### 2015-04-29 paper/paper.tex
Add one parallel algorithm.

\documentclass{article}
\usepackage{amsmath, amssymb}
\usepackage[retainorgcmds]{IEEEtrantools}
\usepackage{filecontents}
\usepackage{hyperref}
\usepackage{graphicx}
\author{Derek Kuo, Henry Milner}
\title{CS267 HW2: Particle Simulation}
\date{3/13/15}

% Some functions for general use.

\newcommand{\code}[1]%
  {\texttt{#1}}

\def\seqn#1\eeqn{\begin{align}#1\end{align}}

\newcommand{\vecName}[1]%
  {\boldsymbol{#1}}

\newcommand{\io}%
  {\text{ i.o. }}

\newcommand{\eventually}%
  {\text{ eventually }}

\newcommand{\tr}%
  {\text{tr}}

\newcommand{\Cov}%
  {\text{Cov}}

\newcommand{\adj}%
  {\text{adj}}

\newcommand{\funcName}[1]%
  {\text{#1}}

\newcommand{\hasDist}%
  {\sim}

\DeclareMathOperator*{\E}%
  {\mathbb{E}}

\newcommand{\Var}%
  {\text{Var}}

\newcommand{\std}%
  {\text{std}}

\newcommand{\grad}%
  {\nabla}

\DeclareMathOperator*{\argmin}{arg\,min}

\DeclareMathOperator*{\argmax}{arg\,max}

\newcommand{\inprod}[2]%
  {\langle #1, #2 \rangle}

\newcommand{\dd}[1]%
  {\frac{\delta}{\delta#1}}

\newcommand{\Reals}%
  {\mathbb{R}}

\newcommand{\indep}%
  {\protect\mathpalette{\protect\independenT}{\perp}} \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}

\newcommand{\defeq}%
  {\buildrel\triangle\over =}

\newcommand{\defn}[1]%
  {\emph{Definition: #1}\\}

\newcommand{\example}[1]%
  {\emph{Example: #1}\\}

\newcommand{\figref}[1]%
  {\figurename~\ref{#1}}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newenvironment{proof}[1][Proof]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\begin{filecontents}{\jobname.bib}
@article{erdos1975problems,
  title={Problems and results on 3-chromatic hypergraphs and some related questions},
  author={Erdos, Paul and Lov{\'a}sz, L{\'a}szl{\'o}},
  journal={Infinite and finite sets},
  volume={10},
  number={2},
  pages={609--627},
  year={1975}
}
@article{beck1991algorithmic,
  title={An algorithmic approach to the Lov{\'a}sz local lemma. I},
  author={Beck, J{\'o}zsef},
  journal={Random Structures \& Algorithms},
  volume={2},
  number={4},
  pages={343--365},
  year={1991},
  publisher={Wiley Online Library}
}
@article{moser2010constructive,
 author = {Moser, Robin A. and Tardos, G\'{a}bor},
 title = {A Constructive Proof of the General Lov\ÁSz Local Lemma},
 journal = {J. ACM},
 issue_date = {January 2010},
 volume = {57},
 number = {2},
 month = feb,
 year = {2010},
 issn = {0004-5411},
 pages = {11:1--11:15},
 articleno = {11},
 numpages = {15},
 url = {http://doi.acm.org/10.1145/1667053.1667060},
 doi = {10.1145/1667053.1667060},
 acmid = {1667060},
 publisher = {ACM},
 address = {New York, NY, USA},
 keywords = {Constructive proof, Lov\'{a}sz local lemma, parallelization},
} 
@article{haeupler2011new,
  title={New constructive aspects of the lovasz local lemma},
  author={Haeupler, Bernhard and Saha, Barna and Srinivasan, Aravind},
  journal={Journal of the ACM (JACM)},
  volume={58},
  number={6},
  pages={28},
  year={2011},
  publisher={ACM}
}
@inproceedings{chung2014distributed,
  title={Distributed algorithms for the Lov{\'a}sz local lemma and graph coloring},
  author={Chung, Kai-Min and Pettie, Seth and Su, Hsin-Hao},
  booktitle={Proceedings of the 2014 ACM symposium on Principles of distributed computing},
  pages={134--143},
  year={2014},
  organization={ACM}
}
@inproceedings{freer2010probabilistic,
  title={When are probabilistic programs probably computationally tractable?},
  author={Freer, Cameron E and Mansinghka, Vikash K and Roy, Daniel M},
  booktitle={NIPS Workshop on Advanced Monte Carlo Methods with Applications},
  year={2010}
}
@article{wainwright2008graphical,
  title={Graphical models, exponential families, and variational inference},
  author={Wainwright, Martin J and Jordan, Michael I},
  journal={Foundations and Trends{\textregistered} in Machine Learning},
  volume={1},
  number={1-2},
  pages={1--305},
  year={2008},
  publisher={Now Publishers Inc.}
}
@inproceedings{papadimitriou1991selecting,
  title={On selecting a satisfying truth assignment},
  author={Papadimitriou, Christos H},
  booktitle={Foundations of Computer Science, 1991. Proceedings., 32nd Annual Symposium on},
  pages={163--169},
  year={1991},
  organization={IEEE}
}
@inproceedings{steurer2010fast,
  title={Fast SDP algorithms for constraint satisfaction problems},
  author={Steurer, David},
  booktitle={Proceedings of the twenty-first annual ACM-SIAM symposium on Discrete Algorithms},
  pages={684--697},
  year={2010},
  organization={Society for Industrial and Applied Mathematics}
}
@inproceedings{polik2007sedumi,
  title={SeDuMi: a package for conic optimization},
  author={Polik, Imre and Terlaky, Tamas and Zinchenko, Yuriy},
  booktitle={IMA workshop on Optimization and Control, Univ. Minnesota, Minneapolis},
  year={2007}
}
@article{malaguti2010survey,
  title={A survey on vertex coloring problems},
  author={Malaguti, Enrico and Toth, Paolo},
  journal={International Transactions in Operational Research},
  volume={17},
  number={1},
  pages={1--34},
  year={2010},
  publisher={Wiley Online Library}
}
@article{gomes2008satisfiability,
  title={Satisfiability solvers},
  author={Gomes, Carla P and Kautz, Henry and Sabharwal, Ashish and Selman, Bart},
  journal={Foundations of Artificial Intelligence},
  volume={3},
  pages={89--134},
  year={2008},
  publisher={Elsevier}
}
@article{balint2013generating,
  title={Generating the uniform random benchmarks for SAT Competition 2013},
  author={Balint, Adrian and Belov, Anton and Heule, Marijn JH and J{\"a}rvisalo, Matti},
  journal={Proceedings of SAT Competition},
  pages={97--98},
  year={2013}
}
@article{belov2014application,
  title={The Application and the Hard Combinatorial Benchmarks in SAT Competition 2014},
  author={Belov, Anton and Heule, Marijn JH and Diepold, Daniel and J{\"a}rvisalo, Matti},
  journal={SAT COMPETITION 2014},
  pages={81}
}
\end{filecontents}
\immediate\write18{bibtex \jobname}

\begin{document}
\maketitle

\begin{abstract}
The algorithmic Lovasz Local Lemma (LLL) is a recent-developed meta-algorithm for solving certain instances of hard combinatorial problems.  In theory, LLL algorithms can be applied whenever a problem can be decomposed into smaller subproblems and the dependencies among subproblems are somewhat sparse.  For example, a k-SAT problem can be decomposed into the problem of satisfying each of its clauses, and LLL algorithms can be useful in solving k-SAT problems whose clauses do not share literals with too many other clauses.  The seminal paper in the field is by Moser and Tardos \cite{moser2010constructive} in 2010.  For many problems, LLL algorithms theoretically lend themselves well to parallelism, as demonstrated in \cite{moser2010constructive} and expanded upon in more recent work \cite{chung2014distributed,haeupler2011new}.  However, we are aware of no published empirical tests of the efficiency of LLL algorithms in practice.  In this paper, we give practical implementations of parallel LLL algorithms for k-SAT, and we investigate their performance.

\section{Introduction}

\subsection{This paper}

\bibliographystyle{plain}
\bibliography{\jobname}

\end{document}

### 2015-04-29 julia/Applications/KsatTests.jl

@@ -1,5 +1,5 @@
 using ArgParse
-using GeneratedData, Problems, Solvers
+using Utils, GeneratedData, Problems, Solvers
 
 function compareGroundTruth(problem:: KsatProblem)
   println("Using exhaustive search to find a solution.")
@@ -11,10 +11,6 @@ function compareGroundTruth(problem:: KsatProblem)
   else
     println("The problem has no solution.")
   end
-end
-
-function parseJuliaArg(solverText:: String)
-  eval(parse(solverText))
 end
 
 function parseArgs()
@@ -29,7 +25,7 @@ function parseArgs()
       help = "whether to solve exhaustively to compare with ground truth"
       action = :store_true
     "--seed", "-e"
-      help = "the random seed"
+      help = "the random seed" #FIXME
       arg_type = Int
   end
 
@@ -37,9 +33,10 @@ function parseArgs()
 end
 
 function run()
-  args = parseArgs()
-  const problemGenerator = parseJuliaArg(args["data-generator"])
-  const solver = parseJuliaArg(args["solver"])
+  # @foo()
+  const args = parseArgs()
+  const problemGenerator = eval(parse(args["data-generator"]))
+  const solver = eval(parse(args["solver"]))
   const problem = generate(problemGenerator)
   if args["compare-ground-truth"]
     compareGroundTruth(problem)

### 2015-04-29 julia/Problems/KsatProblem.jl

@@ -1,4 +1,4 @@
-export KsatProblem, SatClause, KsatSolution, isSuccessful, unsuccessfulKsatSolution, KsatAssignment, isSatisfied, checkSuccess, uniformRandomAssignment
+export KsatProblem, numClauses, SatClause, KsatSolution, isSuccessful, unsuccessfulKsatSolution, KsatAssignment, isSatisfied, checkSuccess, uniformRandomAssignment
 
 # A disjunction of variables.
 immutable SatClause
@@ -26,10 +26,13 @@ end
 const MAX_DISPLAYED_LITERALS = 200
 function Base.string(this:: KsatProblem)
   const maxDisplayedClauses = int(MAX_DISPLAYED_LITERALS / this.k)
-  const numClauses = length(this.clauses)
-  const numDisplayedClauses = min(maxDisplayedClauses, numClauses)
-  const clausesString = join(map(c -> "($(string(c)))", this.clauses[1:min(length(this.clauses), numDisplayedClauses)]), "&")
-  "$(this.k)-SAT problem $(clausesString)$(numDisplayedClauses < numClauses ? "..." : "")"
+  const numDisplayedClauses = min(maxDisplayedClauses, numClauses(this))
+  const clausesString = join(map(c -> "($(string(c)))", this.clauses[1:numDisplayedClauses]), "&")
+  "$(this.k)-SAT problem $(clausesString)$(numDisplayedClauses < numClauses(this) ? "..." : "")"
+end
+
+function numClauses(this:: KsatProblem)
+  length(this.clauses)
 end
 
 

### 2015-04-29 julia/Solvers/IndependentSets.jl

@@ -1,0 +1,69 @@
+export maximalIndependentSet, lubyIndependentSet
+
+# Some algorithms for finding independent sets in undirected graphs.
+# Depends on KsatDependencyGraph.jl.
+# 
+# Some of the algorithms here could, in principle, be parallelized or 
+# distributed.  Currently we have not done that, but we have marked places that
+# would need to be modified to make algorithms run in parallel.
+
+typealias NodeMarker Int64
+
+# Luby's algorithm for finding maximal independent sets.  It is a Las Vegas
+# algorithm, meaning that its running time is random, but it always returns
+# a maximal independent set.
+# See, for example, http://courses.csail.mit.edu/6.852/08/papers/Luby.pdf .
+function maximalIndependentSet(this:: DependencyGraph)
+  remainingGraph = this
+  independentSet = Set{Int64}()
+  while !isempty(nodes(remainingGraph))
+    const newIndependentNodes = lubyIndependentSet(remainingGraph)
+    #NOTE: Could be parallelized.
+    union!(independentSet, newIndependentNodes)
+    const coveredNodes = union(map(node -> neighbors(remainingGraph, node), newIndependentNodes)...)
+    const remainingNodes = setdiff(nodes(remainingGraph), coveredNodes)
+    remainingGraph = inducedSubgraph(remainingGraph, remainingNodes)
+  end
+  independentSet
+end
+
+# One iteration of Luby's algorithm for finding maximal independent sets.
+# This merely finds an independent (but not necessarily maximal) set.
+# See, for example, http://courses.csail.mit.edu/6.852/08/papers/Luby.pdf .
+function lubyIndependentSet(this:: DependencyGraph)
+  const ranks = markNodesRandomly(this)
+  findLocalMinima(this, ranks)
+end
+
+# Assign a random marker of type NodeMarker to each vertex of @this.
+function markNodesRandomly(this:: DependencyGraph)
+  #NOTE: Could be parallelized.
+  marks = Dict{Int64,NodeMarker}()
+  const vs = nodes(this)
+  const randomMarks = rand(NodeMarker, length(vs))
+  for (i, v) in enumerate(vs)
+    marks[v] = randomMarks[i]
+  end
+  marks
+end
+
+# Find all local minima of @values (which maps nodes to markers) in graph
+# @this.  A local minimum is a node in @this whose neighbors in @this have
+# value greater than or equal to it.
[…]

### 2015-04-29 julia/Solvers/KsatDependencyGraph.jl

@@ -1,34 +1,111 @@
-export KsatDependencyGraph, makeKsatDependencyGraph, hasEdge
+export DependencyGraph, hasEdge, hasNode, degree, neighbors, KsatDependencyGraph, makeKsatDependencyGraph, inducedSubgraph
 
-immutable KsatDependencyGraph
+abstract DependencyGraph
+
+function nodes(this:: DependencyGraph)
+  raiseAbstract("nodes", this)
+end
+
+# A Set{Int64} of directed edges in the graph (each undirected edge (i,j)
+# counts as both (i,j) and (j,i)).
+function edges(this:: DependencyGraph)
+  raiseAbstract("edges", this)
+end
+
+function neighbors(this:: DependencyGraph, node:: Int64)
+  raiseAbstract("neighbors", this)
+end
+
+# True if there is an edge between @a and @b.
+function hasEdge(this:: DependencyGraph, a:: Int64, b:: Int64)
+  (a, b) in edges(this)
+end
+
+# True if this graph contains a node with ID @node.
+function hasNode(this:: DependencyGraph, node:: Int64)
+  node in nodes(this)
+end
+
+# The degree of @node.
+function degree(this:: DependencyGraph, node:: Int64)
+  length(neighbors(this, node))
+end
+
+function maxDegree(this:: DependencyGraph)
+  maxDeg = 0
+  for node in nodes(this)
+    maxDeg = max(maxDeg, degree(this, node))
+  end
+  maxDeg
+end
+
+function totalDegree(this:: DependencyGraph)
+  length(edges(this))
+end
+
+function inducedSubgraph(this:: DependencyGraph, nodeSubset:: Set{Int64})
+  KsatDependencySubgraph(this, intersect(nodes(this), nodeSubset))
+end
+
+
+immutable KsatDependencyGraph <: DependencyGraph
   problem:: KsatProblem
+  #TODO: Could use a full IntSet for this.
+  nodes:: Set{Int64}
   # Two clauses have an edge if they share a variable (its sign doesn't
   # matter).  This is a symmetric relation, so this dictionary holds edges
   # in both directions.
   edges:: Set{(Int64, Int64)}
+  neighbors:: Dict{Int64, Set{Int64}}
   maxDegree:: Int64
   totalDegree:: Int64
 end
 
+function nodes(this:: DependencyGraph)
+  this.nodes
+end
+
[…]

### 2015-04-29 julia/Solvers/ParallelLllKsatSolver.jl

@@ -1,0 +1,107 @@
+export ParallelLllKsatSolver, solve, IndependentSetFinder, MoserTardosIndependentSetFinder, SimpleChungIndependentSetFinder
+
+using Problems
+
+abstract IndependentSetFinder
+
+
+# This is a _sequential_ implementation of a parallel algorithm.
+# The algorithms here could be parallelized or distributed.  Currently we have
+# not done that, but we have marked places that would need to be modified to 
+# make algorithms run in parallel.
+
+immutable ParallelLllKsatSolver <: Solver{KsatProblem}
+  # An upper bound on the probability that we fail to find a satisfying
+  # solution given that one exists.  This, along with the problem size,
+  # determines the maximum number of iterations taken to find a solution.
+  # When no satisfying solution exists,
+  # the solver will always take that many iterations, since this solver
+  # produces no certificate of failure.
+  failureProbability:: Float64
+  #NOTE: Needs to be parallelized if the rest of the algorithm is to be
+  # parallelized.
+  independentSetFinder:: IndependentSetFinder
+end
+
+function solve(this:: ParallelLllKsatSolver, problem:: KsatProblem)
+  # See Moser and Tardos 2010, algorithm 1.1.
+  const n = problem.numVariables
+  println("Problem: $(string(problem))")
+  assignment = uniformRandomAssignment(n)
+  annotatedProblem = annotateKsatProblem(problem, assignment)
+  #NOTE: Could be parallelized.
+  const graph = makeKsatDependencyGraph(problem)
+  const maxNumIterations = calculateNumIterations(this.independentSetFinder, problem, graph, this.failureProbability)
+  println("Using at most $(maxNumIterations) iterations for $(problem.k)-SAT problem with $(n) variables and $(numClauses(problem)) clauses.")
+  for i = 1:maxNumIterations
+    #NOTE: Could be parallelized.
+    const unsat = unsatisfiedClauses(annotatedProblem)
+    if isempty(unsat)
+      println("Found successful solution after $(i) iterations.")
+      return toSuccessfulSolution(annotatedProblem)
+    end
+    const independentSet = find(this.independentSetFinder, inducedSubgraph(graph, unsat))
+    #FIXME
+    println("Found an independent set of size $(length(independentSet))")
+    #NOTE: Could be parallelized.  For example, if clauses were distributed
[…]

### 2015-04-29 julia/Solvers/RandomWalkKsatSolver.jl

@@ -18,7 +18,7 @@ function solve(this:: RandomWalkKsatSolver, problem:: KsatProblem)
   const numAssignmentsPerAttempt = 3*n
   const expectedNumStepsRequired = n^1.5 * (4.0/3.0)^n
   const numAttempts = int(ceil(2*(-1)*log2(this.failureProbability)*expectedNumStepsRequired))
-  println("Making $(numAttempts) attempts for $(problem.k)-SAT problem with $(n) variables and $(length(problem.clauses)) clauses, with $(numAssignmentsPerAttempt) assignments per attempt.")
+  println("Making $(numAttempts) attempts for $(problem.k)-SAT problem with $(n) variables and $(numClauses(problem)) clauses, with $(numAssignmentsPerAttempt) assignments per attempt.")
   println("Problem: $(string(problem))")
   findSolution(numAttempts, problem)
 end
@@ -57,7 +57,7 @@ function solve(this:: SimpleRandomKsatSolver, problem:: KsatProblem)
     clause = problem.clauses[clauseIdx]
     variableInClauseIdx = rand(1:problem.k)
     variable = clause.variables[variableInClauseIdx]
-    # println("Flipping variable $(variable) ($(variableInClauseIdx) in clause $(clauseIdx)/$(length(problem.clauses)) $(clause))")
+    # println("Flipping variable $(variable) ($(variableInClauseIdx) in clause $(clauseIdx)/$(numClauses(problem)) $(clause))")
     currentValue = annotatedProblem.currentAssignment[variable]
     updateAssignment!(annotatedProblem, variable, currentValue, !currentValue)
   end

### 2015-04-29 julia/Solvers/SimpleMoserTardosKsatSolver.jl

@@ -16,7 +16,7 @@ function solve(this:: SimpleMoserTardosKsatSolver, problem:: KsatProblem)
   # See Moser and Tardos 2010, algorithm 1.1.
   const n = problem.numVariables
   const maxNumIterations = 1000 #FIXME
-  println("Using at most $(maxNumIterations) iterations for $(problem.k)-SAT problem with $(n) variables and $(length(problem.clauses)) clauses.")
+  println("Using at most $(maxNumIterations) iterations for $(problem.k)-SAT problem with $(n) variables and $(numClauses(problem)) clauses.")
   println("Problem: $(string(problem))")
   assignment = uniformRandomAssignment(n)
   annotatedProblem = annotateKsatProblem(problem, assignment)

### 2015-04-29 julia/Solvers/module.jl

@@ -3,8 +3,10 @@ module Solvers
 include("./Solvers.jl")
 include("./KsatUtils.jl")
 include("./KsatDependencyGraph.jl")
+include("./IndependentSets.jl")
 include("./ExhaustiveKsatSolver.jl")
 include("./RandomWalkKsatSolver.jl")
 include("./SimpleMoserTardosKsatSolver.jl")
+include("./ParallelLllKsatSolver.jl")
 
 end

### 2015-04-29 julia/packages.jl

@@ -1,1 +1,4 @@
 # Installs all packages required for this project.
+
+Pkg.add("Distribution")
+Pkg.add("ArgParse")

### 2015-04-30 julia/Solvers/IndependentSets.jl
Add Weak MIS algorithm from Chung et al.  It may still be buggy.  Example command to run: ``./run Applications/KsatTests.jl -s "ParallelLllKsatSolver(WeakMisFinder())" -d "RandomKsatGenerator(6,18,250)" -g``

@@ -20,7 +20,7 @@ function maximalIndependentSet(this:: DependencyGraph)
     const newIndependentNodes = lubyIndependentSet(remainingGraph)
     #NOTE: Could be parallelized.
     union!(independentSet, newIndependentNodes)
-    const coveredNodes = union(map(node -> neighbors(remainingGraph, node), newIndependentNodes)...)
+    const coveredNodes = neighborhood(remainingGraph, newIndependentNodes)
     const remainingNodes = setdiff(nodes(remainingGraph), coveredNodes)
     remainingGraph = inducedSubgraph(remainingGraph, remainingNodes)
   end

### 2015-04-30 julia/Solvers/KsatDependencyGraph.jl

@@ -16,6 +16,10 @@ function neighbors(this:: DependencyGraph, node:: Int64)
   raiseAbstract("neighbors", this)
 end
 
+function neighborhood(this:: DependencyGraph, nodes:: Set{Int64})
+  union(map(node -> neighbors(this, node), nodes)...)
+end
+
 # True if there is an edge between @a and @b.
 function hasEdge(this:: DependencyGraph, a:: Int64, b:: Int64)
   (a, b) in edges(this)
@@ -85,7 +89,7 @@ function inducedSubgraph(this:: KsatDependencyGraph, nodeSubset:: Set{Int64})
 end
 
 function makeKsatDependencyGraph(problem:: KsatProblem)
-  nodes = Set(1:problem.numVariables)
+  nodes = Set(1:numClauses(problem))
   edges = Set{(Int64, Int64)}()
   neighbors = Dict{Int64, Set{Int64}}()
   maxDegree = 0

### 2015-04-30 julia/Solvers/ParallelLllKsatSolver.jl

@@ -1,8 +1,25 @@
-export ParallelLllKsatSolver, solve, IndependentSetFinder, MoserTardosIndependentSetFinder, SimpleChungIndependentSetFinder
+export ParallelLllKsatSolver, solve
+
+export IndependentSetFinder, MoserTardosIndependentSetFinder, SimpleChungIndependentSetFinder, WeakMisFinder, buildFinderFunc, calculateNumIterations
 
 using Problems
 
+using Distributions
+
+
 abstract IndependentSetFinder
+
+# Return a function from a subgraph of @problemGraph to an independent set in
+# that subgraph.
+function buildFinderFunc(this:: IndependentSetFinder, problemGraph:: DependencyGraph)
+  raiseAbstract("buildFinderFunc", this)
+end
+
+# Compute the number of iterations required by ParallelLllKsatSolver when
+# using this method to find independent sets.
+function calculateNumIterations(this:: IndependentSetFinder, problem:: KsatProblem, graph:: DependencyGraph)
+  raiseAbstract("calculateNumIterations", this)
+end
 
 
 # This is a _sequential_ implementation of a parallel algorithm.
@@ -11,27 +28,21 @@ abstract IndependentSetFinder
 # make algorithms run in parallel.
 
 immutable ParallelLllKsatSolver <: Solver{KsatProblem}
-  # An upper bound on the probability that we fail to find a satisfying
-  # solution given that one exists.  This, along with the problem size,
-  # determines the maximum number of iterations taken to find a solution.
-  # When no satisfying solution exists,
-  # the solver will always take that many iterations, since this solver
-  # produces no certificate of failure.
-  failureProbability:: Float64
   #NOTE: Needs to be parallelized if the rest of the algorithm is to be
   # parallelized.
   independentSetFinder:: IndependentSetFinder
 end
 
 function solve(this:: ParallelLllKsatSolver, problem:: KsatProblem)
-  # See Moser and Tardos 2010, algorithm 1.1.
+  # See Moser and Tardos 2010, algorithms 1 and 2.
   const n = problem.numVariables
   println("Problem: $(string(problem))")
   assignment = uniformRandomAssignment(n)
   annotatedProblem = annotateKsatProblem(problem, assignment)
   #NOTE: Could be parallelized.
   const graph = makeKsatDependencyGraph(problem)
-  const maxNumIterations = calculateNumIterations(this.independentSetFinder, problem, graph, this.failureProbability)
[…]

### 2015-04-30 julia/Solvers/SimpleMoserTardosKsatSolver.jl

@@ -3,13 +3,6 @@ export SimpleMoserTardosKsatSolver, solve
 using Problems
 
 immutable SimpleMoserTardosKsatSolver <: Solver{KsatProblem}
-  # An upper bound on the probability that we fail to find a satisfying
-  # solution given that one exists.  This, along with the problem size,
-  # determines the maximum number of iterations taken to find a solution.
-  # When no satisfying solution exists,
-  # the solver will always take that many iterations, since this solver
-  # produces no certificate of failure.
-  failureProbability:: Float64
 end
 
 function solve(this:: SimpleMoserTardosKsatSolver, problem:: KsatProblem)

### 2015-04-30 julia/Solvers/IndependentSets.jl
Fixed several bugs in WeakMisFinder, which now works.

@@ -1,5 +1,7 @@
 export maximalIndependentSet, lubyIndependentSet
 
+using Distributions
+
 # Some algorithms for finding independent sets in undirected graphs.
 # Depends on KsatDependencyGraph.jl.
 # 
@@ -31,32 +33,52 @@ end
 # This merely finds an independent (but not necessarily maximal) set.
 # See, for example, http://courses.csail.mit.edu/6.852/08/papers/Luby.pdf .
 function lubyIndependentSet(this:: DependencyGraph)
-  const ranks = markNodesRandomly(this)
-  findLocalMinima(this, ranks)
+  const marked = markNodesRandomly(this)
+  findLocalMinima(marked)
 end
 
 # Assign a random marker of type NodeMarker to each vertex of @this.
 function markNodesRandomly(this:: DependencyGraph)
+  markNodesRandomly(this, DiscreteUniform(typemin(NodeMarker), typemax(NodeMarker)))
+end
+
+# Assign IID draws from @distribution to each node of @this.
+# @param distribution should be a distribution on elements of type NodeMarker.
+function markNodesRandomly(this:: DependencyGraph, distribution:: Sampleable{Univariate,Discrete})
   #NOTE: Could be parallelized.
   marks = Dict{Int64,NodeMarker}()
   const vs = nodes(this)
-  const randomMarks = rand(NodeMarker, length(vs))
+  const randomMarks = rand(distribution, length(vs))
   for (i, v) in enumerate(vs)
     marks[v] = randomMarks[i]
   end
-  marks
+  MarkedGraph{NodeMarker}(this, marks)
 end
 
-# Find all local minima of @values (which maps nodes to markers) in graph
-# @this.  A local minimum is a node in @this whose neighbors in @this have
-# value greater than or equal to it.
-function findLocalMinima(this:: DependencyGraph, values:: Dict{Int64,NodeMarker})
+# Find all strict local minima of @values (which maps nodes to markers) in graph
+# @this.  A strict local minimum is a node in @this whose neighbors in @this
+# have value greater than it.
+function findLocalMinima(this:: MarkedGraph{NodeMarker})
+  findLocalMinima(this, <)
+end
+
+# Find all strict local minima of @values (which maps nodes to markers) in graph
+# @this.  A strict local minimum is a node in @this whose neighbors in @this
+# have value greater than it according to @lessThanFunc.
+# @param lessThanFunc should be a binary operator like < that induces a
+# total order on nodes.
[…]

### 2015-04-30 julia/Solvers/KsatDependencyGraph.jl

@@ -1,4 +1,4 @@
-export DependencyGraph, hasEdge, hasNode, degree, neighbors, KsatDependencyGraph, makeKsatDependencyGraph, inducedSubgraph
+export DependencyGraph, nodes, edges, hasEdge, hasNode, degree, neighbors, neighborhood, maxDegree, MarkedGraph, KsatDependencyGraph, makeKsatDependencyGraph, inducedSubgraph
 
 abstract DependencyGraph
 
@@ -17,7 +17,7 @@ function neighbors(this:: DependencyGraph, node:: Int64)
 end
 
 function neighborhood(this:: DependencyGraph, nodes:: Set{Int64})
-  union(map(node -> neighbors(this, node), nodes)...)
+  union(nodes, map(node -> neighbors(this, node), nodes)...)
 end
 
 # True if there is an edge between @a and @b.
@@ -49,6 +49,12 @@ end
 
 function inducedSubgraph(this:: DependencyGraph, nodeSubset:: Set{Int64})
   KsatDependencySubgraph(this, intersect(nodes(this), nodeSubset))
+end
+
+
+immutable MarkedGraph{MarkType}
+  graph:: DependencyGraph
+  marks:: Dict{Int64, MarkType}
 end
 
 
@@ -97,7 +103,7 @@ function makeKsatDependencyGraph(problem:: KsatProblem)
   for (clauseIdx, clause) in enumerate(problem.clauses)
     numEdges = 0
     for (otherClauseIdx, otherClause) in enumerate(problem.clauses)
-      if checkForEdge(clause, otherClause)
+      if clauseIdx != otherClauseIdx && checkForEdge(clause, otherClause)
         push!(edges, (clauseIdx, otherClauseIdx))
         push!(get!(() -> Set{Int64}(), neighbors, clauseIdx), otherClauseIdx)
         numEdges += 1

### 2015-04-30 julia/Solvers/ParallelLllKsatSolver.jl

@@ -52,6 +52,7 @@ function solve(this:: ParallelLllKsatSolver, problem:: KsatProblem)
       return toSuccessfulSolution(annotatedProblem)
     end
     const independentSet = independentSetFunc(inducedSubgraph(graph, unsat))
+    # println("Using an independent set of size $(length(independentSet)) on iteration $i")
     #NOTE: Could be parallelized.  For example, if clauses were distributed
     # across machines and separate variable states maintained per machine, each
     # time a clause i's variables are updated, the updates should be
@@ -121,9 +122,9 @@ immutable WeakMisFinder <: IndependentSetFinder end
 
 function buildFinderFunc(this:: WeakMisFinder, graph:: DependencyGraph)
   const d = maxDegree(graph)
+  const maxNumIterations = 4*e^2*log(2*e*(d+1.0)^4)
   
   function findIndependentSet(subgraph:: DependencyGraph)
-    const maxNumIterations = 4*e^2*log(2*e*(d+1.0)^4)
     # This variable is called S in Chung et al.
     independentSet = Set{Int64}()
     for epoch = 1:maxNumIterations
@@ -131,6 +132,9 @@ function buildFinderFunc(this:: WeakMisFinder, graph:: DependencyGraph)
       # member of the independent set we are growing.
       #NOTE: Could be parallelized.
       candidateNodes = setdiff(nodes(subgraph), neighborhood(subgraph, independentSet))
+      if isempty(candidateNodes)
+        break
+      end
       # This variable is called G' in Chung et al.
       candidateSubgraph = inducedSubgraph(subgraph, candidateNodes)
       # In each phase, we find high-degree vertices (with the threshold lowered
@@ -143,24 +147,30 @@ function buildFinderFunc(this:: WeakMisFinder, graph:: DependencyGraph)
       for phase = 1:int(ceil(log(d)))
         #NOTE: Could be parallelized.
         minDegree = d / 2^phase
-        marks = Dict{Int64,Bool}()
+        # This variable is called V_i in Chung et al.
+        # In this simplified version of Luby's algorithm, each node is marked
+        # with a 0 or 1, and only nodes that are marked 1 and have no neighbors
+        # marked 1 are chosen.  (So it is Luby's algorithm with the order on
+        # marks switched, the marks limited to {0,1}, and a nonuniform sampling
+        # distribution for the marks.)
+        samplingProbability = 1.0/(d*2^(-phase+1.0)+1.0)
[…]

### 2015-04-30 mpi/test.cpp
Add some skeleton files for MPI implementation.

@@ -1,0 +1,29 @@
+// This is just an example MPI source.  Compile and run with:
+//   OMPI_MPICXX="g++-4.9 -Wall -std=c++0x" mpic++ -o test test.cpp
+//   mpirun -np 2 ./test
+
+#include <mpi.h>
+#include <iostream>
+
+int main(int argc, char *argv[])
+{
+  MPI_Init(&argc, &argv);
+
+  int rank;
+  MPI_Comm_rank(MPI_COMM_WORLD, &rank);
+  if (rank == 0) {
+    int value = 17;
+    int result = MPI_Send(&value, 1, MPI_INT, 1, 0, MPI_COMM_WORLD);
+    if (result == MPI_SUCCESS)
+      std::cout << "Rank 0 OK!" << std::endl;
+  } else if (rank == 1) {
+    int value;
+    int result = MPI_Recv(&value, 1, MPI_INT, 0, 0, MPI_COMM_WORLD,
+			  MPI_STATUS_IGNORE);
+    if (result == MPI_SUCCESS && value == 17)
+      std::cout << "Rank 1 OK!" << std::endl;
+  }
+  MPI_Finalize();
+  return 0;
+}
+

### 2015-05-11 mpi/CollectionUtils.py
Python MPI code finally works.  Example: time mpirun -np 4 python ParallelKsatApplication.py --generator "MpiKsatProblem.RandomMpiKsatGenerator(3, 5, 20)" --solver "LllKsatSolver.LllKsatSolver(IndependentSetFinder.SimpleChungIndependentSetFinder())" --problemSeed 0 --solverSeed 10

@@ -4,11 +4,14 @@
 import copy
 import itertools
 
+def iterlen(iterator):
+  return sum(1 for i in iterator)
+
 def intersection(sets):
   if len(sets) == 0:
     return set()
   base = copy.copy(sets[0])
-  for s in itertools.islice(sets, 1):
+  for s in itertools.islice(sets, 1, None):
     base &= s
   return base
 
@@ -16,7 +19,7 @@ def intersection_update(sets):
   if len(sets) == 0:
     return set()
   base = sets[0]
-  for s in itertools.islice(sets, 1):
+  for s in itertools.islice(sets, 1, None):
     base &= s
   return base
 
@@ -30,7 +33,7 @@ def union_update(sets):
   if len(sets) == 0:
     return set()
   base = sets[0]
-  for s in itertools.islice(sets, 1):
+  for s in itertools.islice(sets, 1, None):
     base |= s
   return base
 
@@ -46,6 +49,6 @@ def addAll_update(dicts):
   if len(dicts) == 0:
     return {}
   base = dicts[0]
-  for d in itertools.islice(dicts, 1):
+  for d in itertools.islice(dicts, 1, None):
     base.update(d)
   return base

### 2015-05-11 mpi/ExhaustiveMpiKsatSolver.py

@@ -98,8 +98,9 @@ class ExhaustiveProcessParallelSolver(MpiKsatSolver):
       MpiKsatAssignment.unpackTo(binaryAssignment, currentLocalAssignment, n)
       return currentLocalAssignment.satisfiesClauses(allClauses)
     satisfyingBinaryAssignments = binaryAssignments.filter(checkBinaryAssignment)
-    numSatisfyingAssignments = satisfyingBinaryAssignments.size()
-    arbitrarySatisfyingAssignment = satisfyingBinaryAssignments.reduce(-1, max)
+    numSatisfyingAssignments, arbitrarySatisfyingAssignment = (satisfyingBinaryAssignments
+      .map(lambda elt: (1, elt))
+      .reduce((0, -1), lambda countAndMaxA, countAndMaxB: (countAndMaxA[0] + countAndMaxB[0], max(countAndMaxA[1], countAndMaxB[1]))))
     
     def buildSolution():
       print "%d out of %d satisfying assignments." % (numSatisfyingAssignments, numPotentialSolutions)

### 2015-05-11 mpi/IndependentSetFinder.py

@@ -1,19 +1,16 @@
 import math
+import itertools
 
 import MpiGraph
 import MpiCollection
 import Utils
 
-# Returns a maximal independent set (in the form of an MpiSet) in
-# @graph.
-def maximalIndependentSet(rand, graph):
-  pass
-
 # Returns an independent set (in the form of an MpiSet) in @graph found 
 # by one iteration of Luby's algorithm.
 def lubyIndependentSet(rand, graph):
   marks = markNodesRandomly(graph, lambda : Utils.randLargeInt(rand))
-  return findLocalMinima(graph, marks)
+  broadcastMarks = marks.collectEverywhere()
+  return findLocalMinima(graph, broadcastMarks)
 
 def markNodesRandomly(graph, distribution):
   return graph.nodesIterator().map(lambda node: (node, distribution())).toDict()
@@ -22,28 +19,10 @@ def findLocalMinima(graph, marks):
   return findLocalMaxima(graph, marks, lambda x, y: x < y)
 
 def findLocalMaxima(graph, marks, comparatorFunc):
-  # We do this in two rounds.  First, we pare down the nodes on a single
-  # processor by removing nodes with larger neighbors residing on the same
-  # processor.  Then the remaining nodes (and their weights) are all-gathered
-  # and a similar algorithm is run again, with each processor checking its
-  # nodes against their neighbors in the rest of the graph.
-  potentialLocalMaxima = findLocalMaximaAmongNodes(graph, graph.nodesIterator(), marks, comparatorFunc)
-  potentialsWithMarks = MpiCollection.dictFromLocal(graph.comm(), {node: marks[node] for node in potentialLocalMaxima})
-  allPotentialsWithMarks = potentialsWithMarks.collectEverywhere()
-  allPotentials = set(allPotentialsWithMarks.keys())
-  remainingGraph = graph.inducedSubgraphPlusFringe(potentialLocalMaxima, allPotentials)
-  localMaxima = findLocalMaximaAmongNodes(remainingGraph, potentialLocalMaxima, allPotentialsWithMarks, comparatorFunc)
-  return localMaxima
-
-# Find elements of @potentialMaxima (a set of nodes) that are strict local 
-# maxima according to @comparatorFunc.
-def findLocalMaximaAmongNodes(graph, potentialMaxima, marks, comparatorFunc):
   def isLocalMaximum(node):
     mark = marks[node]
-    neighbors = graph.localNeighbors(node)
[…]

### 2015-05-11 mpi/LllKsatSolver.py

@@ -2,16 +2,10 @@ from MpiKsatSolver import MpiKsatSolver
 from MpiUtils import onMaster
 import IndependentSetFinder
 import MpiGraph
+import MpiCollection
 import MpiKsatAssignment
 import KsatSolution
 import Utils
-
-# The set of indices of clauses in @problem that are unsatisfied under
-# @assignment.
-def unsatisfiedClauses(problem, assignment):
-  localClausesByIdx = problem.localClausesByIdx()
-  localAssignment = assignment.localAssignment()
-  return setFromLocal(set(clauseIdx for clauseIdx, clause in problem.localClausesByIdx() if localAssignment.satisfiesClause(clause)))
 
 # A class of LLL algorithms for solving distributed k-SAT problems.  Solvers
 # are parameterized by a method for finding independent sets in clause
@@ -24,23 +18,21 @@ class LllKsatSolver(MpiKsatSolver):
   def solve(self, rand, problem):
     comm = problem.comm()
     n = problem.numVariables
-    graph = problem.toDependencyGraph()
-    currentAssignment = MpiKsatAssignment.uniformRandomMpiKsatAssignment(rand, n)
+    broadcastProblem = problem.toBroadcast()
+    graph = broadcastProblem.toDependencyGraph(comm)
+    currentAssignment = MpiKsatAssignment.uniformRandomMpiKsatAssignment(comm, rand, n)
     maxNumIterations = self.independentSetFinder.calculateNumIterations(problem, graph)
-    independentSetFunc = self.independentSetFunc.buildFinderFunc(rand, graph)
+    if comm.rank == 0: print "Solving problem %s... with %s iterations." % (problem, maxNumIterations)
+    independentSetFunc = self.independentSetFinder.buildFinderFunc(rand, graph)
     for i in xrange(maxNumIterations):
       # At the beginning of each iteration of this loop, @currentAssignment
       # is consistent across processors.
-      unsatisfiedClauses = unsatisfiedClauses(problem, currentAssignment)
-      if unsatisfiedClauses.toCollection().sizeEverywhere() == 0:
-        return onMaster(comm, lambda : KsatSolution.success(currentAssignment.localAssignment()))
-      unsatSubgraph = graph.inducedSubgraph(unsatisfiedClauses)
+      localAssignment = currentAssignment.localAssignment()
[…]

### 2015-05-11 mpi/MpiCollection.py

@@ -5,6 +5,10 @@ import CollectionUtils
 from MpiChunks import MpiChunks
 from MpiUtils import onMaster
 
+#FIXME: The collections API evolved organically and is fairly broken.  There
+# should be a proper inheritance hierarchy, with everything inheriting from
+# MpiCollection.
+
 # A simple distributed collection in MPI.  Each processor has its own
 # subcollection.  The implementation follows the MPMD model of MPI.  The
 # programmer should populate the local part of the collection with local
@@ -15,20 +19,24 @@ from MpiUtils import onMaster
 # data.
 # 
 # Due to the simplicity of this implementation, there are many caveats:
-#  - Currently none of the operations on this collection are lazy.
+#  - Some of the operations on this collection are lazy, and other are not.
+#    Check the implementation before relying on a particular behavior.  
+#    reified() can be used to get a materialized copy of the collection.
 #  - It is assumed that no operations on this collection change its
-#    partitioning.  Explicit repartitioning could be supported.
+#    partitioning.  Explicit partitioning could be supported.
 class MpiCollection(object):
   #FIXME
-  def __init__(self, distributedChunks):#, partitioner):
+  def __init__(self, distributedChunks):
     self.distributedChunks = distributedChunks
-    # self.partitioner = partitioner
   
   def map(self, f):
-    return MpiCollection(self.mapChunks(lambda chunk: [f(elt) for elt in chunk]))#, self.partitioner)
+    return MpiCollection(self.mapChunks(lambda chunk: (f(elt) for elt in chunk)))
   
   def filter(self, pred):
-    return MpiCollection(self.mapChunks(lambda chunk: [elt for elt in chunk if pred(elt)]))#, self.partitioner)
+    return MpiCollection(self.mapChunks(lambda chunk: (elt for elt in chunk if pred(elt))))
+  
+  def reified(self):
+    return MpiCollection(self.mapChunks(list))
   
   # Should be called on all processors; will return None on all but processor
   # 0.
@@ -54,12 +62,13 @@ class MpiCollection(object):
   def collectEverywhere(self):
     return itertools.chain.from_iterable(self.distributedChunks.collectEverywhere())
 
-  def mapChunks(self, f):
-    return self.distributedChunks.mapChunks(f)#, self.partitioner)
[…]

### 2015-05-11 mpi/MpiGraph.py

@@ -1,6 +1,8 @@
 import itertools
 
+import CollectionUtils
 import SetUtils
+import MpiCollection
 
 # A distributed undirected graph.  Nodes are simply integer IDs.  Nodes are 
 # distributed, and each node knows its neighbors, some of which may reside
@@ -12,20 +14,25 @@ class Graph(object):
   def __init__(self):
     pass
   
+  def comm(self):
+    raise NotImplementedError()
+  
   def nodesIterator(self):
-    pass
+    raise NotImplementedError()
   
-  def nodesSet(self):
-    pass
+  def neighborsIterator(self, node):
+    raise NotImplementedError()
   
-  def localNeighbors(self, node):
-    pass
+  def filter(self, f):
+    return LazySubgraph(self, f)
   
-  def inducedSubgraphPlusFringe(self, nodes, fringe):
-    return LazyFringedSubgraph(self, nodes, fringe)
+  def sizeEverywhere(self):
+    return self.nodesIterator().sizeEverywhere()
   
-  def inducedSubgraph(self, nodes):
-    return LazySubgraph(self, nodes)
+  def maxDegreeEverywhere(self):
+    return (self.nodesIterator()
+      .map(lambda node: CollectionUtils.iterlen(self.neighborsIterator(node)))
+      .reduceEverywhere(0, max))
   
 
 # A graph composed of a list of nodes distributed across machines, plus the
@@ -42,54 +49,33 @@ class ConcreteGraph(Graph):
     return self.nodesV.comm()
   
   def nodesIterator(self):
-    return self.nodesV
-  
-  def nodesSet(self):
-    return self.nodesV
-  
-  def localNeighbors(self, node):
-    #TODO: May be slow.  Should be O(#neighbors).
-    return self.neighborsV[node] & self.nodesSet()
+    return self.nodesV.toCollection()
+
+  def neighborsIterator(self, node):
+    return self.neighborsV[node]
 
 def graphFromLocal(comm, nodes, neighbors):
   return ConcreteGraph(MpiCollection.setFromLocal(comm, nodes), MpiCollection.dictFromLocal(comm, neighbors))
 
-class LazyFringedSubgraph(Graph):
-  def __init__(self, supergraph, nodeSubset, fringe):
-    self.supergraph = supergraph
-    #FIXME: May need to reify this.
-    self.nodeSubset = nodeSubset
-    self.fringe = fringe
-    super(LazyFringedSubgraph, self).__init__()
-  
-  def comm(self):
-    return self.supergraph.comm()
-  
[…]

### 2015-05-11 mpi/MpiKsatAssignment.py

@@ -19,7 +19,7 @@ class KsatAssignment(object):
     return sum(2**i if value else 0 for i, value in enumerate(self.values))
 
   def __repr__(self):
-    return "".join(itertools.imap(str, self.values))
+    return "".join((str(value) for value in self.values))
 
 def unpack(binaryAssignment, n):
   assignment = emptyKsatAssignment(n)

### 2015-05-11 mpi/MpiKsatProblem.py

@@ -21,6 +21,7 @@ class SatClause(object):
     return itertools.izip(self.literals, self.signs)
   
   def sharesLiteral(self, other):
+    #TODO: Could cache this, if it is expensive.
     otherLiterals = set(other.literals)
     return any(ownLiteral in otherLiterals for ownLiteral in self.literals)
 
@@ -31,31 +32,25 @@ def randomSatClause(k, numVariables, rand):
   return SatClause(literals, signs)
 
 
+class BroadcastKsatProblem(object):
+  def __init__(self, k, numVariables, clausesByIdx, localClauseIndices):
+    self.k = k
+    self.numVariables = numVariables
+    self.clausesByIdx = clausesByIdx
+    self.localClauseIndices = localClauseIndices
+  
+  def toDependencyGraph(self, comm):
+    def findNeighborsIndices(nodeIdx, node):
+      return set(neighborIdx for neighborIdx, neighbor in self.clausesByIdx.items() if nodeIdx != neighborIdx and node.sharesLiteral(neighbor))
+    neighborsOfLocalNodes = {localNodeIdx: findNeighborsIndices(localNodeIdx, self.clausesByIdx[localNodeIdx]) for localNodeIdx in self.localClauseIndices}
+    return MpiGraph.graphFromLocal(comm, self.localClauseIndices, neighborsOfLocalNodes)
+
 # A distributed k-SAT problem.
 class MpiKsatProblem(object):
   def __init__(self, k, numVariables, distributedClausesV):
     self.k = k
     self.numVariables = numVariables
     self.distributedClausesV = distributedClausesV
-  
-  def toDependencyGraph(self):
-    # We assume the number of nodes is small; in that case, the most
-    # communication-efficient way to build the graph is to broadcast the nodes 
-    # and compute edges everywhere.
-    localNodes = self.localClausesByIdx()
-    allNodesByIdx = self.distributedClauses().values().collectEverywhere()
-    
-    def findNeighborsIndices(node):
-      set(neighborIdx for neighborIdx, neighbor in enumerate(allNodesByIdx) if node.sharesLiteral(neighbor))
-    
-    #TODO: Could be more efficient ways to do this.  For example, we could find 
-    # edges by shared variable rather than by examining all pairs of nodes.
-    neighborsOfLocalNodes = {localNodeIdx: findNeighborsIndices(localNode) for localNodeIdx, localNode in localNodes}
[…]

### 2015-05-11 mpi/MpiUtils.py

@@ -8,6 +8,14 @@ import sys
 def reseed(comm, rand):
   return np.random.RandomState(rand.randint(0, sys.maxint) + 53*comm.rank)
 
+# Return a new RandomState whose seed is a deterministic function of @seed
+# and of @idx.  Useful for generating a range of random values that does not
+# change when the partitioning changes, e.g.:
+#   seed = rand.randint()
+#   makeRange(10).map(lambda item: makeRandForItem(seed, item))
+def makeRandForItem(seed, idx):
+  return np.random.RandomState(seed + 59*idx)
+
 # Execute some computation embodied in @thunk on the master, or return None
 # on slaves.  This is just a convenience method for this common pattern in MPI
 # code.  (For example, MpiCollection.collect() will only collect on the master,

### 2015-05-11 mpi/ParallelKsatApplication.py

@@ -13,18 +13,21 @@ def parseArgs():
                      help='a Python expression yielding the KsatGenerator to run')
   parser.add_argument('--solver', action='store',
                      help='a Python expression yielding the MpiKsatSolver to use')
-  parser.add_argument('--seed', action='store', default=None, type=int,
-                     help='the random seed to use')
+  parser.add_argument('--problemSeed', action='store', default=None, type=int,
+                     help='the random seed to use when generating the problem')
+  parser.add_argument('--solverSeed', action='store', default=None, type=int,
+                     help='the random seed to use when solving the problem')
   return parser.parse_args()
 
 def run():
   args = parseArgs()
   comm = MPI.COMM_WORLD
-  rand = np.random.RandomState(args.seed if args.seed is not None else None)
-  problem = eval(args.generator).generate(comm, rand)
+  problemRand = np.random.RandomState(args.problemSeed if args.problemSeed is not None else None)
+  problem = eval(args.generator).generate(comm, problemRand)
   solver = eval(args.solver)
-  
-  solution = solver.solve(rand, problem)
+
+  solverRand = np.random.RandomState(args.solverSeed if args.solverSeed is not None else None)
+  solution = solver.solve(solverRand, problem)
   
   if comm.rank == 0:
     print "Finished running solver %s; found %s" % (args.solver, solution)

### 2015-05-11 mpi/ExhaustiveMpiKsatSolver.py
Fix minor bugs in previous commit.

@@ -98,11 +98,12 @@ class ExhaustiveProcessParallelSolver(MpiKsatSolver):
       MpiKsatAssignment.unpackTo(binaryAssignment, currentLocalAssignment, n)
       return currentLocalAssignment.satisfiesClauses(allClauses)
     satisfyingBinaryAssignments = binaryAssignments.filter(checkBinaryAssignment)
-    numSatisfyingAssignments, arbitrarySatisfyingAssignment = (satisfyingBinaryAssignments
+    resultStatistics = (satisfyingBinaryAssignments
       .map(lambda elt: (1, elt))
       .reduce((0, -1), lambda countAndMaxA, countAndMaxB: (countAndMaxA[0] + countAndMaxB[0], max(countAndMaxA[1], countAndMaxB[1]))))
     
     def buildSolution():
+      numSatisfyingAssignments, arbitrarySatisfyingAssignment = resultStatistics
       print "%d out of %d satisfying assignments." % (numSatisfyingAssignments, numPotentialSolutions)
       if arbitrarySatisfyingAssignment > -1:
         return KsatSolution.success(MpiKsatAssignment.unpack(arbitrarySatisfyingAssignment, n))

### 2015-05-11 mpi/IndependentSetFinder.py

@@ -73,7 +73,7 @@ class SimpleChungIndependentSetFinder(IndependentSetFinder):
       #TODO: Probably need a minimum threshold here, since the theoretical
       # analysis gives only asymptotic guarantees.  But it's not clear what
       # it should be.
-      return max(100, int(ceil(log(base, m))))
+      return max(100, int(math.ceil(math.log(base, m))))
   
   def buildFinderFunc(self, rand, graph):
     marks = markNodesRandomly(graph, lambda : Utils.randLargeInt(rand))

### 2015-05-11 mpi/LllKsatSolver.py

@@ -22,7 +22,7 @@ class LllKsatSolver(MpiKsatSolver):
     graph = broadcastProblem.toDependencyGraph(comm)
     currentAssignment = MpiKsatAssignment.uniformRandomMpiKsatAssignment(comm, rand, n)
     maxNumIterations = self.independentSetFinder.calculateNumIterations(problem, graph)
-    if comm.rank == 0: print "Solving problem %s... with %s iterations." % (problem, maxNumIterations)
+    if comm.rank == 0: print "Solving problem with %s iterations." % maxNumIterations
     independentSetFunc = self.independentSetFinder.buildFinderFunc(rand, graph)
     for i in xrange(maxNumIterations):
       # At the beginning of each iteration of this loop, @currentAssignment

### 2015-05-11 julia/Applications/KsatTests_solution.jl
Performance Measurement

@@ -44,9 +44,9 @@ function run()
   const solver = eval(parse(args["solver"]))
 
   x = Float64[]
-  const times = 100
+  const times = 20
 
-  #sum_maxDegree = Int64[]
+  sum_maxDegree = Int64[]
   #sum_avgDegree = Float64[]
 
 
@@ -55,10 +55,10 @@ function run()
     const problem = generate(problemGenerator)
     #println("[Seed:$(i)] $(string(problem))")
 
-    #const graph = makeKsatDependencyGraph(problem)
-    #max = maxDegree(graph)
-    #println("$(max)")
-    #push!(sum_maxDegree,max)
+    const graph = makeKsatDependencyGraph(problem)
+    max = maxDegree(graph)
+    println("[Seed:$(i)] $(max)")
+    push!(sum_maxDegree,max)
     
     #push!(sum_avgDegree,totalDegree(graph)/length(graph.nodes))
 
@@ -70,16 +70,16 @@ function run()
 
   end
 
-  #sort!(sum_maxDegree)
+  sort!(sum_maxDegree)
   #sort!(sum_avgDegree)
   
   sort!(x)
 
-  #println("$(string(sum_maxDegree[1]))\t$(string(sum_maxDegree[2]))\t$(string(sum_maxDegree[3]))\t$(string(sum_maxDegree[4]))\t$(string(sum_maxDegree[10]))")
+  println("$(string(sum_maxDegree[1]))\t$(string(sum_maxDegree[3]))\t$(string(sum_maxDegree[10]))\t$(string(sum_maxDegree[18]))\t$(string(sum_maxDegree[20]))")
 
-  #println("$(string(sum_avgDegree[1]))\t$(string(sum_avgDegree[2]))\t$(string(sum_avgDegree[5]))\t$(string(sum_avgDegree[9]))\t$(string(sum_avgDegree[10]))")
+  #println("$(string(sum_avgDegree[1]))\t$(string(sum_avgDegree[3]))\t$(string(sum_avgDegree[10]))\t$(string(sum_avgDegree[18]))\t$(string(sum_avgDegree[20]))")
 
-  println("$(string(x[1]))\t$(string(x[10]))\t$(string(x[50]))\t$(string(x[90]))\t$(string(x[100]))")
+  println("$(string(x[1]))\t$(string(x[3]))\t$(string(x[10]))\t$(string(x[18]))\t$(string(x[20]))")
 
 
 

### 2015-05-11 julia/Solvers/ExhaustiveKsatSolver.jl

@@ -10,10 +10,10 @@ function solve(this:: ExhaustiveKsatSolver, problem:: KsatProblem)
 
 
 #  const numPotentialSolutions = 2^n
-  if n < 12
+  if n < 11
     const numPotentialSolutions = 2^n
   else
-    const numPotentialSolutions = 2^12
+    const numPotentialSolutions = 2^11
   end
 
   for binaryAssignment:: Uint64 = 1:numPotentialSolutions

### 2015-05-11 julia/run_solution.sh

@@ -2,7 +2,7 @@
 
 for k in 8
 do
-    for m in 80 
+    for m in 1080 1280
     do
 	echo "k=$k m=$m"
         for n in 32

### 2015-05-11 julia/run_solution_parallel_SimpleChungIndependentSet.sh

@@ -2,14 +2,14 @@
 
 for k in 8
 do
-    for m in 880
+    for m in 1280
     do
 	echo "k=$k m=$m"
         for n in 32
         do
         	for run in {1..10}
         	do
-            	./run ./Applications/KsatTests.jl -s "ParallelLllKsatSolver(SimpleChungIndependentSetFinder())" -d "RandomKsatGenerator($k,$n,$m)" -e 49 -l $run
+            	./run ./Applications/KsatTests.jl -s "ParallelLllKsatSolver(SimpleChungIndependentSetFinder())" -d "RandomKsatGenerator($k,$n,$m)" -e 77 -l $run
         	done
         done
     done

### 2015-05-11 mpi/ExhaustiveMpiKsatSolver.py
Add a sampling solver.

@@ -4,8 +4,10 @@ from MpiKsatSolver import MpiKsatSolver
 from MpiChunks import MpiChunks
 import MpiCollection
 from MpiUtils import onMaster
+import MpiUtils
 import MpiKsatAssignment
 import KsatSolution
+import Utils
 
 def collectAndCombineLocalAssignments(comm, locallySatisfyingAssignments):
   allLocalAssignments = locallySatisfyingAssignments.collect()
@@ -105,6 +107,49 @@ class ExhaustiveProcessParallelSolver(MpiKsatSolver):
     def buildSolution():
       numSatisfyingAssignments, arbitrarySatisfyingAssignment = resultStatistics
       print "%d out of %d satisfying assignments." % (numSatisfyingAssignments, numPotentialSolutions)
+      if arbitrarySatisfyingAssignment > -1:
+        return KsatSolution.success(MpiKsatAssignment.unpack(arbitrarySatisfyingAssignment, n))
+      else:
+        return KsatSolution.failure()
+    
+    return onMaster(problem.comm(), buildSolution)
+
+# A solver that broadcasts clauses and partitions a random subset of the
+# set of assignments across machines.  Does essentially no communication.  Not
+# guaranteed to find a solution if one exists.
+#TODO: Refactor to share code with the exhaustive version.
+class SamplingProcessParallelSolver(MpiKsatSolver):
+  def __init__(self, numSamples):
+    self.numSamples = numSamples
+    super(SamplingProcessParallelSolver, self).__init__()
+  
+  def solve(self, rand, problem):
+    comm = problem.comm()
+    n = problem.numVariables
+    numPotentialSolutions = 2**n
+    numSamples = self.numSamples
+    startingSeed = Utils.randLargeInt(rand)
+    binaryAssignments = (MpiCollection.makeRange(comm, numSamples)
+      .map(lambda idx: MpiUtils.makeRandForItem(startingSeed, idx).randint(0, numPotentialSolutions)))
+    currentAssignment = MpiKsatAssignment.emptyMpiKsatAssignment(problem.comm(), n)
+    currentLocalAssignment = currentAssignment.localAssignment()
+    #FIXME: collectEverywhere() probably should return a list, not an iterable.
+    allClauses = list(problem.distributedClauses().values().collectEverywhere())
+    
+    if comm.rank == 0:
+      print allClauses
+    
[…]

### 2015-05-11 mpi/MpiGraph.py

@@ -1,7 +1,6 @@
 import itertools
 
 import CollectionUtils
-import SetUtils
 import MpiCollection
 
 # A distributed undirected graph.  Nodes are simply integer IDs.  Nodes are 

### 2015-05-11 mpi/IndependentSetFinder.py
Fix an issue with random seeds on multiple machines in the LLL solver.

@@ -1,26 +1,29 @@
 import math
 import itertools
+import operator
 
 import MpiGraph
 import MpiCollection
+import MpiUtils
 import Utils
 
 # Returns an independent set (in the form of an MpiSet) in @graph found 
 # by one iteration of Luby's algorithm.
 def lubyIndependentSet(rand, graph):
-  marks = markNodesRandomly(graph, lambda : Utils.randLargeInt(rand))
+  marks = markNodesRandomly(graph, lambda node: Utils.randLargeInt(rand))
   broadcastMarks = marks.collectEverywhere()
   return findLocalMinima(graph, broadcastMarks)
 
 def markNodesRandomly(graph, distribution):
-  return graph.nodesIterator().map(lambda node: (node, distribution())).toDict()
+  return graph.nodesIterator().map(lambda node: (node, distribution(node))).toDict()
 
 def findLocalMinima(graph, marks):
-  return findLocalMaxima(graph, marks, lambda x, y: x < y)
+  return findLocalMaxima(graph, marks, operator.lt)
 
 def findLocalMaxima(graph, marks, comparatorFunc):
   def isLocalMaximum(node):
     mark = marks[node]
+    # print "Checking whether %d is a local maximum: mark = %d, neighbors = %s, comparisons = %s" % (node, mark, {neighbor: marks[neighbor] for neighbor in graph.neighborsIterator(node)}, {neighbor: comparatorFunc(mark, marks[neighbor]) for neighbor in graph.neighborsIterator(node)})
     return all((comparatorFunc(mark, marks[neighbor]) for neighbor in graph.neighborsIterator(node)))
   return graph.filter(isLocalMaximum).nodesIterator()
 
@@ -76,7 +79,8 @@ class SimpleChungIndependentSetFinder(IndependentSetFinder):
       return max(100, int(math.ceil(math.log(base, m))))
   
   def buildFinderFunc(self, rand, graph):
-    marks = markNodesRandomly(graph, lambda : Utils.randLargeInt(rand))
+    startingSeed = Utils.randLargeInt(rand)
+    marks = markNodesRandomly(graph, lambda nodeIdx: MpiUtils.makeRandForItem(startingSeed, nodeIdx))
     broadcastMarks = marks.collectEverywhere()
     return lambda subgraph: findLocalMinima(subgraph, broadcastMarks)
 

### 2015-05-11 mpi/LllKsatSolver.py

@@ -6,6 +6,7 @@ import MpiCollection
 import MpiKsatAssignment
 import KsatSolution
 import Utils
+import CollectionUtils
 
 # A class of LLL algorithms for solving distributed k-SAT problems.  Solvers
 # are parameterized by a method for finding independent sets in clause
@@ -29,8 +30,11 @@ class LllKsatSolver(MpiKsatSolver):
       # is consistent across processors.
       localAssignment = currentAssignment.localAssignment()
       unsatSubgraph = graph.filter(lambda clauseIdx: not localAssignment.satisfiesClause(broadcastProblem.clausesByIdx[clauseIdx]))
+      # print "%d unsatisfied clauses on iteration %d on machine %d: %s." % (CollectionUtils.iterlen(unsatSubgraph.nodesIterator().localElements()), i, comm.rank, {clauseIdx: broadcastProblem.clausesByIdx[clauseIdx] for clauseIdx in unsatSubgraph.nodesIterator().localElements()})
       localIndependentSet = independentSetFunc(unsatSubgraph)
       locallyModifiedVariables = MpiCollection.dictFromLocal(comm, {})
+      #FIXME
+      localIndependentSetSize = 0
       for clauseIdx in localIndependentSet.localElements():
         clause = broadcastProblem.clausesByIdx[clauseIdx]
         #TODO: Only communicate modified variables to machines that need them.
@@ -39,6 +43,8 @@ class LllKsatSolver(MpiKsatSolver):
         # machine is large).
         updates = {variable: Utils.randBit(rand) for variable in clause.literals}
         locallyModifiedVariables.localDict().update(updates)
+        localIndependentSetSize += 1
+      # print "Using an independent set of size %d on iteration %d on machine %d." % (localIndependentSetSize, i, comm.rank)
       #TODO: Probably inefficient serialization here.  Should serialize as
       # a list of ints and a list of bools.
       allModifiedVariables = locallyModifiedVariables.collectEverywhere()

### 2015-05-11 mpi/MpiBroadcast.py

@@ -15,3 +15,10 @@ class MpiMutableBroadcast(object):
   
   def comm(self):
     return self.chunks.comm()
+
+# Broadcast a value from processor 0 to every processor.
+def broadcast(comm, localValueThunk):
+  chunks = MpiChunks(comm, localValueThunk() if comm.rank == 0 else None)
+  allChunks = chunks.collectEverywhere()
+  broadcastChunk = allChunks[0]
+  return MpiMutableBroadcast(comm, broadcastChunk)

### 2015-05-11 mpi/MpiChunks.py

@@ -4,7 +4,8 @@ from MpiUtils import onMaster
 # with a local chunk of arbitrary data.
 # Currently none of the operations on this object are lazy, so
 # mutating it is okay, subject to the usual issues with distributed data
-# structures.
+# structures.  However, lazy collections may be implemented on top of this
+# by passing lazy mappers to mapChunks().
 class MpiChunks(object):
   def __init__(self, comm, localChunk):
     self.comm = comm

### 2015-05-11 mpi/MpiCollection.py

@@ -1,4 +1,5 @@
 import itertools
+import operator
 
 import Utils
 import CollectionUtils
@@ -44,6 +45,11 @@ class MpiCollection(object):
     collectedChunks = self.distributedChunks.collect()
     return onMaster(self.comm(), lambda : itertools.chain.from_iterable(collectedChunks))
 
+  # Should be called on all processors.  Will return the entire collection
+  # everywhere (a broadcast).
+  def collectEverywhere(self):
+    return itertools.chain.from_iterable(self.distributedChunks.collectEverywhere())
+
   def reduce(self, zero, plus):
     return self.mapChunks(lambda chunk: reduce(plus, chunk, zero)).reduce(zero, plus)
 
@@ -51,21 +57,11 @@ class MpiCollection(object):
     return self.mapChunks(lambda chunk: reduce(plus, chunk, zero)).reduceEverywhere(zero, plus)
 
   def size(self):
-    numElements = self.mapChunks(lambda chunk: len(chunk)).reduce(0, lambda x, y: x + y)
+    numElements = self.mapChunks(_chunkLen).reduce(0, operator.add)
     return onMaster(self.comm(), lambda : numElements)
 
   def sizeEverywhere(self):
-    return self.mapChunks(lambda chunk: len(chunk)).reduceEverywhere(0, lambda x, y: x + y)
-
-  # Should be called on all processors.  Will return the entire collection
-  # everywhere.
-  def collectEverywhere(self):
-    return itertools.chain.from_iterable(self.distributedChunks.collectEverywhere())
-
-  # A new MpiCollection, with each local collection containing all the elements
-  # in @self.
-  def broadcast(self):
-    return collectionFromLocal(self.comm(), self.collectEverywhere())
+    return self.mapChunks(_chunkLen).reduceEverywhere(0, operator.add)
 
   def mapChunks(self, f):
     return self.distributedChunks.mapChunks(f)
@@ -85,6 +81,9 @@ class MpiCollection(object):
 
   def comm(self):
     return self.distributedChunks.comm
+
+def _chunkLen(chunk):
+  return len(chunk) if type(chunk) == set or type(chunk) == list else CollectionUtils.iterlen(chunk)
 
 def collectionFromLocal(comm, localCollection):
   return MpiCollection(MpiChunks(comm, localCollection))

### 2015-05-11 mpi/MpiKsatProblem.py

@@ -5,6 +5,8 @@ import sys
 import MpiCollection
 import MpiUtils
 import MpiGraph
+import MpiBroadcast
+import MpiUtils
 import Utils
 import CollectionUtils
 
@@ -63,6 +65,14 @@ class MpiKsatProblem(object):
     
   def localClausesByIdx(self):
     return self.distributedClausesV.localDict()
+  
+  def isSatisfiedBy(self, assignmentThunk):
+    broadcastAssignment = MpiBroadcast.broadcast(self.comm(), assignmentThunk)
+    isSatisfied = (self.distributedClausesV
+      .values()
+      .map(lambda clause: broadcastAssignment.value().satisfiesClause(clause))
+      .reduce(True, lambda x, y: x and y))
+    return MpiUtils.onMaster(self.comm(), lambda : isSatisfied)
   
   def toBroadcast(self):
     allClausesByIdx = self.distributedClausesV.collectEverywhere()

### 2015-05-11 mpi/MpiUtils.py

@@ -6,7 +6,7 @@ import sys
 # machines.  Useful if we want to run a deterministic distributed job without
 # actually using the same seed on every machine.
 def reseed(comm, rand):
-  return np.random.RandomState(rand.randint(0, sys.maxint) + 53*comm.rank)
+  return np.random.RandomState(int((rand.randint(0, sys.maxint) + 53*comm.rank) % sys.maxint))
 
 # Return a new RandomState whose seed is a deterministic function of @seed
 # and of @idx.  Useful for generating a range of random values that does not
@@ -14,7 +14,7 @@ def reseed(comm, rand):
 #   seed = rand.randint()
 #   makeRange(10).map(lambda item: makeRandForItem(seed, item))
 def makeRandForItem(seed, idx):
-  return np.random.RandomState(seed + 59*idx)
+  return np.random.RandomState(int((seed + 7*idx) % sys.maxint))
 
 # Execute some computation embodied in @thunk on the master, or return None
 # on slaves.  This is just a convenience method for this common pattern in MPI

### 2015-05-11 mpi/ParallelKsatApplication.py

@@ -26,11 +26,18 @@ def run():
   problem = eval(args.generator).generate(comm, problemRand)
   solver = eval(args.solver)
 
+  #NOTE: The solver will perform differently for different numbers of machines,
+  # even if the randoms seed is the same.
   solverRand = np.random.RandomState(args.solverSeed if args.solverSeed is not None else None)
   solution = solver.solve(solverRand, problem)
   
   if comm.rank == 0:
     print "Finished running solver %s; found %s" % (args.solver, solution)
+  
+  satisfactionCheck = problem.isSatisfiedBy(lambda : solution.assignment())
+  if comm.rank == 0:
+    if satisfactionCheck != solution.isSuccessful():
+      print "Solution check failure: Solver claimed success was %s but was %s!" % (solution.isSuccessful(), satisfactionCheck)
 
 if __name__ == "__main__":
   run()

### 2015-05-11 mpi/MpiBroadcast.py
Fix a bug that caused a null dereference when no solution was found.

@@ -17,8 +17,8 @@ class MpiMutableBroadcast(object):
     return self.chunks.comm()
 
 # Broadcast a value from processor 0 to every processor.
-def broadcast(comm, localValueThunk):
-  chunks = MpiChunks(comm, localValueThunk() if comm.rank == 0 else None)
+def broadcast(comm, localValue):
+  chunks = MpiChunks(comm, localValue)
   allChunks = chunks.collectEverywhere()
   broadcastChunk = allChunks[0]
   return MpiMutableBroadcast(comm, broadcastChunk)

### 2015-05-11 mpi/MpiKsatProblem.py

@@ -66,11 +66,11 @@ class MpiKsatProblem(object):
   def localClausesByIdx(self):
     return self.distributedClausesV.localDict()
   
-  def isSatisfiedBy(self, assignmentThunk):
-    broadcastAssignment = MpiBroadcast.broadcast(self.comm(), assignmentThunk)
+  def isSatisfiedBy(self, solution):
+    broadcastSolution = MpiBroadcast.broadcast(self.comm(), solution)
     isSatisfied = (self.distributedClausesV
       .values()
-      .map(lambda clause: broadcastAssignment.value().satisfiesClause(clause))
+      .map(lambda clause: broadcastSolution.value().assignment() is not None and broadcastSolution.value().assignment().satisfiesClause(clause))
       .reduce(True, lambda x, y: x and y))
     return MpiUtils.onMaster(self.comm(), lambda : isSatisfied)
   

### 2015-05-11 mpi/ParallelKsatApplication.py

@@ -34,7 +34,7 @@ def run():
   if comm.rank == 0:
     print "Finished running solver %s; found %s" % (args.solver, solution)
   
-  satisfactionCheck = problem.isSatisfiedBy(lambda : solution.assignment())
+  satisfactionCheck = problem.isSatisfiedBy(solution)
   if comm.rank == 0:
     if satisfactionCheck != solution.isSuccessful():
       print "Solution check failure: Solver claimed success was %s but was %s!" % (solution.isSuccessful(), satisfactionCheck)

### 2015-05-11 mpi/ExhaustiveMpiKsatSolver.py
Make sampling solver work with more than 62 variables.

@@ -126,32 +126,31 @@ class SamplingProcessParallelSolver(MpiKsatSolver):
   def solve(self, rand, problem):
     comm = problem.comm()
     n = problem.numVariables
-    numPotentialSolutions = 2**n
     numSamples = self.numSamples
+    processorRand = MpiUtils.reseed(comm, rand)
     startingSeed = Utils.randLargeInt(rand)
-    binaryAssignments = (MpiCollection.makeRange(comm, numSamples)
-      .map(lambda idx: MpiUtils.makeRandForItem(startingSeed, idx).randint(0, numPotentialSolutions)))
-    currentAssignment = MpiKsatAssignment.emptyMpiKsatAssignment(problem.comm(), n)
-    currentLocalAssignment = currentAssignment.localAssignment()
-    #FIXME: collectEverywhere() probably should return a list, not an iterable.
     allClauses = list(problem.distributedClauses().values().collectEverywhere())
+    satisfyingAssignments = (MpiCollection.makeRange(comm, numSamples)
+      .map(lambda idx: MpiKsatAssignment.uniformRandomKsatAssignment(processorRand, n))
+      .filter(lambda assignment: assignment.satisfiesClauses(allClauses)))
     
     if comm.rank == 0:
       print allClauses
     
-    def checkBinaryAssignment(binaryAssignment):
-      MpiKsatAssignment.unpackTo(binaryAssignment, currentLocalAssignment, n)
-      return currentLocalAssignment.satisfiesClauses(allClauses)
-    satisfyingBinaryAssignments = binaryAssignments.filter(checkBinaryAssignment)
-    resultStatistics = (satisfyingBinaryAssignments
+    def reduceResultStatistics(countAndArbitraryElementA, countAndArbitraryElementB):
+      newCount = countAndArbitraryElementA[0] + countAndArbitraryElementB[0]
+      newArbitraryElement = countAndArbitraryElementA[1] if countAndArbitraryElementB[1] is None else countAndArbitraryElementB[1]
+      return (newCount, newArbitraryElement)
+    
+    resultStatistics = (satisfyingAssignments
       .map(lambda elt: (1, elt))
-      .reduce((0, -1), lambda countAndMaxA, countAndMaxB: (countAndMaxA[0] + countAndMaxB[0], max(countAndMaxA[1], countAndMaxB[1]))))
+      .reduce((0, None), reduceResultStatistics))
     
     def buildSolution():
[…]

### 2015-05-11 mpi/ExhaustiveMpiKsatSolver.py
Minor cleanup.

@@ -35,6 +35,7 @@ class ExhaustiveDataParallelSolver(MpiKsatSolver):
       print list(allClauses)
     
     n = problem.numVariables
+    assert n < 63
     numPotentialSolutions = 2**n
     locallySatisfyingAssignments = MpiChunks(problem.comm(), [])
     assignment = MpiKsatAssignment.emptyMpiKsatAssignment(problem.comm(), n)
@@ -86,6 +87,7 @@ class ExhaustiveProcessParallelSolver(MpiKsatSolver):
   def solve(self, rand, problem):
     comm = problem.comm()
     n = problem.numVariables
+    assert n < 63
     numPotentialSolutions = 2**n
     binaryAssignments = MpiCollection.makeRange(comm, numPotentialSolutions)
     currentAssignment = MpiKsatAssignment.emptyMpiKsatAssignment(problem.comm(), n)

### 2015-05-11 mpi/ExhaustiveMpiKsatSolver.py
change for performance measurement

@@ -61,6 +61,7 @@ class ExhaustiveDataParallelSolver(MpiKsatSolver):
     
     def buildSolution():
       print "%d out of %d satisfying assignments." % (numSatisfyingAssignments, numPotentialSolutions)
+      
       if satisfyingAssignment is not None:
         return KsatSolution.success(satisfyingAssignment)
       else:
@@ -136,8 +137,8 @@ class SamplingProcessParallelSolver(MpiKsatSolver):
       .map(lambda idx: MpiKsatAssignment.uniformRandomKsatAssignment(processorRand, n))
       .filter(lambda assignment: assignment.satisfiesClauses(allClauses)))
     
-    if comm.rank == 0:
-      print allClauses
+    #if comm.rank == 0:
+      #print allClauses
     
     def reduceResultStatistics(countAndArbitraryElementA, countAndArbitraryElementB):
       newCount = countAndArbitraryElementA[0] + countAndArbitraryElementB[0]
@@ -150,7 +151,10 @@ class SamplingProcessParallelSolver(MpiKsatSolver):
     
     def buildSolution():
       numSatisfyingAssignments, arbitrarySatisfyingAssignment = resultStatistics
-      print "%d out of %d satisfying assignments." % (numSatisfyingAssignments, numSamples)
+      #print "%d out of %d satisfying assignments." % (numSatisfyingAssignments, numSamples)
+      if numSatisfyingAssignments*1.0/numSamples <= 0.0002 and numSatisfyingAssignments != 0:
+        print "problem difficulty = " + str(numSatisfyingAssignments*1.0/numSamples)
+      
       if arbitrarySatisfyingAssignment > -1:
         return KsatSolution.success(arbitrarySatisfyingAssignment)
       else:

### 2015-05-11 mpi/LllKsatSolver.py

@@ -7,6 +7,7 @@ import MpiKsatAssignment
 import KsatSolution
 import Utils
 import CollectionUtils
+import datetime
 
 # A class of LLL algorithms for solving distributed k-SAT problems.  Solvers
 # are parameterized by a method for finding independent sets in clause
@@ -19,12 +20,31 @@ class LllKsatSolver(MpiKsatSolver):
   def solve(self, rand, problem):
     comm = problem.comm()
     n = problem.numVariables
+
+    
+    #broadcastStartTime = now()
+    broadcastStartTime = datetime.datetime.now()
     broadcastProblem = problem.toBroadcast()
+    broadcastEndTime = datetime.datetime.now()
+
+    #graphCreationStartTime = now()
+    graphCreationStartTime = datetime.datetime.now()
     graph = broadcastProblem.toDependencyGraph(comm)
+    graphCreationEndTime = datetime.datetime.now()
+    
     currentAssignment = MpiKsatAssignment.uniformRandomMpiKsatAssignment(comm, rand, n)
     maxNumIterations = self.independentSetFinder.calculateNumIterations(problem, graph)
     if comm.rank == 0: print "Solving problem with %s iterations." % maxNumIterations
+
+    #initialMarkingsStartTime = now()
+    initialMarkingsStartTime = datetime.datetime.now()
     independentSetFunc = self.independentSetFinder.buildFinderFunc(rand, graph)
+    initialMarkingsEndTime = datetime.datetime.now()
+
+
+    findIndependentSetAndResampleTotalTime = datetime.timedelta()
+    broadcastModifiedVariablesTotalTime = datetime.timedelta()
+
     for i in xrange(maxNumIterations):
       # At the beginning of each iteration of this loop, @currentAssignment
       # is consistent across processors.
@@ -35,6 +55,9 @@ class LllKsatSolver(MpiKsatSolver):
       locallyModifiedVariables = MpiCollection.dictFromLocal(comm, {})
       #FIXME
       localIndependentSetSize = 0
+      
+      #findIndependentSetAndResampleStartTime = now()
+      findIndependentSetAndResampleStartTime = datetime.datetime.now()
       for clauseIdx in localIndependentSet.localElements():
         clause = broadcastProblem.clausesByIdx[clauseIdx]
         #TODO: Only communicate modified variables to machines that need them.
@@ -44,12 +67,25 @@ class LllKsatSolver(MpiKsatSolver):
[…]

### 2015-05-11 mpi/ParallelKsatApplication.py

@@ -1,6 +1,7 @@
 import argparse
 from mpi4py import MPI
 import numpy as np
+import datetime
 
 import MpiKsatProblem
 import ExhaustiveMpiKsatSolver
@@ -29,10 +30,16 @@ def run():
   #NOTE: The solver will perform differently for different numbers of machines,
   # even if the randoms seed is the same.
   solverRand = np.random.RandomState(args.solverSeed if args.solverSeed is not None else None)
+
+  wholeSolveStartTime = datetime.datetime.now()
   solution = solver.solve(solverRand, problem)
+  wholeSolveEndTime = datetime.datetime.now()
+
+  #if comm.rank == 0:
+  #  print "\nTotal Solving Time is " + str((wholeSolveEndTime - wholeSolveStartTime).total_seconds()) + "\n"
   
-  if comm.rank == 0:
-    print "Finished running solver %s; found %s" % (args.solver, solution)
+  #if comm.rank == 0:
+    #print "Finished running solver %s; found %s" % (args.solver, solution)
   
   satisfactionCheck = problem.isSatisfiedBy(solution)
   if comm.rank == 0:

### 2015-05-12 mpi/MpiUtils.py
Fix random seed generation for machines with 32-bit numpy.

@@ -1,12 +1,16 @@
 import numpy as np
 import sys
 
+# Some versions of numpy only support seeds up to this.  This is unfortunate,
+# since collisions can make our algorithm fail.
+MAXIMUM_RANDOM_SEED = 2**32-1
+
 # Return a new RandomState whose seed is a deterministic function of the
 # seed in @rand and the machine ID, and which is different for different
 # machines.  Useful if we want to run a deterministic distributed job without
 # actually using the same seed on every machine.
 def reseed(comm, rand):
-  return np.random.RandomState(int((rand.randint(0, sys.maxint) + 53*comm.rank) % sys.maxint))
+  return np.random.RandomState(int((rand.randint(0, sys.maxint) + 53*comm.rank) % MAXIMUM_RANDOM_SEED))
 
 # Return a new RandomState whose seed is a deterministic function of @seed
 # and of @idx.  Useful for generating a range of random values that does not
@@ -14,7 +18,7 @@ def reseed(comm, rand):
 #   seed = rand.randint()
 #   makeRange(10).map(lambda item: makeRandForItem(seed, item))
 def makeRandForItem(seed, idx):
-  return np.random.RandomState(int((seed + 7*idx) % sys.maxint))
+  return np.random.RandomState(int((seed + 7*idx) % MAXIMUM_RANDOM_SEED))
 
 # Execute some computation embodied in @thunk on the master, or return None
 # on slaves.  This is just a convenience method for this common pattern in MPI

### 2015-05-12 paper/paper.tex
Final Report
Partial work on the paper.

@@
 \usepackage[retainorgcmds]{IEEEtrantools}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
 \usepackage{filecontents}
@@
 \author{Derek Kuo, Henry Milner}
-\title{CS267 HW2: Particle Simulation}
-\date{3/13/15}
+\title{CS267 Final Project: An Empirical Investigation of the Algorithmic Lovász Local Lemma}
+\date{5/11/15}
+
+% Some macros specific to this paper.
+\newcommand{\ksat}{\texttt{k-SAT}}
+
+\newcommand{\polylog}{\text{polylog}}
+
+\newcommand{\lovasz}{Lov\'{a}sz}
 
@@
 }
+
 @article{gomes2008satisfiability,
@@
 }
+
+@article{belov2014sat,
+  title={SAT COMPETITION 2014},
+  author={Belov, Anton and Diepold, Daniel and Heule, Marijn JH and J{\"a}rvisalo, Matti},
+  year={2014}
+}
+
 \end{filecontents}
@@
 \begin{abstract}
-The algorithmic Lovasz Local Lemma (LLL) is a recent-developed meta-algorithm for solving certain instances of hard combinatorial problems.
-In theory, LLL algorithms can be applied whenever a problem can be decomposed into smaller subproblems and the dependencies among subproblems are somewhat sparse.
-For example, a k-SAT problem can be decomposed into the problem of satisfying each of its clauses, and LLL algorithms can be useful in solving k-SAT problems whose clauses do not share literals with too many other clauses.
+The algorithmic \lovasz~ Local Lemma (LLL) is a recent-developed meta-algorithm for solving certain instances of hard combinatorial problems.
+In theory, LLL algorithms can be applied whenever a problem can be decomposed into smaller subproblems and the dependencies among subproblems are, in a certain sense, somewhat sparse.
+For example, a \ksat problem can be decomposed into the problem of satisfying each of its clauses, and LLL algorithms can be useful in solving \ksat problems whose clauses do not share literals with too many other clauses.
 The seminal paper in the field is by Moser and Tardos \cite{moser2010constructive} in 2010.
-For many problems, LLL algorithms theoretically lend themselves well to parallelism, as demonstrated in \cite{moser2010constructive} and expanded upon in more recent work \cite{chung2014distributed,haeupler2011new}.
-However, we are aware of no published empirical tests of the efficiency of LLL algorithms in practice.
-In this paper, we give practical implementations of parallel LLL algorithms for k-SAT, and we investigate their performance.
+For many problems, LLL algorithms theoretically lend themselves well to parallelism, as demonstrated in \cite{moser2010constructive} and expanded upon in more recent work \cite{chung2014distributed,haeupler2011new} that focuses in particular on distributed algorithms.
+However, we are aware of no published empirical tests of the efficiency of LLL algorithms in practice; the existing literature focuses entirely on proving results about their asymptotic performance.
+In this paper, we give concrete implementations of serial and distributed LLL algorithms for \ksat, and we give a preliminary evaluation of their performance.
+\end{abstract}
 
 \section{Introduction}
+\label{sec:intro}
+In 1975, Erd\"{o}s and \lovasz~ \cite{erdos1975problems} developed a technique, now called the \lovasz~ Local Lemma (LLL), for proving the existence of solutions to combinatorial problems.
+The technique is a ``probabilistic method'': Given a set of events that we want to avoid (for example, each event could correspond to the violations of a problem constraint), it is proved that there is a strictly positive probability that none of these events happen under a random potential solution drawn from a suitable distribution on potential solutions.
+The main precondition of the LLL is that the statistical dependencies among events are \emph{sparse}, with the sparsity being of the same magnitude as the marginal probability of each single event being unsatisfied under the sampling distribution.
+We will discuss the technique later in more detail.
 
-\subsection{This paper}
+For many years the LLL was viewed as a tool useful only in proofs.
+In 1991, Beck \cite{beck1991algorithmic} found randomized algorithms for several combinatorial problems that actually \emph{find} solutions, and which finish in expected polynomial time under conditions similar to the preconditions of the LLL.
+The CS theory community continued to improve on these results, until in 2010 Moser and Tardos \cite{moser2010constructive} published a meta-algorithm for solving, in expected polynomial time, any combinatorial problem that satisfies the LLL preconditions.
+
+An interesting aspect of the Moser-Tardos meta-algorithm is that it can be parallelized.
+Moser and Tardos analyzed a simple parallel version of their algorithm in their original paper.
+More recently, Chung et al \cite{chung2014distributed} and Haeupler et al \cite{haeupler2011new} developed algorithms requiring less communication in a distributed environment.
+All of these algorithms finish in $\tilde{O}(1)$ time given a number of processors that grows linearly with the problem description length.
+However, the big-$\tilde{O}$ hides logarithmic factors and potentially large constants.
+
+Though they seem promising, to our knowledge neither the Moser-Tardos meta-algorithm nor its parallel variants have ever been tested empirically.
+Several questions of practical interest are left open by the existing literature:
+\begin{enumerate}
+  \item When the preconditions of the LLL are satisfied, does the Moser-Tardos meta-algorithm actually run quickly, or does the analysis hide large constant factors?
+  \item Under those conditions, is the Moser-Tardos meta-algorithm competitive with existing solvers for hard combinatorial problems?
+  \item Are there interesting and difficult problems that satisfy the preconditions of the LLL?
+  \item Are the preconditions of the LLL actually necessary in practice for the Moser-Tardos meta-algorithm to work efficiently?
+How quickly does its performance degrade as the preconditions become more unsatisfied?
+  \item Are the parallel LLL algorithms useful in practice?
+\end{enumerate}
+
+In this paper, we study the last question: Are the parallel LLL algorithms useful in practice?
+In particular, we consider algorithms designed for \emph{distributed-memory} systems, though shared-memory parallelism and GPUs would both be interesting future directions.
+In order to answer our question it will be necessary to give partial answers to the others, but we leave more thorough investigations for future work.
+
+To study applications of the algorithmic LLL, we must choose combinatorial problems to solve.
+We focus on \ksat.
+In \ksat, we are given $n$ boolean variables and $m$ disjunctive (``OR'') clauses of $k$ possibly-negated variables each, and we must find an assignment of the variables that satisfies all of the clauses.
+The decision version of \ksat is NP-complete, but this does not necessarily mean that instances of practical importance (or, possibly, all but a small number of worst-case instances) are hard, and there is a long and ongoing history of work on heuristic search methods for the problem.
+In fact, there is an annual competition for SAT solvers \cite{belov2014sat}.
+(Our benchmarks will be based on the guidelines in that competition.)
+
+\subsection{This Paper}
+In section \ref{sec:alll}, we discuss in detail the Moser-Tardos meta-algorithm.
+In section \ref{sec:implementation}, we describe our implementation of particular forms of the algorithm and the design choices we made.
+In section \ref{sec:benchmarks} we comment on the design of benchmarks, which is nontrivial for \ksat.
+In section \ref{sec:scaling}, we present scaling results for our chosen benchmarks.
+We conclude in section \ref{sec:conclusion} with a discussion of our results and directions for future work.
+
+\section{The Algorithmic \lovasz~ Local Lemma}
+\label{sec:alll}
+
+The basic setup for LLL algorithms is as follows: Let $\mathcal{A} = \{A_1, \cdots, A_m\} \in \{0,1\}^m$ be a set of discrete variables (which we take to be binary for sake of exposition), and let $\mathcal{E} = \{E_1, \cdots, E_n\}$ be a set of functions $\{0,1\}^m \to \{0,1\}$ mapping the variables to truth values.
+The $E_i$ are called ``events,'' and $E_i$ is said to ``happen'' under some assignment to $\mathcal{A}$ if $E_i(\mathcal{A}) = 1$.
+We would like to find an assignment of the variables so that none of the events happen.
+
+This framework encompasses many constraint satisfaction problems.
+For example, we can encode graph coloring as follows: the variables are binary encodings of the color assignments to vertices, and event $E_{ij}$ happens if vertices $i$ and $j$ are assigned the same color.
+Then if we find an assignment of the variables for which none of the events happen, we have found a graph coloring.
+Encoding \ksat is even more straightforward: the $A_i$ are the problem variables, and each $E_i$ corresponds to a clause involving $k$ variables.
+$E_i$ happens if the corresponding clause is false, and an assignment solves the problem if none of the events happen.
+In \ksat, each event $E_i$ is a function of only a smaller subset of the variables.
+We denote that subset of $\mathcal{A}$ by $\operatorname{vbl}(E_i)$.
+LLL algorithms typically need $|\operatorname{vbl}(E_i)| << |\mathcal{A}|$.
+
+The simplest LLL algorithm is \emph{extremely} simple.  It is Algorithm \ref{alg:simple-lll} below.
+
+\begin{algorithm}[H]
+\label{alg:simple-lll}
+\begin{algorithmic}
+\State Initialize all the $A_i$ as IID samples from a uniform Bernoulli distribution.
+\While{Any of the $E_j$ happen under the current value of $\mathcal{A}$}
+  \State Let $E_j$ be an arbitrary event that happens under $\mathcal{A}$.
+  \State Reample each $A_i \in \operatorname{vlb}(E_j)$ IID uniform Bernoulli.
+\EndWhile
+\end{algorithmic}
+\caption{The simplest LLL algorithm.}
+\end{algorithm}
+
+Of course, the decision version of \ksat is NP-hard, so we would not expect this algorithm to work for all instances.
+Moser and Tardos \cite{moser2010constructive} prove that it runs in expected polynomial time when the events are not individually too hard to satisfy with random assignment and when there are not too many dependencies among the events.
+Concretely:
+\begin{enumerate}
+  \item Let $p$ be the maximum probability that any single event happens under a random assignment to all the variables.
+(In \ksat, this is $2^{-k}$, since a clause is unsatisfied only if all of its k variables are assigned incorrectly, and the single variable assignments happen independently with probability $1/2$.)
+  \item Construct an undirected dependency graph for the events, in which there is an edge between $E_j$ and $E_k$ if they are statistically dependent under random assignment to the $A_i$.
+In \ksat, there is an edge between clauses $E_j$ and $E_k$ if they share any variable.
+Let $d$ be the maximum degree of events in this graph.
+\end{enumerate}
+Moser and Tardos prove that the above algorithm finishes in expected polynomial time if $e p d < 1$, where $e$ is the base of the natural logarithm.
+
+For example, a \ksat problem satisfies this condition if each clause shares variables with no more than $\frac{2^k}{e}$ other clauses.
+If we generate a random \ksat instance with $m$ variables and $n$ clauses, the probability that two clauses share a variable is approximately $\frac{k^2}{m}$, so the expected value of $d$ is approximately $\frac{n k^2}{m}$.
+In that case the proof by Moser and Tardos applies if $n$ is less than $\frac{2^k m}{k^2 e}$.
+For comparison, the maximum number of clauses in a \ksat instance is $2^k {m \choose k}$.
+It is known that random 3-SAT instances are hardest when $n \approxeq 4.26 m$ \cite{gomes2008satisfiability}, so these are relatively easy instances.
+In fact, for $k = 3$ these instances do not even use all the variables, since $\frac{2^3 m}{3^2 e} = \frac{8}{9e} < 1/3$.
+
+The event dependency graph is useful for seeing an easy way to parallelize the simple LLL algorithm.
+Notice that if independent events $E_j$ and $E_k$ happen under the current assignment to $\mathcal{A}$, and event $E_j$'s variables are chosen for resampling, then (by independence) $E_k$ will still happen after the resampling, so the next iteration could resample its variables.
+This means we can resample variables depended-on by any independent set of events that currently happen, and this resampling can be done in parallel.
+The difficult part is finding a large independent set and communicating variable assignments.
+Chung et al \cite{chung2014distributed} and Srinivasan et al \cite{haeupler2011new} give alternative parallel LLL algorithms (similar to Algorithm \ref{alg:simple-lll}, but somewhat more complicated) that can be run more efficiently on distributed computers.
+
+Last, it should be noted that the simple LLL algorithm is quite similar to stochastic local search heuristics for \ksat that have been known for many years.
+It would be surprising if it performed much better than those older methods, or the more sophisticated heuristics that have followed them.
+So we expect to find negative results.
+The parallel LLL algorithms are less obvious, so there is more hope that they outperform or outscale existing algorithms on instances that satisfy Moser's condition for efficiency.
 

### 2015-05-12 mpi/ParallelKsatApplication.py
Temporary debugging code.

@@ -21,18 +21,27 @@ def parseArgs():
   return parser.parse_args()
 
 def run():
+  print "0"
   args = parseArgs()
+  print "1"
   comm = MPI.COMM_WORLD
+  print "2"
   problemRand = np.random.RandomState(args.problemSeed if args.problemSeed is not None else None)
+  print "3"
   problem = eval(args.generator).generate(comm, problemRand)
+  print "4"
   solver = eval(args.solver)
+  print "5"
 
   #NOTE: The solver will perform differently for different numbers of machines,
   # even if the randoms seed is the same.
   solverRand = np.random.RandomState(args.solverSeed if args.solverSeed is not None else None)
+  print "6"
 
   wholeSolveStartTime = datetime.datetime.now()
+  print "7"
   solution = solver.solve(solverRand, problem)
+  print "8"
   wholeSolveEndTime = datetime.datetime.now()
 
   #if comm.rank == 0:

### 2015-05-12 mpi/ExhaustiveMpiKsatSolver.py
performance measurement

@@ -152,8 +152,8 @@ class SamplingProcessParallelSolver(MpiKsatSolver):
     def buildSolution():
       numSatisfyingAssignments, arbitrarySatisfyingAssignment = resultStatistics
       #print "%d out of %d satisfying assignments." % (numSatisfyingAssignments, numSamples)
-      if numSatisfyingAssignments*1.0/numSamples <= 0.0002 and numSatisfyingAssignments != 0:
-        print "problem difficulty = " + str(numSatisfyingAssignments*1.0/numSamples)
+      #if numSatisfyingAssignments*1.0/numSamples <= 0.0002 and numSatisfyingAssignments != 0:
+      print "problem difficulty = " + str(numSatisfyingAssignments*1.0/numSamples)
       
       if arbitrarySatisfyingAssignment > -1:
         return KsatSolution.success(arbitrarySatisfyingAssignment)

### 2015-05-12 mpi/ParallelKsatApplication.py
Performance Measurement

@@ -21,31 +21,31 @@ def parseArgs():
   return parser.parse_args()
 
 def run():
-  print "0"
+#  print "0"
   args = parseArgs()
-  print "1"
+#  print "1"
   comm = MPI.COMM_WORLD
-  print "2"
+#  print "2"
   problemRand = np.random.RandomState(args.problemSeed if args.problemSeed is not None else None)
-  print "3"
+#  print "3"
   problem = eval(args.generator).generate(comm, problemRand)
-  print "4"
+#  print "4"
   solver = eval(args.solver)
-  print "5"
+#  print "5"
 
   #NOTE: The solver will perform differently for different numbers of machines,
   # even if the randoms seed is the same.
   solverRand = np.random.RandomState(args.solverSeed if args.solverSeed is not None else None)
-  print "6"
+#  print "6"
 
   wholeSolveStartTime = datetime.datetime.now()
-  print "7"
+#  print "7"
   solution = solver.solve(solverRand, problem)
-  print "8"
+#  print "8"
   wholeSolveEndTime = datetime.datetime.now()
 
-  #if comm.rank == 0:
-  #  print "\nTotal Solving Time is " + str((wholeSolveEndTime - wholeSolveStartTime).total_seconds()) + "\n"
+  if comm.rank == 0:
+    print "\nTotal Solving Time is " + str((wholeSolveEndTime - wholeSolveStartTime).total_seconds()) + "\n"
   
   #if comm.rank == 0:
     #print "Finished running solver %s; found %s" % (args.solver, solution)

### 2015-05-12 mpi/ParallelKsatApplication.py
push

@@ -21,27 +21,21 @@ def parseArgs():
   return parser.parse_args()
 
 def run():
-  print "0"
   args = parseArgs()
-  print "1"
   comm = MPI.COMM_WORLD
-  print "2"
   problemRand = np.random.RandomState(args.problemSeed if args.problemSeed is not None else None)
-  print "3"
   problem = eval(args.generator).generate(comm, problemRand)
-  print "4"
   solver = eval(args.solver)
-  print "5"
 
   #NOTE: The solver will perform differently for different numbers of machines,
   # even if the randoms seed is the same.
   solverRand = np.random.RandomState(args.solverSeed if args.solverSeed is not None else None)
-  print "6"
+
 
   wholeSolveStartTime = datetime.datetime.now()
-  print "7"
+
   solution = solver.solve(solverRand, problem)
-  print "8"
+
   wholeSolveEndTime = datetime.datetime.now()
 
   #if comm.rank == 0:

### 2015-05-12 paper/paper.tex
Partial work on the paper.

--- 2015-04-29 paper/paper.tex
@@
-\documentclass{article}
+\documentclass[twocolumn]{article}
 \usepackage{amsmath, amssymb}
 \usepackage[retainorgcmds]{IEEEtrantools}
+\usepackage{algorithm}
+\usepackage{algpseudocode}
 \usepackage{filecontents}
@@
 \author{Derek Kuo, Henry Milner}
-\title{CS267 HW2: Particle Simulation}
-\date{3/13/15}
+\title{CS267 Final Project: An Empirical Investigation of the Algorithmic Lovász Local Lemma}
+\date{5/11/15}
+
+% Some macros specific to this paper.
+\newcommand{\ksat}{\texttt{k-SAT}~}
+
+\newcommand{\polylog}{\text{polylog}}
+
+\newcommand{\lovasz}{Lov\'{a}sz~}
 
@@
 }
+@article{alon1986fast,
+  title={A fast and simple randomized parallel algorithm for the maximal independent set problem},
+  author={Alon, Noga and Babai, L{\'a}szl{\'o} and Itai, Alon},
+  journal={Journal of algorithms},
+  volume={7},
+  number={4},
+  pages={567--583},
+  year={1986},
+  publisher={Elsevier}
+}
+@article{luby1986simple,
+  title={A simple parallel algorithm for the maximal independent set problem},
+  author={Luby, Michael},
+  journal={SIAM journal on computing},
+  volume={15},
+  number={4},
+  pages={1036--1053},
+  year={1986},
+  publisher={SIAM}
+}
 @inproceedings{freer2010probabilistic,
@@
 }
+@inproceedings{coja2014asymptotic,
+  title={The asymptotic k-sat threshold},
+  author={Coja-Oghlan, Amin},
+  booktitle={Proceedings of the 46th Annual ACM Symposium on Theory of Computing},
+  pages={804--813},
+  year={2014},
+  organization={ACM}
+}
+
 @article{gomes2008satisfiability,
@@
 \begin{abstract}
-The algorithmic Lovasz Local Lemma (LLL) is a recent-developed meta-algorithm for solving certain instances of hard combinatorial problems.
-In theory, LLL algorithms can be applied whenever a problem can be decomposed into smaller subproblems and the dependencies among subproblems are somewhat sparse.
-For example, a k-SAT problem can be decomposed into the problem of satisfying each of its clauses, and LLL algorithms can be useful in solving k-SAT problems whose clauses do not share literals with too many other clauses.
+The algorithmic \lovasz Local Lemma (LLL) is a recent-developed meta-algorithm for solving certain instances of hard combinatorial problems.
+In theory, LLL algorithms can be applied whenever a problem can be decomposed into smaller subproblems and the dependencies among subproblems are, in a certain sense, somewhat sparse.
+For example, a \ksat problem can be decomposed into the problem of satisfying each of its clauses, and LLL algorithms can be useful in solving \ksat problems whose clauses do not share literals with too many other clauses.
 The seminal paper in the field is by Moser and Tardos \cite{moser2010constructive} in 2010.
-For many problems, LLL algorithms theoretically lend themselves well to parallelism, as demonstrated in \cite{moser2010constructive} and expanded upon in more recent work \cite{chung2014distributed,haeupler2011new}.
-However, we are aware of no published empirical tests of the efficiency of LLL algorithms in practice.
-In this paper, we give practical implementations of parallel LLL algorithms for k-SAT, and we investigate their performance.
+For many problems, LLL algorithms theoretically lend themselves well to parallelism, as demonstrated in \cite{moser2010constructive} and expanded upon in more recent work \cite{chung2014distributed,haeupler2011new} that focuses in particular on distributed algorithms.
+However, we are aware of no published empirical tests of the efficiency of LLL algorithms in practice; the existing literature focuses entirely on proving results about their asymptotic performance.
+In this paper, we give concrete implementations of serial and distributed LLL algorithms for \ksat, and we give a preliminary evaluation of their performance.
+\end{abstract}
 
 \section{Introduction}
+\label{sec:intro}
+In 1975, Erd\"{o}s and \lovasz \cite{erdos1975problems} developed a technique, now called the \lovasz Local Lemma (LLL), for proving the existence of solutions to combinatorial problems.
+The technique is a ``probabilistic method'': Given a set of events that we want to avoid (for example, each event could correspond to the violations of a problem constraint), it is proved that there is a strictly positive probability that none of these events happen under a random potential solution drawn from a suitable distribution on potential solutions.
+The main precondition of the LLL is that the statistical dependencies among events are \emph{sparse}, with the sparsity being of the same magnitude as the marginal probability of each single event being unsatisfied under the sampling distribution.
+We will discuss the technique later in more detail.
 
-\subsection{This paper}
+For many years the LLL was viewed as a tool useful only in proofs.
+In 1991, Beck \cite{beck1991algorithmic} found randomized algorithms for several combinatorial problems that actually \emph{find} solutions, and which finish in expected polynomial time under conditions similar to the preconditions of the LLL.
+The CS theory community continued to improve on these results, until in 2010 Moser and Tardos \cite{moser2010constructive} published a meta-algorithm for solving, in expected polynomial time, any combinatorial problem that satisfies the LLL preconditions.
+
+An interesting aspect of the Moser-Tardos meta-algorithm is that it can be parallelized.
+Moser and Tardos analyzed a simple parallel version of their algorithm in their original paper.
+More recently, Chung et al \cite{chung2014distributed} and Haeupler et al \cite{haeupler2011new} developed algorithms requiring less communication in a distributed environment.
+All of these algorithms finish in $\tilde{O}(1)$ time given a number of processors that grows linearly with the problem description length.
+However, the big-$\tilde{O}$ hides logarithmic factors and potentially large constants.
+
+Though they seem promising, to our knowledge neither the Moser-Tardos meta-algorithm nor its parallel variants have ever been tested empirically.
+Several questions of practical interest are left open by the existing literature:
+\begin{enumerate}
+  \item When the preconditions of the LLL are satisfied, does the Moser-Tardos meta-algorithm actually run quickly, or does the analysis hide large constant factors?
+  \item Under those conditions, is the Moser-Tardos meta-algorithm competitive with existing solvers for hard combinatorial problems?
+  \item Are there interesting and difficult problems that satisfy the preconditions of the LLL?
+  \item Are the preconditions of the LLL actually necessary in practice for the Moser-Tardos meta-algorithm to work efficiently?
+How quickly does its performance degrade as the preconditions become more unsatisfied?
+  \item Are the parallel LLL algorithms useful in practice?
+\end{enumerate}
+
+In this paper, we study only the last question: Are the parallel LLL algorithms useful in practice?
+In particular, we consider algorithms designed for \emph{distributed-memory} systems (though shared-memory parallelism and GPUs would both be interesting future directions).
+In order to answer our question it will be necessary to give partial answers to the others, but we leave more thorough investigations for future work.
+
+To study applications of the algorithmic LLL, we must choose combinatorial problems to solve.
+We focus on \ksat.
+In \ksat, we are given $n$ boolean variables and $m$ disjunctive (``OR'') clauses of $k$ possibly-negated variables each, and we must find an assignment of the variables that satisfies all of the clauses.
+The decision version of \ksat is NP-complete, but this does not necessarily mean that instances of practical importance (or, possibly, all but a small number of worst-case instances) are hard, and there is a long and ongoing history of work on heuristic search methods for the problem.
+In fact, there is an annual competition for SAT solvers \cite{belov2014sat}.
+(Our benchmarks will be based on the guidelines in that competition.)  We do not expect that our implementations will be competitive with state-of-the-art SAT solvers.
+Our interest is primarily in finding out whether there is any practical potential for an algorithm whose development has thus far been guided exclusively by the desire to prove results about its performance.
+
+Our main contribution is a collection of serial and distributed implementations of LLL algorithms for \ksat.
+Our serial code is written in Julia, and our distributed code is written in Python, using MPI via the mpy4py library.
+All code is available at \texttt{https://github.com/henryem/lll} .
+
+\subsection{This Paper}
+In section \ref{sec:alll}, we discuss in detail the Moser-Tardos meta-algorithm.
+In section \ref{sec:implementation}, we describe our implementation of particular forms of the algorithm and the design choices we made.
+In section \ref{sec:benchmarks} we comment on the design of benchmarks, which is nontrivial for \ksat.
+In section \ref{sec:scaling}, we present scaling results for our chosen benchmarks.
+We conclude in section \ref{sec:discussion} with a discussion of our results and directions for future work.
+
+\section{The Algorithmic \lovasz Local Lemma}
+\label{sec:alll}
+LLL algorithms can be useful when a problem can be phrased as follows.
+Let $\mathcal{A} = \{A_1, \cdots, A_n\} \in \{0,1\}^n$ be a set of discrete variables (which we take to be binary for sake of exposition), and let $\mathcal{E} = \{E_1, \cdots, E_m\}$ be a set of functions $\{0,1\}^n \to \{0,1\}$ mapping the variables to truth values.
+The $E_i$ are called ``events,'' and $E_i$ is said to ``happen'' under some assignment to $\mathcal{A}$ if $E_i(\mathcal{A}) = 1$.
+We would like to find an assignment of the variables so that none of the events happen.
+
+This framework encompasses many constraint satisfaction problems.
+For example, we can encode graph coloring as follows: the variables are binary encodings of the color assignments to vertices, and event $E_{ij}$ happens if vertices $i$ and $j$ are assigned the same color.
+Then if we find an assignment of the variables for which none of the events happen, we have found a graph coloring.
+Encoding \ksat is even more straightforward: the $A_i$ are the problem variables, and each $E_i$ corresponds to a clause involving $k$ variables.
+$E_i$ happens if the corresponding clause is false, and an assignment solves the problem if none of the events happen.
+In \ksat, each event $E_i$ is a function of only a smaller subset of the variables (of size $k$), and it is functionally independent of the rest.
+We denote that subset of $\mathcal{A}$ by $\operatorname{vbl}(E_i)$.
+If we imagine randomly and independently determining the variable values, then $E_i$ is statistically dependent only on $\operatorname{vbl}(E_i)$.
+LLL algorithms typically need $|\operatorname{vbl}(E_i)| << |\mathcal{A}|$.
+
+\subsection{The Moser-Tardos meta-algorithm}
+The simplest LLL algorithm, the Moser-Tardos meta-algorithm, is \emph{extremely} simple.  It is algorithm \ref{alg:mt-meta} below.  Its specialization to \ksat is algorithm \ref{alg:mt-ksat}.
+%FIXME: Fix numbering.
+
+\begin{algorithm}[H]
+\label{alg:mt-meta}
+\begin{algorithmic}
+\State Initialize all the $A_i$ as i.i.d. samples from a uniform Bernoulli distribution.
+\While{Any of the $E_j$ happen under the current value of $\mathcal{A}$}
+  \State Let $E_j$ be an arbitrary event that happens under $\mathcal{A}$.
+  \State Resample each $A_i \in \operatorname{vbl}(E_j)$ i.i.d. uniform Bernoulli.
+\EndWhile
+\end{algorithmic}
+\caption{The Moser-Tardos meta-algorithm.}
+\end{algorithm}
+
+\begin{algorithm}[H]
+\label{alg:mt-ksat}
+\begin{algorithmic}
+\State Initialize all variables $V_i$ as i.i.d. samples from a uniform Bernoulli distribution.
+\While{Any clause is violated}
+  \State Let $C_j$ be an arbitrary clause that is violated.
+  \State Resample each $V_i \in \operatorname{vbl}(C_j)$ i.i.d. uniform Bernoulli.
+\EndWhile
+\end{algorithmic}
+\caption{The Moser-Tardos meta-algorithm, specialized to \ksat.}
+\end{algorithm}
+
+\begin{figure}[ht]
+  \includegraphics[scale=0.5]{figures/mt-directed-example.png}
+  \caption{One iteration of an example run of the Moser-Tardos meta-algorithm for \ksat.
+The problem is $(!a|b)\&(b|!d)\&(a|c)$.
+The variables $a$, $b$, $c$, and $d$ are initialized uniformly at random before the algorithm starts.}
+  \label{fig:mt-directed-example}
+\end{figure}
+
+To illustrate the algorithm, it helps to associate an LLL problem with a bipartite \emph{directed probabilistic graph} (sometimes also called a \emph{Bayes' net}).
+Each variable and each event is a node in this graph, and there is a directed edge from each variable to each event that depends on it.
+(The events are deterministic given the variables, but in a randomized algorithm the process that generates the variable values is random.)  Figure \ref{fig:mt-directed-example} shows an example run of the Moser-Tardos meta-algorithm for a \ksat problem through the lense of this directed graph.
+
+Of course, the decision version of \ksat is NP-hard, so we would not expect this algorithm to work well for all instances.
+Moser and Tardos \cite{moser2010constructive} prove that it runs in expected polynomial time when the events are not individually too hard to satisfy with random assignment and when there are not too many dependencies among the events.
+
+To make this concrete, it is useful to associate another graph with an LLL problem: its dependency graph.
+This is an undirected graph in which each event is a node, and in which there is an edge between events $E_j$ and $E_k$ if they are statistically dependent under independent random assignment to the $A_i$, or equivalently if any variable in the directed probablistic graph points to both $E_j$ and $E_k$.
+In \ksat, there is an edge between clauses $E_j$ and $E_k$ if they share any variable.
+Figure \ref{fig:mt-undirected-example} has an example dependency graph for the problem pictured in figure \ref{fig:mt-directed-example}.
+
+\begin{figure}[ht]
+  \includegraphics[scale=0.5]{figures/mt-undirected-example.png}
+  \caption{The dependency graph for the problem pictured in figure \ref{fig:mt-directed-example}.}
+  \label{fig:mt-undirected-example}
+\end{figure}
+
+Let $d$ be the maximum degree of events in the dependency graph for a problem instance.
+And let $p$ be the maximum probability that any single event happens under a uniform random assignment to all the variables.
+In \ksat, this is $2^{-k}$, since a disjunctive clause is unsatisfied only if all of its k variables are assigned incorrectly, and the single variable assignments happen independently with probability $1/2$.
+Then Moser and Tardos prove that algorithm \ref{alg:mt-meta} finishes in expected polynomial time if $e p d < 1$, where $e$ is the base of the natural logarithm.
+These are the same conditions under which the ordinary \lovasz Local Lemma guarantees the \emph{existence} of a solution.
+
+A \ksat problem satisfies this condition if each clause shares variables with no more than $\frac{2^k}{e}$ other clauses.
+If we generate a random \ksat instance with $n$ variables and $m$ clauses (and random signs), the probability that two clauses share a variable is approximately $\frac{k^2}{n}$, so the expected degree of any node is approximately $\frac{m k^2}{n}$.
+If we make the optimistic assumption that the maximum degree equals the expected degree, then the proof by Moser and Tardos applies if $m$ is less than $\frac{2^k n}{k^2 e}$.
+For comparison, the maximum number of clauses in a \ksat instance is $2^k {n \choose k}$.
+It is known that random 3-SAT instances are hardest when $m \approxeq 4.26 n$ \cite{gomes2008satisfiability}, so these are relatively easy instances.
+In fact, for $k = 3$ these instances do not even use all the variables, since $\frac{2^3 m}{3^2 e} = \frac{8}{9e} < 1/3$.
+For larger $k$, the instances for which the proof by Moser and Tardos applies can be much less trivial than that, but they can never meet the threshold, which is asymptotically $\log 2 2^k - \frac{1}{2}(1 + \log 2)$\cite{coja2014asymptotic}.
+Of course, we should not expect to be able provably to solve hard instances of NP-complete problems in polynomial time!
+We will provide empirical results for problems that are at varying degrees of proximity to the hardness threshold.
+
+\subsection{Parallelizing Moser-Tardos}
+The event dependency graph is useful for seeing a way to parallelize algorithm \ref{alg:mt-meta}.
+Notice that if independent events $E_j$ and $E_k$ happen under the current assignment to $\mathcal{A}$, and event $E_j$'s variables are chosen for resampling, then (by independence) $E_k$ will still happen after the resampling, so there exists a valid serial execution of the algorithm in which the next iteration resamples $E_k$'s variables.
+This means we can resample variables depended-on by any independent set of events that currently happen, and this resampling can be done in parallel.
+Figure \ref{fig:mt-independence-example} gives an example.
+
+\begin{figure}[ht]
+  \includegraphics[scale=0.4]{figures/mt-independence-example.png}
+  \caption{The dependency graph for the problem pictured in figure \ref{fig:mt-directed-example}, with an independent set highlighted.
+If these two clauses are unsatisfied at some stage of the algorithm, the variables of both can be resampled simultaneously.
+If we could only identify a smaller independent set, there would be less opportunity for parallelism.}
+  \label{fig:mt-independence-example}
+\end{figure}
+
+The larger the independent set we can find, the more parallelism is available.
+In fact, Moser and Tardos prove that the algorithm terminates in $O(\log(m))$ iterations -- that is, it has sub-linear computational depth -- if the variables of a \emph{maximal} independent set are resampled at each step.
+(Note that the requirement is not for a \emph{maximum} independent set, finding which is itself an NP-complete problem, but merely an independent set that cannot be greedily extended.)  Since the set of events that happen changes on each iteration, it is necessary to find new independent sets on each iteration, so this is not a fixed cost.
+In practice, this is a major bottleneck of the algorithm.
+Moser and Tardos suggest using Luby's algorithm \cite{luby1986simple}.
+In short, that algorithm grows an independent set by assigning random IDs to each node in the graph, then iteratively adding all remaining strict local minima to the independent set, and removing the added nodes and their neighbors.
+It finds a maximal independent set using $O(\log m)$ rounds of communication and $O(m \log m)$ total work, in expectation and with high probability.
+Chung et al \cite{chung2014distributed} and Srinivasan et al \cite{haeupler2011new} give alternative parallel LLL algorithms that can be run more efficiently on distributed computers.
+These algorithms differ only in the methods they use to find independent sets; the requirement for a maximal independent set is typically relaxed in some fashion.
+
+%FIXME: Mention LOCAL model?
+
+For our parallel code, we have implemented the so-called ``simple'' algorithm of Chung et al (algorithm 2 in \cite{chung2014distributed}) using some techniques suggested by Haeupler et al \cite{haeupler2011new}.
+We call this algorithm \ref{alg:chung-simple}, and pseudocode is given below.
+
+\begin{algorithm}[H]
+\label{alg:chung-simple}
+\begin{algorithmic}
+\State Initialize all the $A_i$ as i.i.d. samples from a uniform Bernoulli distribution.
+\State Mark each of the $E_j$ with a random integer ID.
+\While{Any of the $E_j$ happen under the current value of $\mathcal{A}$}
+  \State Let $U = \{E_j: E_j\text{ currently happens}\}$. 
+  \State Let $S = \{E_j: E_j \in U, ID(E_j) < \min_{k \in \Gamma(j) \cap U} ID(E_k)\}$, the set of strict local minima among unsatisfied events.
+  \State Resample each $A_i \in \cup_{E_j \in S} \operatorname{vbl}(E_j)$ i.i.d. uniform Bernoulli.
+\EndWhile
+\end{algorithmic}
+\caption{The ``simple'' algorithm of Chung et al.  Here $\Gamma(j)$ is the set of neighbors of event $E_j$ in the dependency graph.}
+\end{algorithm}
+
+Algorithm \ref{alg:chung-simple} simply uses Luby's algorithm, but stops after a single iteration rather than continuing until a maximal independent set has been found.
+Further, the random IDs are reused across iterations of algorithm 1.
+We will see that these two modifications substantially reduce the communication requirements of the Moser-Tardos meta-algorithm.
+The pseudocode does not address the parallelism or communication patterns used in the algorithm, which is the subject of section \ref{sec:implementation}.
+It should be noted, however, that algorithm \ref{alg:chung-simple} is proven to run in polynomial time only when $e p d^2 < 1$, a stronger requirement.
+Chung et al provide another algorithm with the same dependency on $d$ as the Moser-Tardos meta-algorithm (algorithm 3 in their paper), but it involves more communication than the simple algorithm.
+We implemented only a serial version of it.
+
+% Last, it should be noted that the simple LLL algorithm is quite similar to stochastic local search heuristics for \ksat that have been known for many years.
+It would be surprising if it performed much better than those older methods, or the more sophisticated heuristics that have followed them.
+So we expect to find negative results.
+The parallel LLL algorithms are less obvious, so there is more hope that they outperform or outscale existing algorithms on instances that satisfy Moser's condition for efficiency.
+
+\section{Our Implementation}
+\label{sec:implementation}
+We have written Julia implementations of algorithms \ref{alg:mt-ksat} and \ref{alg:chung-simple}, and algorithm 3 in \cite{chung2014distributed}.
+Each of these algorithms is a modification of the Moser-Tardos meta-algorithm with a particular way of finding independent sets.
+Since there is potentially a tradeoff between the time spent to find independent sets and the number of iterations the algorithm must perform, these implementations are useful for independently examining the number of iterations taken by each algorithm.
+Otherwise, since it is serial, the Julia code is straightforward.
+
+We have also written a Python implementation of algorithm \ref{alg:chung-simple} using the MPI bindings in the mpi4py library.
+Since this code is targeted at a distributed-memory environment, we had to consider how to parallelize each step of the algorithm.
+We made rather different design choices than suggested by Chung et al.
+
+Chung et al assume the LOCAL model of communication.
+In this model, each clause of the \ksat problem is assigned its own processor and may communicate with its neighbors in the dependency graph.
+The running time of an algorithm in this model is proportional to the number of communication rounds.
+Since a clause's variables are shared only with its neighbors in the dependency graph, variable updates need only be shared among neighbors.
+And Luby's algorithm for independent sets only requires nodes to communicate with their neighbors to discover whether any are currently unsatisfied and have smaller IDs.
+In principle, we could use this as a recipe for our code.
+In practice, however, using one processor per clause is tremendously wasteful of computational resources.
+
+Instead, we assume that responsibility the clauses is divided among a small number of processors.
+We assume that the number of processors is small enough, and the dependency graph dense enough, that at least one edge will cross between the subgraphs of the dependency graph on each processor.
+(This assumption holds true even for the easiest \ksat instances we use in our benchmarks.)  This means that the variable assignments must be broadcast across the clusters, at the cost of one phase of $O(n)$ communication per iteration.
+More importantly, when computing independent sets, each processor may need to check whether every problem clause is satisfied.
+To accomplish this, it would be more expensive to broadcast the satisfiedness of every clause than to simply compute it on each machine.
+So we broadcast all of the the problem clauses to each processor so that each processor can compute satisfiedness of any clause given an assignment to the variables.
+This means that the only coupling among the nodes is the variable assignments; all other computation proceeds independently.
+
+Algorithm \ref{alg:implementation} gives pseudocode for our implementation.
+
+\begin{algorithm}[H]
+\label{alg:implementation}
+\begin{algorithmic}
+\State Broadcast all clauses.
+\State Assign each clause to a processor $p(j)$.
+\State Mark each clause with a random integer ID, and broadcast these marks.
+\State Broadcast an initial random assignment.
+\While{True}
+  \State Let $U_k$ be the set of violated clauses assigned to processor $k$.
+  \State Let $U$ be the set of all violated clauses, constructed lazily on each processor.
+  \State Let $S_k = \{E_j: E_j \in U_k, ID(E_j) < \min_{k \in \Gamma(j) \cap U} ID(E_k)\}$.
+  \State Let $V_k = \cup_{E_j \in S_k} \operatorname{vbl}(E_j)$.
+  \State On each processor $k$, resample each variable in $V_k$.
+  \State Broadcast the new variable assignments.
+  \State Terminate if no new variables have been assigned anywhere.
+\EndWhile
+\end{algorithmic}
+\caption{The ``simple'' algorithm of Chung et al, as we have implemented it.}
+\end{algorithm}
+
+\section{Benchmarking \ksat}
+\label{sec:benchmarks}
+Since \ksat is NP-complete, benchmarking it is potentially difficult.  Particular problems %FIXME
+
+\section{Scale Results}
+\label{sec:scaling}
+
+\section{Discussion}
+\label{sec:discussion}
+Our results leave several open questions:
+\begin{description}
+  \item[Graph creation:] Can we speed up or eliminate graph creation?  This turns out to be a major bottleneck in our implementation, which is unsurprising, since it takes $O(m^2)$ total work.
+  %FIXME
+\end{description}
 

### 2015-05-12 julia/run_solution_parallel_SimpleChungIndependentSet.sh
LLL Condition and Iteration Cost Chart

@@ -1,13 +1,13 @@
 
 
-for k in 8
+for n in 256
 do
-    for m in 1280
+    for m in 8 32 128 512
     do
-	echo "k=$k m=$m"
-        for n in 32
+        for k in 2 3 4 8 16
         do
-        	for run in {1..10}
+        echo "k=$k m=$m n=$n"
+        	for run in 1
         	do
             	./run ./Applications/KsatTests.jl -s "ParallelLllKsatSolver(SimpleChungIndependentSetFinder())" -d "RandomKsatGenerator($k,$n,$m)" -e 77 -l $run
         	done

### 2015-05-12 paper/paper.tex
First cut of paper.

@@
 }
+@article{peleg2000distributed,
+  title={Distributed computing},
+  author={Peleg, David},
+  journal={SIAM Monographs on discrete mathematics and applications},
+  volume={5},
+  year={2000},
+  publisher={Springer}
+}
 @inproceedings{steurer2010fast,
@@
 \subsection{The Moser-Tardos meta-algorithm}
-The simplest LLL algorithm, the Moser-Tardos meta-algorithm, is \emph{extremely} simple.  It is algorithm \ref{alg:mt-meta} below.  Its specialization to \ksat is algorithm \ref{alg:mt-ksat}.
-%FIXME: Fix numbering.
+The simplest LLL algorithm, the Moser-Tardos meta-algorithm, is \emph{extremely} simple.  It is algorithm 1 below.  Its specialization to \ksat is algorithm 2.
 
@@
 \label{alg:mt-meta}
-\begin{algorithmic}
+\begin{algorithmic}[1]
 \State Initialize all the $A_i$ as i.i.d. samples from a uniform Bernoulli distribution.
@@
 \label{alg:mt-ksat}
-\begin{algorithmic}
+\begin{algorithmic}[1]
 \State Initialize all variables $V_i$ as i.i.d. samples from a uniform Bernoulli distribution.
@@
 And let $p$ be the maximum probability that any single event happens under a uniform random assignment to all the variables.
-In \ksat, this is $2^{-k}$, since a disjunctive clause is unsatisfied only if all of its k variables are assigned incorrectly, and the single variable assignments happen independently with probability $1/2$.
-Then Moser and Tardos prove that algorithm \ref{alg:mt-meta} finishes in expected polynomial time if $e p d < 1$, where $e$ is the base of the natural logarithm.
+In \ksat, $p = 2^{-k}$, since a disjunctive clause is unsatisfied only if all of its k variables are assigned incorrectly, and the single variable assignments happen independently with probability $1/2$.
+Then Moser and Tardos prove that algorithm 1 finishes in expected polynomial time if $e p d < 1$, where $e$ is the base of the natural logarithm.
 These are the same conditions under which the ordinary \lovasz Local Lemma guarantees the \emph{existence} of a solution.
@@
 A \ksat problem satisfies this condition if each clause shares variables with no more than $\frac{2^k}{e}$ other clauses.
-If we generate a random \ksat instance with $n$ variables and $m$ clauses (and random signs), the probability that two clauses share a variable is approximately $\frac{k^2}{n}$, so the expected degree of any node is approximately $\frac{m k^2}{n}$.
+For instance, if we generate a random \ksat instance with $n$ variables and $m$ clauses (and random signs), the probability that two clauses share a variable is approximately $\frac{k^2}{n}$, so the expected degree of any node is approximately $\frac{m k^2}{n}$.
 If we make the optimistic assumption that the maximum degree equals the expected degree, then the proof by Moser and Tardos applies if $m$ is less than $\frac{2^k n}{k^2 e}$.
@@
 \subsection{Parallelizing Moser-Tardos}
-The event dependency graph is useful for seeing a way to parallelize algorithm \ref{alg:mt-meta}.
+The event dependency graph is useful for seeing a way to parallelize algorithm 1.
 Notice that if independent events $E_j$ and $E_k$ happen under the current assignment to $\mathcal{A}$, and event $E_j$'s variables are chosen for resampling, then (by independence) $E_k$ will still happen after the resampling, so there exists a valid serial execution of the algorithm in which the next iteration resamples $E_k$'s variables.
@@
 
-%FIXME: Mention LOCAL model?
-
 For our parallel code, we have implemented the so-called ``simple'' algorithm of Chung et al (algorithm 2 in \cite{chung2014distributed}) using some techniques suggested by Haeupler et al \cite{haeupler2011new}.
-We call this algorithm \ref{alg:chung-simple}, and pseudocode is given below.
+We call this algorithm 3, and pseudocode is given below.
 
@@
 
-Algorithm \ref{alg:chung-simple} simply uses Luby's algorithm, but stops after a single iteration rather than continuing until a maximal independent set has been found.
+Algorithm 3 simply uses Luby's algorithm, but stops after a single iteration rather than continuing until a maximal independent set has been found.
 Further, the random IDs are reused across iterations of algorithm 1.
@@
 The pseudocode does not address the parallelism or communication patterns used in the algorithm, which is the subject of section \ref{sec:implementation}.
-It should be noted, however, that algorithm \ref{alg:chung-simple} is proven to run in polynomial time only when $e p d^2 < 1$, a stronger requirement.
+It should be noted, however, that algorithm 3 is proven to run in polynomial time only when $e p d^2 < 1$, a stronger requirement.
 Chung et al provide another algorithm with the same dependency on $d$ as the Moser-Tardos meta-algorithm (algorithm 3 in their paper), but it involves more communication than the simple algorithm.
@@
 \label{sec:implementation}
-We have written Julia implementations of algorithms \ref{alg:mt-ksat} and \ref{alg:chung-simple}, and algorithm 3 in \cite{chung2014distributed}.
+We have written Julia implementations of algorithms 2 and 3, and algorithm 3 in \cite{chung2014distributed}.
 Each of these algorithms is a modification of the Moser-Tardos meta-algorithm with a particular way of finding independent sets.
@@
 
-We have also written a Python implementation of algorithm \ref{alg:chung-simple} using the MPI bindings in the mpi4py library.
+We have also written a Python implementation of algorithm 3 using the MPI bindings in the mpi4py library.
 Since this code is targeted at a distributed-memory environment, we had to consider how to parallelize each step of the algorithm.
@@
 
-Chung et al assume the LOCAL model of communication.
+Chung et al assume the LOCAL model of communication \cite{peleg2000distributed}.
 In this model, each clause of the \ksat problem is assigned its own processor and may communicate with its neighbors in the dependency graph.
@@
 In practice, however, using one processor per clause is tremendously wasteful of computational resources.
+It would be practical only for very small problems that could be more efficiently solved serially.
 
-Instead, we assume that responsibility the clauses is divided among a small number of processors.
-We assume that the number of processors is small enough, and the dependency graph dense enough, that at least one edge will cross between the subgraphs of the dependency graph on each processor.
-(This assumption holds true even for the easiest \ksat instances we use in our benchmarks.)  This means that the variable assignments must be broadcast across the clusters, at the cost of one phase of $O(n)$ communication per iteration.
-More importantly, when computing independent sets, each processor may need to check whether every problem clause is satisfied.
-To accomplish this, it would be more expensive to broadcast the satisfiedness of every clause than to simply compute it on each machine.
-So we broadcast all of the the problem clauses to each processor so that each processor can compute satisfiedness of any clause given an assignment to the variables.
-This means that the only coupling among the nodes is the variable assignments; all other computation proceeds independently.
+Instead, we assume that responsibility for the clauses is divided among a small number of processors $p$.
+Specifically, we assume that the number of processors is small enough, and the dependency graph dense enough, that at least one edge will cross between the subgraphs of the dependency graph on each processor, with high probability.
+(This assumption holds true even for the easiest \ksat instances we use in our benchmarks.)  This leads to several design decisions:
+\begin{enumerate}
+  \item When computing independent sets, each processor may need to check the satisfiedness of every problem clause in order to determine whether its own clauses are local minima in the dependency graph.
+To accomplish this, we could compute the satisfiedness of every clause in parallel and then broadcast them.
+But this involves $O(m)$ communication per iteration, which is more expensive than simply computing it each machine at a cost of $O(m)$ computation.
+So we initially broadcast all of the the problem clauses to each processor, and after initialization, each processor can compute satisfiedness of any clause given an up-to-date assignment to the variables.
+  \item The processor responsible for each clause must identify its neighbors in the dependency graph.
+This involves a scan that takes $O(m^2)$ work (though a more sophisticated algorithm that is harder to parallelize takes $O(m k)$ work).
+However, given that all the clauses have been broadcast to every processor, this step is embarrassingly parallel.
+  \item Our assumption on $p$ implies that each processor has at least one clause that depends on each variable, with high probability.
+Therefore the variable assignments must be broadcast across the clusters on each iteration, at the cost of one phase of $O(n)$ communication.
+  \item With the above two choices, the only coupling among the nodes is the variable assignments.
+All other computation proceeds independently, and the variable broadcast is the only communication required (other than the startup cost of broadcasting the clauses).
+\end{enumerate}
 
-Algorithm \ref{alg:implementation} gives pseudocode for our implementation.
+Algorithm 4 gives pseudocode for our implementation.
 
@@
 \State Assign each clause to a processor $p(j)$.
+\State For each clause $j$, compute the neighbor set $\Gamma(j)$ on processor $p(j)$.
 \State Mark each clause with a random integer ID, and broadcast these marks.
-\State Broadcast an initial random assignment.
+\State Broadcast an initial random assignment of the variables.
 \While{True}
@@
 \label{sec:benchmarks}
-Since \ksat is NP-complete, benchmarking it is potentially difficult.  Particular problems %FIXME
+Ideally we would use real problems to test our implementation.
+Since \ksat is NP-complete, there are surely instances that will take exponential time, and there is no complete theoretical characterization of these instances to guide our selection of synthetic instances.
+Unfortunately, due to time constraints we are forced to leave real problems for future work.
 
-\section{Scale Results}
+Instead, we focus on randomly-generated synthetic instances.
+A randomly-generated \ksat instance with parameters $m$, $n$, and $k$ has $m$ clauses, each of which selects one of the ${n \choose k}$ subsets of $k$ variables uniformly at random.
+Each variable in each clause is negated with probability $1/2$.
+It is important to be careful about the selection of problem parameters, since many choices lead to uninteresting problems.
+If $m$ is too small, most variable assignments will solve most instances, while if $m$ is too large, most instances will be unsatisfiable.
+The annual SAT competition \cite{belov2014sat} provides some guidance on the choice of $m$, $n$, and $k$, and our parameter choices are similar to those in the 2009 competition.
+However, it is still unclear how to characterize the difficulty of a particular random instance.
+Even with parameters that typically generate hard problems, some instances will be easy or impossible.
+To generate our benchmarks, we use rejection sampling to find instances for which no more than a small fraction of assignments solve the instance, taking this as our proxy for problem difficulty.
+
+\section{Results}
 \label{sec:scaling}
+We perform all of our experiments on the Hopper cluster.
+
+\subsection{Strong scaling and profiling}
+Strong scaling for one instance is displayed in figure \ref{fig:strong}.
+The algorithm scales almost perfectly for this large problem.
+Profiling our code, we found to our surprise that the initial neighbor identification stage was the primary bottleneck.
+This involves iterating over each pair of nodes and is embarrassingly parallel, so it takes $O(m^2/p)$ time, where $p$ is the number of processors used.
+This explains the favorable scaling.
+Profiling results are displayed in figures \ref{fig:profiling} and \ref{fig:profiling-strong}.
+
+\begin{figure}[ht]
+  \includegraphics[scale=0.4]{figures/strong.png}
+  \caption{Strong scaling for a large problem.}
+  \label{fig:strong}
+\end{figure}
+
+\begin{figure}[ht]
+  \includegraphics[scale=0.4]{figures/profiling.png}
+  \caption{Profiling results for varying problem sizes, in log scale.}
+  \label{fig:profiling}
+\end{figure}
+
+\begin{figure}[ht]
+  \includegraphics[scale=0.4]{figures/profiling-strong.png}
+  \caption{Profiling results for a large problem using 1 and 2 Hopper nodes.}
+  \label{fig:profiling-strong}
+\end{figure}
+
+\subsection{Weak scaling}
+For the reasons described in section \ref{sec:benchmarks}, weak scaling is difficult to establish, since the difficulty of a problem is not simple to define.
+We attempted to use as a proxy for problem difficulty the probability that a random assignment to the variables would solve the problem.
+This proxy was reasonable for small problems, but unreasonable for large ones.
+Figure \ref{fig:weak} displays the results of this test.
+It appears from this figure that the algorithm scales worse than linearly even as $p$ increases linearly with $m$, but this is likely an artifact of our poor proxy for problem difficulty.
+Strong scaling seems to be a more meaningful measure of the utility of parallelism for this problem.
+
+\begin{figure}[ht]
+  \includegraphics[scale=0.4]{figures/weak.png}
+  \caption{An attempted experiment to demonstrate weak scaling.  Large problems are much harder than our proxy suggests.}
+  \label{fig:weak}
+\end{figure}
+
+\subsection{The LLL condition}
+All theoretical results for algorithm 4 require the condition $e p d^2 < 1$, as discussed above.
+Theoretical results for other LLL algorithms require $e p d < 1$.
+However, in our benchmarks we observe no such hard requirement, nor a threshold effect at $e p d \approxeq 1$.
+Figure \ref{fig:lll-condition} compares the number of iterations taken to solve an instance with the value of $e p d$.
+There is no noticeable change at $e p d = 1$.
+This suggests that there may be room for improvement in the theoretical results.
+
+\begin{figure}[ht]
+  \includegraphics[scale=0.4]{figures/lll-condition.png}
+  \caption{The LLL condition value of a problem ($e p d$) versus the number of iterations taken by algorithm 3.
+The problems become harder as $e p d$ increases, but there is no sharp threshold at $1$.}
+  \label{fig:lll-condition}
+\end{figure}
 
@@
 \label{sec:discussion}
-Our results leave several open questions:
+Overall, our results are positive.
+Algorithm 4 performs quite well on the range of problems we have tested, and it exhibits strong scaling.
+By naive random search, we could find no cases in which a problem has a satisfying solution that could not be found in fewer than a hundred iterations of algorithm 4.
+The current bottleneck, neighbor identification, is embarrassingly parallel.
+
+In addition to the questions enumerated in section \ref{sec:intro}, our results leave several open questions for future work:
 \begin{description}
-  \item[Graph creation:] Can we speed up or eliminate graph creation?  This turns out to be a major bottleneck in our implementation, which is unsurprising, since it takes $O(m^2)$ total work.
-  %FIXME
+  \item[Neighbor identification:] Can we speed up neighbor identification?
+An initial random assignment of all variables will satisfy, on average, all but $m 2^{-k}$ clauses, and many clauses may never be unsatisfied during a run of the algorithm.
+If we computed neighbor sets lazily, we could avoid doing any work for these always-satisfied clauses, potentially reducing the cost of neighbor identification by a factor of $2^k$.
+  \item[GPUs:] Our implementation targets a distributed-memory environment, but GPUs could be better choices.
+In a GPU, an implementation closer to the LOCAL computational model, with $p \approxeq m$, might be appropriate.
 \end{description}

### 2015-05-19 paper/cs267-report/paper.tex
Add another paper and a poster.

--- 2015-05-12 paper/paper.tex
@@
 \author{Derek Kuo, Henry Milner}
-\title{CS267 Final Project: An Empirical Investigation of the Algorithmic Lovász Local Lemma}
+\title{CS267 Final Project: An Empirical Investigation of the Algorithmic Lov\'{a}sz Local Lemma}
 \date{5/11/15}
@@
 In 1975, Erd\"{o}s and \lovasz \cite{erdos1975problems} developed a technique, now called the \lovasz Local Lemma (LLL), for proving the existence of solutions to combinatorial problems.
-The technique is a ``probabilistic method'': Given a set of events that we want to avoid (for example, each event could correspond to the violations of a problem constraint), it is proved that there is a strictly positive probability that none of these events happen under a random potential solution drawn from a suitable distribution on potential solutions.
+The technique is a ``probabilistic method'': Given a set of events that we want to avoid (for example, each event could correspond to the violation of one problem constraint), it is proved that there is a strictly positive probability that none of these events if we draw randomly from a suitable distribution on potential solutions.
 The main precondition of the LLL is that the statistical dependencies among events are \emph{sparse}, with the sparsity being of the same magnitude as the marginal probability of each single event being unsatisfied under the sampling distribution.
@@
 
-To study applications of the algorithmic LLL, we must choose combinatorial problems to solve.
+To study applications of the algorithmic LLL, we must choose a combinatorial problem to solve.
 We focus on \ksat.
@@
 In fact, there is an annual competition for SAT solvers \cite{belov2014sat}.
-(Our benchmarks will be based on the guidelines in that competition.)  We do not expect that our implementations will be competitive with state-of-the-art SAT solvers.
+(Our benchmarks will be based on the guidelines in that competition.)  We do not expect that our implementations will be competitive with state-of-the-art SAT solvers, and such comparisons are outside the scope of this project.
 Our interest is primarily in finding out whether there is any practical potential for an algorithm whose development has thus far been guided exclusively by the desire to prove results about its performance.
@@
 \label{sec:alll}
-LLL algorithms can be useful when a problem can be phrased as follows.
+LLL algorithms can be useful for problems that can be phrased as follows.
 Let $\mathcal{A} = \{A_1, \cdots, A_n\} \in \{0,1\}^n$ be a set of discrete variables (which we take to be binary for sake of exposition), and let $\mathcal{E} = \{E_1, \cdots, E_m\}$ be a set of functions $\{0,1\}^n \to \{0,1\}$ mapping the variables to truth values.
@@
 If we imagine randomly and independently determining the variable values, then $E_i$ is statistically dependent only on $\operatorname{vbl}(E_i)$.
-LLL algorithms typically need $|\operatorname{vbl}(E_i)| << |\mathcal{A}|$.
+LLL algorithms typically need $|\operatorname{vbl}(E_i)| \ll |\mathcal{A}|$.
 
@@
 In fact, for $k = 3$ these instances do not even use all the variables, since $\frac{2^3 m}{3^2 e} = \frac{8}{9e} < 1/3$.
-For larger $k$, the instances for which the proof by Moser and Tardos applies can be much less trivial than that, but they can never meet the threshold, which is asymptotically $\log 2 2^k - \frac{1}{2}(1 + \log 2)$\cite{coja2014asymptotic}.
+For larger $k$, the instances for which the proof by Moser and Tardos applies can be much less trivial than that, but they can never meet the threshold, which is asymptotically $(\log 2) 2^k - \frac{1}{2}(1 + \log 2)$\cite{coja2014asymptotic}.
 Of course, we should not expect to be able provably to solve hard instances of NP-complete problems in polynomial time!

### 2015-05-19 paper/cs270-report/paper.tex
Add another paper and a poster.

\documentclass[twocolumn]{article}
\usepackage{amsmath, amssymb}
\usepackage[retainorgcmds]{IEEEtrantools}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{filecontents}
\usepackage{hyperref}
\author{Henry Milner}
\title{Toward A Non-Atomic Algorithmic Lov\'{a}sz~ Local Lemma}
\date{5/15/15}

% Some macros specific to this paper.
\newcommand{\ksat}{\texttt{k-SAT}~}

\newcommand{\lovasz}{Lov\'{a}sz~}

% The name of the generic algorithm.
\newcommand{\mt}{MT~}

% Some convenience functions for homework problems.
\newcommand{\problem}[1]%
  {\section*{#1.}}

\newcommand{\problemSubpart}[1]%
  {\noindent\emph{#1.}}

\newcommand{\problemNamedSubpart}[1]%
  {\noindent\emph{#1}}

% Some convenience functions for note-taking.
\newcommand{\topic}[1]%
  {\section*{#1}}

% Some functions for general use.

\def\seqn#1\eeqn{\begin{align}#1\end{align}}

\newcommand{\vecName}[1]%
  {\boldsymbol{#1}}

\newcommand{\io}%
  {\text{ i.o. }}

\newcommand{\eventually}%
  {\text{ eventually }}

\newcommand{\tr}%
  {\text{tr}}

\newcommand{\Cov}%
  {\text{Cov}}

\newcommand{\adj}%
  {\text{adj}}

\newcommand{\funcName}[1]%
  {\text{#1}}

\newcommand{\hasDist}%
  {\sim}

\DeclareMathOperator*{\E}%
  {\mathbb{E}}

\newcommand{\Var}%
  {\text{Var}}

\newcommand{\std}%
  {\text{std}}

\newcommand{\grad}%
  {\nabla}

\DeclareMathOperator*{\argmin}{arg\,min}

\DeclareMathOperator*{\argmax}{arg\,max}

\newcommand{\inprod}[2]%
  {\langle #1, #2 \rangle}

\newcommand{\dd}[1]%
  {\frac{\delta}{\delta#1}}

\newcommand{\Reals}%
  {\mathbb{R}}

\newcommand{\indep}%
  {\protect\mathpalette{\protect\independenT}{\perp}} \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}

\newcommand{\defeq}%
  {\buildrel\triangle\over =}

\newcommand{\defn}[1]%
  {\emph{Definition: #1}\\}

\newcommand{\example}[1]%
  {\emph{Example: #1}\\}

\newcommand{\figref}[1]%
  {\figurename~\ref{#1}}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newenvironment{proof}[1][Proof]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\begin{filecontents}{\jobname.bib}
@article{erdos1975problems,
  title={Problems and results on 3-chromatic hypergraphs and some related questions},
  author={Erdos, Paul and Lov{\'a}sz, L{\'a}szl{\'o}},
  journal={Infinite and finite sets},
  volume={10},
  number={2},
  pages={609--627},
  year={1975}
}
@article{beck1991algorithmic,
  title={An algorithmic approach to the Lov{\'a}sz local lemma. I},
  author={Beck, J{\'o}zsef},
  journal={Random Structures \& Algorithms},
  volume={2},
  number={4},
  pages={343--365},
  year={1991},
  publisher={Wiley Online Library}
}
@article{moser2010constructive,
 author = {Moser, Robin A. and Tardos, G\'{a}bor},
 title = {A Constructive Proof of the General Lov\ÁSz Local Lemma},
 journal = {J. ACM},
 issue_date = {January 2010},
 volume = {57},
 number = {2},
 month = feb,
 year = {2010},
 issn = {0004-5411},
 pages = {11:1--11:15},
 articleno = {11},
 numpages = {15},
 url = {http://doi.acm.org/10.1145/1667053.1667060},
 doi = {10.1145/1667053.1667060},
 acmid = {1667060},
 publisher = {ACM},
 address = {New York, NY, USA},
 keywords = {Constructive proof, Lov\'{a}sz local lemma, parallelization},
}
@inproceedings{moser2009constructive,
  title={A constructive proof of the Lov{\'a}sz local lemma},
  author={Moser, Robin A},
  booktitle={Proceedings of the forty-first annual ACM symposium on Theory of computing},
  pages={343--350},
  year={2009},
  organization={ACM}
}
@misc{tao2009entropy,
  author = {Tao, Terence},
  title = {Moser's entropy compression argument},
  journal = {Blog},
  type = {Blog},
  number = {August 5},
  year = {2009},
  howpublished = {\url{https://terrytao.wordpress.com/2009/08/05/mosers-entropy-compression-argument/}}
}
@misc{fortnow2009entropy,
  author = {Fortnow, Lance},
  title = {A Kolmogorov Complexity Proof of the Lov\'{a}sz Local Lemma},
  journal = {Blog},
  type = {Blog},
  number = {June 02},
  year = {2009},
  howpublished = {\url{http://blog.computationalcomplexity.org/2009/06/kolmogorov-complexity-proof-of-lov.html}}
}
@article{haeupler2011new,
  title={New constructive aspects of the lovasz local lemma},
  author={Haeupler, Bernhard and Saha, Barna and Srinivasan, Aravind},
  journal={Journal of the ACM (JACM)},
  volume={58},
  number={6},
  pages={28},
  year={2011},
  publisher={ACM}
}
@inproceedings{freer2010probabilistic,
  title={When are probabilistic programs probably computationally tractable?},
  author={Freer, Cameron E and Mansinghka, Vikash K and Roy, Daniel M},
  booktitle={NIPS Workshop on Advanced Monte Carlo Methods with Applications},
  year={2010}
}
@article{wainwright2008graphical,
  title={Graphical models, exponential families, and variational inference},
  author={Wainwright, Martin J and Jordan, Michael I},
  journal={Foundations and Trends{\textregistered} in Machine Learning},
  volume={1},
  number={1-2},
  pages={1--305},
  year={2008},
  publisher={Now Publishers Inc.}
}
@inproceedings{achlioptas2014random,
  title={Random Walks that Find Perfect Objects and the Lov{\'a}sz Local Lemma},
  author={Achlioptas, Dimitris and Iliopoulos, Fotis},
  booktitle={Foundations of Computer Science (FOCS), 2014 IEEE 55th Annual Symposium on},
  pages={494--503},
  year={2014},
  organization={IEEE}
}
@inproceedings{papadimitriou1991selecting,
  title={On selecting a satisfying truth assignment},
  author={Papadimitriou, Christos H},
  booktitle={Foundations of Computer Science, 1991. Proceedings., 32nd Annual Symposium on},
  pages={163--169},
  year={1991},
  organization={IEEE}
}
@inproceedings{schoning1999probabilistic,
  title={A probabilistic algorithm for k-SAT and constraint satisfaction problems},
  author={Schoning, T},
  booktitle={Foundations of Computer Science, 1999. 40th Annual Symposium on},
  pages={410--414},
  year={1999},
  organization={IEEE}
}
@article{alon1991parallel,
  title={A parallel algorithmic version of the local lemma},
  author={Alon, Noga},
  journal={Random Structures \& Algorithms},
  volume={2},
  number={4},
  pages={367--378},
  year={1991},
  publisher={Wiley Online Library}
}
@inproceedings{chung2014distributed,
  title={Distributed algorithms for the Lov{\'a}sz local lemma and graph coloring},
  author={Chung, Kai-Min and Pettie, Seth and Su, Hsin-Hao},
  booktitle={Proceedings of the 2014 ACM symposium on Principles of distributed computing},
  pages={134--143},
  year={2014},
  organization={ACM}
}
@inproceedings{gonzalez2011parallel,
  title={Parallel gibbs sampling: From colored fields to thin junction trees},
  author={Gonzalez, Joseph and Low, Yucheng and Gretton, Arthur and Guestrin, Carlos},
  booktitle={International Conference on Artificial Intelligence and Statistics},
  pages={324--332},
  year={2011}
}
@book{alon2004probabilistic,
  title={The probabilistic method},
  author={Alon, Noga and Spencer, Joel H},
  year={2004},
  publisher={John Wiley \& Sons}
}
@inproceedings{steurer2010fast,
  title={Fast SDP algorithms for constraint satisfaction problems},
  author={Steurer, David},
  booktitle={Proceedings of the twenty-first annual ACM-SIAM symposium on Discrete Algorithms},
  pages={684--697},
  year={2010},
  organization={Society for Industrial and Applied Mathematics}
}
@inproceedings{polik2007sedumi,
  title={SeDuMi: a package for conic optimization},
  author={Polik, Imre and Terlaky, Tamas and Zinchenko, Yuriy},
  booktitle={IMA workshop on Optimization and Control, Univ. Minnesota, Minneapolis},
  year={2007}
}
@article{shannon2001mathematical,
  title={A mathematical theory of communication},
  author={Shannon, Claude Elwood},
  journal={ACM SIGMOBILE Mobile Computing and Communications Review},
  volume={5},
  number={1},
  pages={3--55},
  year={2001},
  publisher={ACM}
}
\end{filecontents}
\immediate\write18{bibtex \jobname}

\begin{document}
\maketitle

\abstract{The \lovasz Local Lemma (LLL) \cite{erdos1975problems} is a probabilistic method for proving the existence of elements of a $\sigma$-algebra $(\Omega, \Sigma)$ that miss all of a set of events $\mathcal{A}$.  The method works when a probability measure can be assigned to $\Sigma$ so that the marginal probability of the events is low and the their statistical dependency graph is sparse.  In the less abstract \emph{variable version} of the LLL problem, the space is a product space of, say, $n$ independent variables and each event corresponds to a binary function of some of the variables.  In 2010, Moser and Tardos \cite{moser2009constructive,moser2010constructive} made the LLL method constructive for variable LLL problems, giving a surprisingly simple expected-polynomial-time algorithm to find variable assignments that make no events happen.  They used a novel entropy compression argument, proving that their algorithm could be used to compress random bits if it took more than a certain amount of expected time to find a solution.  Since then, there has been a flurry of follow-up work.  In this paper, we briefly review the Moser-Tardos result and a recent generalization by Achlioptas and Iliopoulos \cite{achlioptas2014random}).  Then we propose a generalization of the entropy compression argument to new ``non-atomic'' settings using \emph{relative entropy compression}.}

\section{Introduction}
The \lovasz Local Lemma (LLL) is a probabilistic method for proving the existence of elements of a probability space $\mathcal{P} = (\Omega, \Sigma, \mu)$ that miss all of a set of events $\mathcal{A} = \{A_1, \cdots, A_m\}$, the search for which we call an ``LLL problem''.  It was introduced by \lovasz and Erd\"{o}s in 1975 \cite{erdos1975problems}; the proof is a fairly straightforward application of the probabilistic method \cite{alon2004probabilistic}.  The LLL has many applications in combinatorial search problems.  A useful restricted version of the problem is the ``variable version,'' in which $\mathcal{P}$ is a product space of $n$ independent random variables $X = (X_1, \cdots, X_n)$.  (We will sometimes write $A_i(X)$ for the indicator that event $i$ happens when the variables take value $X$.)  For example, we can use the LLL to prove the existence of solutions to certain \ksat problems by imagining a probability space on $n$ random variables in which each event $A_j$ occurs if a corresponding disjunctive clause $C_j \in \mathcal{C} = \{C_1, \cdots, C_m\}$ of $k$ variables is false.  In that setup, ``missing all events'' corresponds to satisfying all clauses.  Not every \ksat problem has a satisfying solution, so we should expect the conditions of the LLL to be restrictive.  We will come back to the \ksat example several times, since it is a very straightforward application.

A seminal paper by Beck \cite{beck1991algorithmic} showed that the LLL could be made \emph{algorithmic} -- that is, when the LLL conditions hold, certain algorithms will \emph{find} a solution quickly.  This started a literature around the ``constructive'' or ``algorithmic'' LLL.  Beck's result required stronger conditions than the non-constructive LLL and worked only for particular problems (like \ksat).  In 2010, Moser and Tardos \cite{moser2009constructive,moser2010constructive} published a breakthrough meta-algorithm for finding solutions to any variable LLL problem under the same conditions as the non-constructive LLL.  In its simplest form, the MT algorithm is strikingly simple:

\begin{algorithm}[H]
\caption{The Moser-Tardos (MT) algorithm, in simple iterative form.}
\label{alg:mt}
\begin{algorithmic}
\State Initialize the $X_i$ IID uniform.
\While{Any of the $A_j$ happen under the current value of $X$}
  \State Let $A_j$ be an arbitrary event that currently happens.
  \State Resample each $X_i \in \operatorname{vbl}(A_j)$ IID uniform.
\EndWhile
\end{algorithmic}
\end{algorithm}

For example, specialized to \ksat, the MT algorithm becomes:

\begin{algorithm}[H]
\caption{The Moser-Tardos algorithm, specialized to \ksat.}
\label{alg:mt-ksat}
\begin{algorithmic}[1]
\State Initialize the $X_i$ IID uniform Bernoulli.
\While{Any clause is violated by $X$}
  \State Let $C_j$ be an arbitrary clause that is violated.
  \State Resample each $X_i \in \operatorname{vbl}(C_j)$ IID uniform Bernoulli.
\EndWhile
\end{algorithmic}
\end{algorithm}

Moser and Tardos prove that their algorithm finishes in expected polynomial time under the \emph{standard LLL condition} $e p (d+1) < 1$, where
\begin{description}
  \item[$e$] is the base of the natural logarithm;
  \item[$p$] is the maximum marginal probability of any event when $X \hasDist \operatorname{Unif}(\Omega)$; and
  \item[$d$] is the maximum degree of any event in the \emph{dependency graph} on events, in which two events are connected by an edge if they are statistically dependent when $X \hasDist \operatorname{Unif}(\Omega)$ .
\end{description}

After the Moser and Tardos paper, interest in the subject has only increased; follow-up work has given new algorithms, new problem domains, new applications, and a range of sufficient conditions weaker than the LLL condition.

\subsection{This paper}
In this paper, we first review the result by Moser and Tardos, using a more illuminating proof by Tao \cite{tao2009entropy} and Fortnow \cite{fortnow2009entropy} that directly highlights the role of entropy compression.  In section \ref{sec:perfect}, we review the extension of the algorithmic LLL to non-variable problems by Achlioptas and Iliopoulos \cite{achlioptas2014random}.  Using these two ideas, in section \ref{sec:nonatomic} we sketch an argument that suggests the entropy compression argument could be extended to novel LLL-like applications in which the state vector of an algorithm cannot be uniquely recovered.  Achlioptas and Iliopoulos call such algorithms (or rather, technically, their associated random walk graphs) ``non-atomic''.  The original work in this paper is preliminary and incomplete, and we look forward to improving it in the near future.

\section{The algorithmic LLL}
\label{sec:lll}
Though the iterative version of the MT algorithm is very simple, a slightly more complicated \emph{recursive} one is useful for analysis:

\begin{algorithm}[H]
\caption{A more complicated recursive version of the Moser-Tardos (MT) algorithm.  Here $\Gamma(i)$ denotes the neighbors of event $i$ in the event dependency graph.}
\label{alg:mt-recursive}
\begin{algorithmic}
\State Initialize the $X_i$ IID uniform.
\ForAll{$i \in [m]$}
  \Call{Fix}{i}
\EndFor

\Function{Fix}{i}
  \If{$! A_i(X)$}
    \Return
  \EndIf
  \State Resample each $X_i \in \operatorname{vbl}(A_i)$ IID uniform.
  \ForAll{$j \in [m]: j \in \Gamma(i), A_j(X)$}
    \Call{Fix}{j}
  \EndFor
\EndFunction

\end{algorithmic}
\end{algorithm}

We would like to prove that this algorithm terminates in polynomial time when the LLL condition is met.  We will briefly review an argument by Tao \cite{tao2009entropy} and Fortnow \cite{fortnow2009entropy} that emphasizes the entropy compression aspect of the proof.  The argument is specialized to \ksat to make the exposition simpler.

The general idea is to imagine running algorithm \ref{alg:mt-recursive} until \texttt{Fix} is called $t$ times.  If $i(s)$ denotes the argument to \texttt{Fix} on iteration $s$, then after $t$ calls, $\sum_{j = 1}^{t} \operatorname{vbl}(A_{t(j)}) = k t$ binary variables have been resampled independently using some random source, and the initial assignment used $n$ bits.  Together, the algorithm state vector, which we call $Z$, encodes $n + k t$ purely random bits.  We will use the assumption $2^C_0 p d < 1$ to give a method to encode the algorithm state in an alternative way $Y$, which we call the ``alternative information''.  (We will use $C_0 = 3$, though the proof is possible under the weaker condition $e p (d + 1) < 1$.)  When $t$ is large, $Y$ will contain fewer bits than $Z$, which contradicts Shannon's source coding theorem \cite{shannon2001mathematical} (it is impossible to encode purely random bits with a scheme that uses fewer bits in expectation).

We note first that, given an innocuous assumption about the problem (which is true for \ksat and can be made true for other problems), there is a 1-to-1 correspondence between the pair $((A_{t(i)})_{i=1}^{t}, \sigma_{t+1})$ and $Z$.  Here $\sigma_t$ denotes the state of $X$ before call $t$.  The assumption we need is that each event $A_i$ happens for only one setting of $\operatorname{vbl}(A_i)$.  For example, in \ksat, a disjunctive clause can only be violated by one setting of its variables.  In section \ref{subsec:trivial} we will explain why this assumption is innocuous in general.  Under the assumption, there is an easy inductive argument: If $\sigma_{i+1}$ and $A_{t(i)}$ are known, then $\sigma_i$ must equal $\sigma_{i+1}$ outside $\operatorname{vbl}(A_{t(i)})$, and on $\operatorname{vbl}(A_{t(i)})$ it must equal the one violating setting, since it was chosen to be \texttt{Fix}ed.  But $\sigma_{t+1}$ and all the events are known, so the whole state vector $Z$ (including the initial state) is determined.

It remains to encode $(A_{t(i)})_{i=1}^{t}$ efficiently given $\sigma_{t+1}$.  Note that we cannot simply encode the events using $\log m$ bits, since this would not give us the bound we want.  Instead we will use a structure called (by Moser and Tardos) a ``witness forest.''  In this forest, each event called directly by the outer loop of the algorithm (and which actually causes a resample) is a root, and an event called inside $\texttt{Fix}(i)$ is a child of event $A_i$.  For \ksat, $p = 2^{-k}$, so by the assumption $2^C_0 p d < 1$, each event is independent of all but at most $2^{k-C_0}$ other events.  So if an event is a child, it can be encoded using at most $2^{k-C_0}$ bits.  Each event is either a root or not, so the roots can be encoded using $m$ bits.  Finally, to encode each tree, we can use an extra bit per element to denote the end of the tree.

So we can encode $Y$ using at most $n + 2m + (k - C_0 + 1)t$ bits, and we can determine $X$ from $Y$, so by Shannon's source coding theorem
\seqn
  n + m + (k - C_0 + 1)t \leq n + k t .
\eeqn

This is violated if $t \geq \frac{m}{C_0 - 1}$, so the algorithm must finish (in expectation) before then.  (In particular, it must finish \emph{eventually}, so there must exist a satisfying solution, and this argument is another proof of the LLL.)

\section{The algorithmic LLL on general sample spaces}
\label{sec:perfect}
Thus far, we have seen only a variable-version LLL problem in which our probability space was a product of $n$ variables.  The MT algorithm requires explicit access to those variables, and it is restricted to handling every event in the same way -- by resampling all of its variables.  Both assumptions are fairly restrictive.  The sample space for some combinatorial problems is better represented in other forms, like permutations \cite{achlioptas2014random}.  And contra the second assumption, many local search algorithms for \ksat (e.g. \texttt{WalkSAT} \cite{papadimitriou1991selecting}) are similar to MT, but will flip only a smaller number of variables on each iteration.  Achlioptas and Iliopoulos \cite{achlioptas2014random} provide an elegant framework that allows the probability space to be arbitrary and the algorithm to choose actions more flexibly, for which variable-version problems and algorithms are a special case.  Examining this generalization is instructive.

Again take our $\sigma$-algebra to be $(\Omega, \Sigma)$, but this time without a product structure and without a measure $\mu$.  Our set of events, which we will now call ``flaws'', are the same as before, but now there is no notion of $\operatorname{vbl}(A_i)$.  Let $U(\sigma)$ denote the set of flaws that happen under state $\sigma$.  Again, our goal is to find any element $\sigma^* \in \Omega$ such that none of the events happen; we call such elements with $U(\sigma^*) = \emptyset$ ``flawless objects.''  Consider any randomized iterative algorithm whose random state is Markovian and stationary, which attempts to find a flawless object.  (The MT algorithm is an example of such an algorithm.)  We can always encode such an algorithm in a random walk graph $\mathcal{T}$ on $\Omega$.  Now let us try to write down the properties of this graph that are really needed for something like the proof in section \ref{sec:lll} to apply.

The first requirement is ``accountability of action.''  In Tao's proof, we needed the fact that the flaw (event) $A_t$ addressed on iteration $t$ uniquely determined the values of $\operatorname{vbl}(A_t)$ in the previous state $\sigma_t$.  We proved this by assuming, by induction, that we knew the values of the variables in $\sigma_{t+1}$.  Given the identity of the flaw, we knew that $\sigma_t$ and $\sigma_{t+1}$ differed only on the variables in $\operatorname{vbl}(A_t)$ differed.  Further, since each flaw could be violated by only one configuration of its variables, we knew what that configuration must have been in $\sigma_t$.  Without variables, we need a different tack to get a correspondence between flaw vectors and algorithm state vectors.  We will still assume that the algorithm ``addresses'' a single flaw on each iteration; given a current state $\sigma_t$ and a flaw to address, the algorithm draws from some distribution to find its next action.  To achieve accountability, we assume that, for every (flaw, state) pair $(A_{t}, \sigma_{t+1})$, there is a unique previous state $\sigma_t$.  In terms of the random walk graph, this means that if we label each weighted edge in the graph with the flaw we addressed at that step, each node has at most one incoming edge for each flaw.  This requirement is called ``atomicity.''

In the variable version of the problem, there are two simple sufficient conditions for atomicity:
\begin{description}
  \item[Simple constraints:] For each flaw $A_i$, there is only one configuration of the variables $\operatorname{vbl}(A_i)$ that makes the flaw happen.  This condition is unimportant, since we can always modify the algorithm (for purposes of proofs) to make it so.  We will discuss this further in section \ref{sec:nonatomic}.
  \item[Algorithmic focusing:] After choosing a flaw $A_t$, the algorithm modifies \emph{only} the variables in $\operatorname{vbl}(A_t)$ on iteration $t$.  This condition is nontrivial.
\end{description}

The second requirement is analogous to the standard LLL condition $e p (d+1)$ (though other, more elaborate results are proven in \cite{achlioptas2014random}).  In section \ref{sec:lll}, we saw that it was desirable (from the perspective of the entropy compression argument) to maximize the entropy of each step.  So we will assume that, having chosen a flaw $A_t$ to address, our algorithm chooses uniformly at random from all the possible next states it could reach.  We call this ``action set'' $A(A_t, \sigma_t)$.  (In the variable version, this would have been of size $2^{\operatorname{vbl}(A_t)}$ for any $\sigma_t$.)  Then, the requirement is that no flaw $A_i$ has too many neighbors with small action sets; specifically,
\seqn
  \label{eqn:degree-req}
  e \sum_{j \in \Gamma(i)} \frac{1}{\min_{\sigma: A_j \in U(\sigma)} |A(\sigma, A_j)|} < 1 - \delta
\eeqn
for some positive $\delta$.

Under these two requirements, Achlioptas and Iliopoulos show that any such algorithm will terminate quickly.  In that case, the running time of the random walk algorithm is $\delta^{-1} (\log_2 |\Omega| + |U(\sigma_1)| + s)$ steps with probability at least $1 - 2^{-s}$.

Interestingly, instead of using a measure $\mu$ on $(\Omega, \Sigma)$, requirement \ref{eqn:degree-req} is entirely in terms of the random walk graph, and it is independent of the starting location $\sigma_1$.  The proof does not require any entropy in the starting state.  Still, the main point of our interest in \cite{achlioptas2014random} is the clear statement of the atomicity requirement.

\section{Non-atomic algorithms}
\label{sec:nonatomic}
In sections \ref{sec:lll} and \ref{sec:perfect}, we saw that the \emph{atomicity} of an \mt algorithm was a key assumption in the entropy-compression argument.  In this section, we conjecture that similar arguments could be made without the atomicity assumption.  It would be very interesting if this conjecture held, since it could open up the algorithmic LLL to applications outside combinatorial search.  We look forward to following up on this conjecture in future work, but for now we will just sketch some arguments.  The work in this section is, to our knowledge, original.

Recall that assumptions on the degree of nodes in the transformation graph of an \mt algorithm allowed us to show that the identity of the flaw addressed at each step could be encoded using fewer bits on average than would be needed to encode the randomness involved in making the step itself.  Assuming atomicity (recall its two requirements: (1) simple constraints and (2) algorithmic focusing) then gave us a way to exactly reconstruct the state of the algorithm before the step was taken.  Now, let us abstractly write $Z$ for a vector of algorithm states (in the MT algorithm for \ksat, the concatenation of the original uniform assignment bits and the vectors of $k$ bits used at each iteration); $Y$ for the hypothetical observed information that encodes information sufficient to reconstruct the algorithm states (in our explication of Tao's MT proof, this includes the witness forest, the set of initially-violated events, and the final assignment); and $\operatorname{Infer}$ for the deterministic function that maps $Y$ to $Z$.  Then the purpose of the atomicity assumption is that $P(Z | Y)$ is a point mass measure at $\operatorname{Infer}(Y)$.

Existing work, including \cite{achlioptas2014random}, assumes that it is necessary for $P(Z | Y)$ to be a point mass, but what would it mean if it were not?  We could recast the original entropy compression argument as a \emph{relative entropy compression} argument.

Consider the state vector for $t$ steps of the \mt algorithm for \ksat, which is of size $M = m + kt$.  Then the ``prior'' on $Z$, the distribution before we observe anything, is $P(Z) = \operatorname{Unif}(2^{m+kt})$.  By Bayes' rule, $H(Y | Z) = H(Z | Y) - H(Z) + H(Y)$.  If we make the reasonable assumption that $Y$ can be recovered from the actual state vector $Z$, then $H(Y | Z) = 0$, and we have
\seqn
  \label{eqn:entropy-general}
  H(Z) = H(Z | Y) + H(Y) .
\eeqn
Typically we will proceed by finding a simple expression for the left hand side (which is typically determined explicitly by the algorithm) and upper-bounding the right hand side.  For example, if we can encode the alternative information $Y$ using $N$ bits, an upper bound on the prior entropy $H(Y)$ is $N$, and then equation \ref{eqn:entropy-general} becomes
\seqn
  \label{eqn:entropy-encoded}
  H(Z) \leq H(Z | Y) + N .
\eeqn
When $P(Z | Y)$ is a point mass, $H(Z | Y) = 0$, and inequality \ref{eqn:entropy-encoded} reduces to
\seqn
  \label{eqn:entropy-tao}
  H(Z) \leq N .
\eeqn
This is exactly the fact that we needed in Tao's proof to show a contradiction when $t$ was too large.

From this perspective, it does not seem strictly necessary that $H(Z | Y) = 0$.  Perhaps in some cases we can use inequality \ref{eqn:entropy-encoded} or equation \ref{eqn:entropy-general} with some other upper bounds on $H(Z | Y)$ or $H(Z)$ to make an argument like Tao's.  We will consider two examples to illustrate this point.

\subsection{Example 1 (trivial)}
\label{subsec:trivial}
As a trivial example, suppose that we have run the MT algorithm for a \ksat problem and encoded the alternative information in $Y$ using a bit string of size at most $h$ for each iteration, as in Tao's proof.  Imagine removing one bit of information from each iteration's bit string in such a way that we can infer from it only a uniform distribution over $2^c$ potential flaws; call this reduced alternative information $Y'$, with encoding length $N' = N - ct$.  This ensures there are $2^{ct}$ possible flaw-encodings, and thus (by atomicity of the underlying MT algorithm for \ksat) our posterior on $Z$, having observed $Y'$, is a uniform distribution over $2^{ct}$ state vectors.  Then we could use Tao's argument to show that inequality \ref{eqn:entropy-encoded} is violated under exactly the conditions when it was violated for $Y$, since $H(Z | Y') + N' = H(Z | Y) + 2^{ct} + N - 2^{ct} = H(Z | Y) + N$.

We can use exactly the same argument to apply Tao's proof in a situation where the ambiguity in $P(Z | Y)$ is induced by non-atomicity of the transformation graph, rather than by the artificial removal of information from $Y$.  So why do we call this example trivial?  It only violates the first condition for atomicity in the framework of Achlioptas and Iliopoulos, which is that if we observe state $\sigma_{t+1}$ after a flaw $f$ is addressed, there is only one possibility for the previous state $\sigma_t$.  They call this condition ``syntactic,'' since it can always be addressed by expanding the flaw set to make flaws more granular.  For example, say that we have a modified version of \ksat in which $2^c$ out of $2^k$ assignments to each clause can make it fail.  We can rephrase any such problem as an ordinary \ksat problem by making $2^c$ copies of each clause, with each copy only disallowing one state.  This would increase the size of $F$ by $2^c$ and increase by $c$ the number of bits required to describe any flaw in the alternative information.  So the argument in our trivial example does not buy us any extra power in handling non-atomicity, only an alternative proof of Achlioptas' and Iliopoulos' comment.

\subsection{Example 2 (Uncountable feasible sets)}
\label{subsec:uncountable}
Consider the following feasibility problem:
\seqn
  \label{prob:feas}
  \operatorname{Find} & x \in \Reals^n \texttt{s.t.} \\
  x \in \mathcal{B}(1, n, ||\cdot||_\infty) \\
  A_i(x) = 0 \forall i \in [m] ,
\eeqn
where the $A_i$ are \emph{arbitrary} indicator functions $\Reals^{m} \to \{0,1\}$ and $\mathcal{B}(1, n, ||\cdot||_\infty)$ is the unit $\infty$-norm ball in $\Reals^n$ (a.k.a. the solid hypercube).  We will assume later that each function depends only on a subset of the elements of the vector $x$; let $\operatorname{vbl}(z) \subseteq [n]$ denote the indices on which a function depends, and let $y[\operatorname{vbl}(z)]$ denote the corresponding elements of a vector $y$.  Finally, let $\mu^d$ denote the uniform measure on $\mathcal{B}(1, d, ||\cdot||_\infty)$.

It is straightforward to write down an MT algorithm for solving this problem:

\begin{algorithm}[H]
\caption{An MT-like algorithm for problem \ref{prob:feas}.}
\label{alg:mt-feas}
\begin{algorithmic}[1]

\State Initialize $x \hasDist \operatorname{Unif(\mathcal{B}(1, n, ||\cdot||_\infty))}$.
\ForAll{$i \in [m]$}
  \Call{Fix}{i}
\EndFor

\Function{Fix}{i}
  \If{$A_i(x) = 0$}
    \Return
  \EndIf
  \State Resample $x[\operatorname{vbl}(A_i)]$ from $\mu^{|\operatorname{vbl}(A_i)}$.
  \ForAll{$j \in [m]: j \in \Gamma(i), A_j(x) = 1$}
    \Call{Fix}{j}
  \EndFor
\EndFunction

\end{algorithmic}
\end{algorithm}

Moser and Tardos prove that this algorithm finishes in polynomial time under the usual LLL conditions, but their proof relies on the finiteness of $n$ and $m$.  The proof of Achlioptas and Iliopoulos would need no assumptions on $n$ and weaker assumptions on $m$, but it fails when the sample space $\Omega$ is infinite (even countably so).  We will provide a proof that does not depend on the dimension that this algorithm finishes in expected polynomial time under conditions analogous to the LLL conditions.  Specifically, we will make the following assumptions:
\begin{enumerate}
  \item Each $A_i$ is a function of only $k$ elements of $x$.  \label{eqn:feas-sparsity-condition}
  \item Each $A_i$ is $\mu^n$-measurable and satisfies $\mu^n(\{x: A_i(x) = 1\}) \leq 2^{-k}$.  (That is, most of the hypercube is feasible for any single constraint.)  Otherwise the constraints can be arbitrarily complicated(!)  \label{eqn:feas-margprob-condition}
  \item The dependency graph induced by our constraints meets a slightly-strengthened version of the usual LLL condition: $2^{C_0} 2^{-k} d < 1$, for some absolute constant $C_0$.  (For reference, the usual condition is $e 2^{-k} (d + 1) < 1$.  The requirement for a strengthened condition may be an artifact of the proof.) \label{eqn:feas-lowdep-condition}
\end{enumerate}

This problem now has many of the ingredients of an LLL problem.  (In fact, we can encode any instance of \ksat that satisfies the LLL conditions in this problem, mapping $x_i$ to \texttt{true} if it is in $[0,1]$ and \texttt{false} otherwise, and encoding constraints accordingly.)  The process for encoding the alternative information and constructing a witness forest from it is almost identical to that in Tao's argument.  A technical difficulty is that the sample space is uncountable, so we can no longer speak of encoding a point using bits; this is where it seems useful to use the entropy inequality \ref{eqn:entropy-general} directly.  A more practical difficulty is that we can no longer exactly reconstruct the algorithm state given the history of corrected flaws.  To fix this, it will suffice to augment the alternative information slightly.

Let us first calculate $H(Z)$ for a run of algorithm \ref{alg:mt-feas} that takes $t$ steps.  This part is easy.  We define $Z$ to be the concatenation of the initial algorithm state (a draw from $\mu^n$) and the $t$ resampling steps (equivalent to draws from $\mu^k$).  When computing entropies, we use as our reference measure $\mu^n$.  Each of our events is independent, so $H(Z) = H(U([-1,1])) (n + t k) = n + k t$.

Now we define the alternative information $Y$ and find an upper bound for its entropy $H(Y)$.  We use almost the same construction as in Tao's proof: $Y$ is the concatenation of
\begin{enumerate}
  \item the witness forest (a number of bits, whose entropy is therefore bounded by its length),
  \item the end state (whose entropy is at most $n$, the same as that of the initial algorithm state),
  \item the initial set of violated constraints (which uses at most $m$ bits), and
  \item one additional piece of information for each iteration, which was not needed in Tao's argument.  To apply Tao's argument directly, we need to exactly reconstruct each state $\sigma_t$ from $Y$.  But at each iteration, even if we know $\sigma_{t+1}$ and the rest of $Y$ tells us exactly the flaw that has been addressed, we cannot in general exactly reconstruct the previous algorithm state $\sigma_t$.  However, we can bound its support, which bounds its entropy under the dominating measure $\mu^n$, and then include addition information in $Y$ sufficient to pin down $\sigma_t$ precisely.  Under our assumptions on the problem, it turns out that this additional data has entropy at most $1$.  Given only the values of $\sigma_t$ on the $n - k$ variables outside $\operatorname{vbl}(A_t)$, we know that $\mu^n(A_t | \{(\sigma_{t+1})_i: i \notin \operatorname{vbl}(A_t)\}) \leq 2^{-k}$, since $A_t$ is independent of those variables and $\mu(A_t) \leq 2^{-k}$ by assumption \ref{eqn:feas-margprob-condition}.  This happens inside a subset of size $2^{k}$, which has $\mu^n$-measure $2^{k-n}$, so the set of possible values for $\sigma_t$ is of $\mu^n$-measure $2^{-n}$, and its entropy (with $\mu^n$ as dominating measure) is $1$. %FIXME: Make the entropy argument more precise.  We are considering each item independently, which gives an upper bound, which is all we want.
\end{enumerate}

From these pieces and their individual entropy bounds (the sum of which bounds the joint entropy), $H(Y) \leq m + n + (k - C_0 + 2) t$, where $C_0$ is the constant in assumption \ref{eqn:feas-lowdep-condition}.

Finally, since we have pinned down $Z$ using $Y$, $H(Z | Y) = 0$.  Putting everything together, we get $m + n + (k - C_0 + 2) t \leq n + k t$, so that the algorithm must terminate (in expectation) before $t > \frac{m}{C_0 - 2}$.  We can choose $C_0 = 3$, giving us the requirement that $8 p d < 1$.  The algorithm terminates only when no constraint is violated, so the algorithm must find a satisfying solution in $m$ calls to \texttt{Fix}.  (Note that the ordinary non-constructive LLL applies to this problem, and our assumptions are slightly stronger than the LLL conditions, so a solution must exist.)

Some final notes on this algorithm:
\begin{enumerate}
  \item We chose our sample space to be a norm ball in finite dimensions, rather than say all of $\Reals^n$, precisely so that our sample space would be compact, allowing us to choose a simple probability measure as our dominating measure when computing entropies.  In general, to bound the number of steps taking by the algorithm, we need a probability measure that $Y$-almost surely dominates $P(Z | Y)$, for which the LLL-required marginal event probabilities hold, and which induces low dependency degree among the events.
  \item We chose the infinity norm ball in particular for a different reason: the sample space needs to be an axis-aligned rectangle, or it is not easy to ensure that the resampling step keeps us inside it while modifying only $k$ variables at a time and covering a large volume.  Again, this is not important for proving something about the number of steps taken by the algorithm, but it is important for the efficiency of the individual steps.
  \item Rather than augment $Y$ to pin down $Z$, we could imagine proving directly that $H(Z | Y)$ is small.  It is natural to view $Z$ and $Y$ as the states and observations of a hidden Markov model, respectively, and use any of the well-known algorithms for inference in HMMs to compute $P(Z | Y)$.  (Such inference could be quite difficult if the constraint functions are complicated, but for purposes of analysis we could imagine performing it.)  At least in this case, augmenting $Y$ was much simpler, but this makes the example weaker, since we are not taking advantage of our freedom in making $H(Z | Y) > 0$.
  \item With respect to the constraint functions themselves, the algorithm can be made efficient as long as we have an oracle that can efficiently find a violated constraint when there is one.  It may be possible to have an efficient version of the algorithm testing only a core of $\operatorname{poly}(n)$ ``core'' constraints using the ideas of Haeupler et al \cite{haeupler2011new}.
\end{enumerate}

%\subsection{Example 3 (An Unfocused Algorithm)}

\section{Future work}
Our primary interest is in seeing whether we can find a general framework in which to apply the relative entropy compression argument.
It would be helpful to find an example of an algorithm for which the ``focusing'' requirement of atomicity -- the requirement that only variables related to the addressed flaw be modified at each step -- is violated in a way that can be rescued by the relative entropy method. %FIXME: When example 3 above is filled in, delete this.
We would also like to understand whether the results of \cite{achlioptas2014random} can be restated using an entropy compression argument, with the hope that our results can be extended to the non-variable setting.

In our review of the literature, several other topics interested us, though we did not get to them in this paper.  Here is a brief literature review:
\begin{description}
  \item[Connections to graphical model inference:] Graphical models are a very natural way to think about constraint satisfaction problems, though they are perhaps too general; sampling from the conditional distribution in a graphical model allows us to solve any CSP.  These connections are touched on in \cite{freer2010probabilistic} and more thoroughly in \cite{haeupler2011new}.  In particular, Haeupler et al introduce the \emph{conditional-LLL} distribution: Conditional on no flaw happening, what is the distribution of variables (or, in the Achlioptas and Iliopoulos setup, the distribution of states)?  They prove that the MT algorithm approximately samples from this distribution.  MCMC is a very old method for sampling in graphical models, and the MT algorithm is a very special instance that seems to have no analogue in the machine learning literature.  It would be interesting to see whether an MT-like algorithm could be a useful version of MCMC for some statistical problems (perhaps bearing provable guarantees, via a relative entropy compression argument, that are not available for other algorithms), or whether the LLL could apply to existing algorithms from machine learning.
  \item[Parallel LLL algorithms:] Beck's algorithm was not parallel, but Alon \cite{alon1991parallel} published a parallel version soon after.  Moser and Tardos \cite{moser2010constructive} used a similar (but much simplified) idea in their original paper.  It is easy to see that the simple iterative version of their algorithm (algorithm \ref{alg:mt}) can be parallelized by resampling the variables of independent sets of events, though it is not trivial that this ensures the algorithm terminates in $O(\log m)$ such steps (in expectation).  More recently, Chung et al \cite{chung2014distributed} gave more practical parallel algorithms, which trade off the size of the independent set for the time taken to find such a set.  We are interested in making parallel MT-like algorithms fast in practice.  In other (unpublished) work, we have implemented some of these algorithms ourselves and found that they give reasonable performance on random \ksat instances, but they do not actually gain much from parallelism.  Given the potential connections between graphical model inference and MT-like algorithms, and the extensive work in parallelizing graphical model inference (for example, by Gonzalez et al \cite{gonzalez2011parallel}), we are curious whether such work can be applied to MT-like algorithms.  It would also be interesting to see whether the general algorithm of Achlioptas and Iliopoulos \cite{achlioptas2014random} can be parallelized in a similar fashion.
\end{description}

\bibliographystyle{plain}
\bibliography{\jobname}

\end{document}

### 2015-05-19 paper/cs270-report/proposal-270.tex
Add another paper and a poster.

\documentclass{article}
\usepackage{amsmath, amssymb}
\usepackage[retainorgcmds]{IEEEtrantools}
\usepackage{filecontents}
\usepackage{hyperref}
\author{Henry Milner}
\title{CS270 Project Proposal: Algorithmic Lovasz Local Lemma}
\date{4/3/15}

% Some convenience functions for homework problems.
\newcommand{\problem}[1]%
  {\section*{#1.}}

\newcommand{\problemSubpart}[1]%
  {\noindent\emph{#1.}}

\newcommand{\problemNamedSubpart}[1]%
  {\noindent\emph{#1}}

% Some convenience functions for note-taking.
\newcommand{\topic}[1]%
  {\section*{#1}}

% Some functions for general use.

\def\seqn#1\eeqn{\begin{align}#1\end{align}}

\newcommand{\vecName}[1]%
  {\boldsymbol{#1}}

\newcommand{\io}%
  {\text{ i.o. }}

\newcommand{\eventually}%
  {\text{ eventually }}

\newcommand{\tr}%
  {\text{tr}}

\newcommand{\Cov}%
  {\text{Cov}}

\newcommand{\adj}%
  {\text{adj}}

\newcommand{\funcName}[1]%
  {\text{#1}}

\newcommand{\hasDist}%
  {\sim}

\DeclareMathOperator*{\E}%
  {\mathbb{E}}

\newcommand{\Var}%
  {\text{Var}}

\newcommand{\std}%
  {\text{std}}

\newcommand{\grad}%
  {\nabla}

\DeclareMathOperator*{\argmin}{arg\,min}

\DeclareMathOperator*{\argmax}{arg\,max}

\newcommand{\inprod}[2]%
  {\langle #1, #2 \rangle}

\newcommand{\dd}[1]%
  {\frac{\delta}{\delta#1}}

\newcommand{\Reals}%
  {\mathbb{R}}

\newcommand{\indep}%
  {\protect\mathpalette{\protect\independenT}{\perp}} \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}

\newcommand{\defeq}%
  {\buildrel\triangle\over =}

\newcommand{\defn}[1]%
  {\emph{Definition: #1}\\}

\newcommand{\example}[1]%
  {\emph{Example: #1}\\}

\newcommand{\figref}[1]%
  {\figurename~\ref{#1}}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newenvironment{proof}[1][Proof]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\begin{filecontents}{\jobname.bib}
@article{erdos1975problems,
  title={Problems and results on 3-chromatic hypergraphs and some related questions},
  author={Erdos, Paul and Lov{\'a}sz, L{\'a}szl{\'o}},
  journal={Infinite and finite sets},
  volume={10},
  number={2},
  pages={609--627},
  year={1975}
}
@article{beck1991algorithmic,
  title={An algorithmic approach to the Lov{\'a}sz local lemma. I},
  author={Beck, J{\'o}zsef},
  journal={Random Structures \& Algorithms},
  volume={2},
  number={4},
  pages={343--365},
  year={1991},
  publisher={Wiley Online Library}
}
@article{moser2010constructive,
 author = {Moser, Robin A. and Tardos, G\'{a}bor},
 title = {A Constructive Proof of the General Lov\ÁSz Local Lemma},
 journal = {J. ACM},
 issue_date = {January 2010},
 volume = {57},
 number = {2},
 month = feb,
 year = {2010},
 issn = {0004-5411},
 pages = {11:1--11:15},
 articleno = {11},
 numpages = {15},
 url = {http://doi.acm.org/10.1145/1667053.1667060},
 doi = {10.1145/1667053.1667060},
 acmid = {1667060},
 publisher = {ACM},
 address = {New York, NY, USA},
 keywords = {Constructive proof, Lov\'{a}sz local lemma, parallelization},
} 
@article{haeupler2011new,
  title={New constructive aspects of the lovasz local lemma},
  author={Haeupler, Bernhard and Saha, Barna and Srinivasan, Aravind},
  journal={Journal of the ACM (JACM)},
  volume={58},
  number={6},
  pages={28},
  year={2011},
  publisher={ACM}
}
@inproceedings{freer2010probabilistic,
  title={When are probabilistic programs probably computationally tractable?},
  author={Freer, Cameron E and Mansinghka, Vikash K and Roy, Daniel M},
  booktitle={NIPS Workshop on Advanced Monte Carlo Methods with Applications},
  year={2010}
}
@article{wainwright2008graphical,
  title={Graphical models, exponential families, and variational inference},
  author={Wainwright, Martin J and Jordan, Michael I},
  journal={Foundations and Trends{\textregistered} in Machine Learning},
  volume={1},
  number={1-2},
  pages={1--305},
  year={2008},
  publisher={Now Publishers Inc.}
}
@inproceedings{papadimitriou1991selecting,
  title={On selecting a satisfying truth assignment},
  author={Papadimitriou, Christos H},
  booktitle={Foundations of Computer Science, 1991. Proceedings., 32nd Annual Symposium on},
  pages={163--169},
  year={1991},
  organization={IEEE}
}
@inproceedings{steurer2010fast,
  title={Fast SDP algorithms for constraint satisfaction problems},
  author={Steurer, David},
  booktitle={Proceedings of the twenty-first annual ACM-SIAM symposium on Discrete Algorithms},
  pages={684--697},
  year={2010},
  organization={Society for Industrial and Applied Mathematics}
}
@inproceedings{polik2007sedumi,
  title={SeDuMi: a package for conic optimization},
  author={Polik, Imre and Terlaky, Tamas and Zinchenko, Yuriy},
  booktitle={IMA workshop on Optimization and Control, Univ. Minnesota, Minneapolis},
  year={2007}
}
\end{filecontents}
\immediate\write18{bibtex \jobname}

\begin{document}
\maketitle

For my project, I will study the algorithmic Lovasz Local Lemma (LLL).  I will begin with a survey of the literature on the subject.  This will cover the nonconstructive lemma due to Erdos and Lovasz \cite{erdos1975problems} (briefly, since it was covered in class), then the first algorithm by Beck \cite{beck1991algorithmic}, and finally the general algorithm and information-theoretic proof by Moser and Tardos \cite{moser2010constructive}.

After the survey, I will (time permitting) explore two questions that motivate my interest in this subject:

\subsection{Connections to MCMC Sampling}
In a 2010 paper, Freer et al \cite{freer2010probabilistic} point out that Moser's algorithm is similar to MCMC sampling in probabilistic graphical models.  In the usual algorithmic LLL setup, we let $\mathcal{E} = \{E_1, \cdots, E_n\}$ be a set of events determined by marginally mutually independent random variables $\mathcal{A} = \{A_1, \cdots, A_m\}$.  We could think of this as a directed probabilistic graphical model (DPGM), with an arrow from each variable $A_i$ to the events $c(A_i)$ dependent on it.  A satisfying assignment is anything that has positive probability when we condition on none of the events happening.  Then Moser's algorithm repeatedly picks parent sets $\pi(E_j)$ that have probability $0$ under the current assignment and resamples them from their marginal distribution.

This is close to a Metropolis-Hastings sampling scheme: the proposal distribution resamples all parents of an arbitrary event, and the proposal is rejected if the event already didn't happen under the previous assignment.  Moser's algorithm cannot actually be a proper Metropolis-Hastings algorithm, since the Markov chain produced when we run it eventually reaches an absorbing state.  Still, Srinivasan et al \cite{haeupler2011new} show that Moser's algorithm approximately samples from the ```conditional LLL''' distribution -- the joint distribution of the $A_i$ conditional on none of the events in $\mathcal{E}$ happening -- which is all that we want out of a sampling algorithm.

I would like to investigate whether an algorithm like Moser's could perform fast sampling in PDGMs that satisfy an LLL-like condition.  The proof techniques in \cite{moser2010constructive} and \cite{haeupler2011new} seem quite different from those in the literature on probablistic inference.  For example, the LLL condition applies to graphs with very high treewidth (at least $k$ in the $k$-SAT problem), while exact inference in general graphs is exponentially hard in the treewidth\cite{wainwright2008graphical}.  I also want to see whether Moser's algorithm and proof can be modified to work when the events in $\mathcal{E}$ are merely dependent on the $A_i$, and not fixed deterministically by them.

\subsection{Practicality}
Is Moser's algorithm useful in practice?  It seems that common local search methods for problems like $k$-SAT and graph coloring are similar to Moser's algorithm, but not exactly the same.  Since Moser's algorithm is so simple (and so similar to algorithms like Papadimitriou's algorithm for $2$-SAT \cite{papadimitriou1991selecting}), it seems likely that practitioners would have tried it heuristically.  But I have not (so far) found any studies of this; perhaps this idea has not yet migrated out of the theory community.  So I would like to write a reasonably performant implementation of the parallel version of Moser's algorithm, and then compare it on CSP benchmarks with off-the-shelf solvers and perhaps with an SDP relaxation \cite{steurer2010fast} using an off-the-shelf SDP solver like SeDuMi \cite{polik2007sedumi}.

\bibliographystyle{plain}
\bibliography{\jobname}

\end{document}

### 2015-05-28 paper/cs270-report/proposal.tex
Some work on an expanded version of the 270 survey.

\documentclass{article}
\usepackage{amsmath, amssymb}
\usepackage[retainorgcmds]{IEEEtrantools}
\usepackage{algpseudocode}
\usepackage{filecontents}
\usepackage{hyperref}
\author{Henry Milner}
\title{CS270 Project Proposal: Algorithmic Lovasz Local Lemma}
\date{4/3/15}

% Some convenience functions for homework problems.
\newcommand{\problem}[1]%
  {\section*{#1.}}

\newcommand{\problemSubpart}[1]%
  {\noindent\emph{#1.}}

\newcommand{\problemNamedSubpart}[1]%
  {\noindent\emph{#1}}

% Some convenience functions for note-taking.
\newcommand{\topic}[1]%
  {\section*{#1}}

% Some functions for general use.

\def\seqn#1\eeqn{\begin{align}#1\end{align}}

\newcommand{\vecName}[1]%
  {\boldsymbol{#1}}

\newcommand{\io}%
  {\text{ i.o. }}

\newcommand{\eventually}%
  {\text{ eventually }}

\newcommand{\tr}%
  {\text{tr}}

\newcommand{\Cov}%
  {\text{Cov}}

\newcommand{\adj}%
  {\text{adj}}

\newcommand{\funcName}[1]%
  {\text{#1}}

\newcommand{\hasDist}%
  {\sim}

\DeclareMathOperator*{\E}%
  {\mathbb{E}}

\newcommand{\Var}%
  {\text{Var}}

\newcommand{\std}%
  {\text{std}}

\newcommand{\grad}%
  {\nabla}

\DeclareMathOperator*{\argmin}{arg\,min}

\DeclareMathOperator*{\argmax}{arg\,max}

\newcommand{\inprod}[2]%
  {\langle #1, #2 \rangle}

\newcommand{\dd}[1]%
  {\frac{\delta}{\delta#1}}

\newcommand{\Reals}%
  {\mathbb{R}}

\newcommand{\indep}%
  {\protect\mathpalette{\protect\independenT}{\perp}} \def\independenT#1#2{\mathrel{\rlap{$#1#2$}\mkern2mu{#1#2}}}

\newcommand{\defeq}%
  {\buildrel\triangle\over =}

\newcommand{\defn}[1]%
  {\emph{Definition: #1}\\}

\newcommand{\example}[1]%
  {\emph{Example: #1}\\}

\newcommand{\figref}[1]%
  {\figurename~\ref{#1}}

\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}
\newenvironment{proof}[1][Proof]{\begin{trivlist}
\item[\hskip \labelsep {\bfseries #1}]}{\end{trivlist}}

\begin{filecontents}{\jobname.bib}
@article{erdos1975problems,
  title={Problems and results on 3-chromatic hypergraphs and some related questions},
  author={Erdos, Paul and Lov{\'a}sz, L{\'a}szl{\'o}},
  journal={Infinite and finite sets},
  volume={10},
  number={2},
  pages={609--627},
  year={1975}
}
@article{beck1991algorithmic,
  title={An algorithmic approach to the Lov{\'a}sz local lemma. I},
  author={Beck, J{\'o}zsef},
  journal={Random Structures \& Algorithms},
  volume={2},
  number={4},
  pages={343--365},
  year={1991},
  publisher={Wiley Online Library}
}
@article{moser2010constructive,
 author = {Moser, Robin A. and Tardos, G\'{a}bor},
 title = {A Constructive Proof of the General Lov\ÁSz Local Lemma},
 journal = {J. ACM},
 issue_date = {January 2010},
 volume = {57},
 number = {2},
 month = feb,
 year = {2010},
 issn = {0004-5411},
 pages = {11:1--11:15},
 articleno = {11},
 numpages = {15},
 url = {http://doi.acm.org/10.1145/1667053.1667060},
 doi = {10.1145/1667053.1667060},
 acmid = {1667060},
 publisher = {ACM},
 address = {New York, NY, USA},
 keywords = {Constructive proof, Lov\'{a}sz local lemma, parallelization},
} 
@article{haeupler2011new,
  title={New constructive aspects of the lovasz local lemma},
  author={Haeupler, Bernhard and Saha, Barna and Srinivasan, Aravind},
  journal={Journal of the ACM (JACM)},
  volume={58},
  number={6},
  pages={28},
  year={2011},
  publisher={ACM}
}
@inproceedings{chung2014distributed,
  title={Distributed algorithms for the Lov{\'a}sz local lemma and graph coloring},
  author={Chung, Kai-Min and Pettie, Seth and Su, Hsin-Hao},
  booktitle={Proceedings of the 2014 ACM symposium on Principles of distributed computing},
  pages={134--143},
  year={2014},
  organization={ACM}
}
@inproceedings{freer2010probabilistic,
  title={When are probabilistic programs probably computationally tractable?},
  author={Freer, Cameron E and Mansinghka, Vikash K and Roy, Daniel M},
  booktitle={NIPS Workshop on Advanced Monte Carlo Methods with Applications},
  year={2010}
}
@article{wainwright2008graphical,
  title={Graphical models, exponential families, and variational inference},
  author={Wainwright, Martin J and Jordan, Michael I},
  journal={Foundations and Trends{\textregistered} in Machine Learning},
  volume={1},
  number={1-2},
  pages={1--305},
  year={2008},
  publisher={Now Publishers Inc.}
}
@inproceedings{papadimitriou1991selecting,
  title={On selecting a satisfying truth assignment},
  author={Papadimitriou, Christos H},
  booktitle={Foundations of Computer Science, 1991. Proceedings., 32nd Annual Symposium on},
  pages={163--169},
  year={1991},
  organization={IEEE}
}
@inproceedings{steurer2010fast,
  title={Fast SDP algorithms for constraint satisfaction problems},
  author={Steurer, David},
  booktitle={Proceedings of the twenty-first annual ACM-SIAM symposium on Discrete Algorithms},
  pages={684--697},
  year={2010},
  organization={Society for Industrial and Applied Mathematics}
}
@inproceedings{polik2007sedumi,
  title={SeDuMi: a package for conic optimization},
  author={Polik, Imre and Terlaky, Tamas and Zinchenko, Yuriy},
  booktitle={IMA workshop on Optimization and Control, Univ. Minnesota, Minneapolis},
  year={2007}
}
\end{filecontents}
\immediate\write18{bibtex \jobname}

\begin{document}
\maketitle

The algorithmic Lovasz Local Lemma (LLL) is a recent-developed class of randomized algorithms for solving certain easy instances of hard combinatorial problems.  The seminal paper in the field is by Moser and Tardos \cite{moser2010constructive} in 2010.  For many problems, LLL algorithms theoretically lend themselves well to parallelism, as demonstrated in more recent work \cite{chung2014distributed,haeupler2011new}.  However, we are aware of no published empirical tests of the efficiency of LLL algorithms in practice.  In our project, we will implement parallel LLL algorithms for graph coloring and k-SAT, and we will compare their performance on benchmarks against other algorithms.

\subsection{A brief introduction to LLL algorithms}
The basic setup for LLL algorithms is as follows: Let $\mathcal{A} = \{A_1, \cdots, A_m\} \in 2^m$ be a set of discrete variables (which we take to be binary for sake of exposition), and let $\mathcal{E} = \{E_1, \cdots, E_n\}$ be a set of functions $[K]^m \to \{\operatorname{true},\operatorname{false}\}$ mapping the variables to truth values.  The $E_i$ are called ``events,'' and $E_i$ is said to ``happen'' if it is true.  We would like to find an assignment of the variables so that none of the events happen.  For example, in k-SAT, the $A_i$ are the problem variables, and each $E_i$ is a clause involving $k$ variables.  The simplest version of the algorithm is \emph{extremely} simple:

\begin{algorithmic}
\State Sample all the $A_i$ IID Bernoulli($1/2$).
\While{Any of the $E_j$ happen}
  \State Let $E_j$ be an arbitrary event that happens.
  \State Sample each $A_i \in \{A_i: E_j\text{ depends on }A_i\}$ IID Bernoulli($1/2$).
\EndWhile
\end{algorithmic}

Of course, the decision version of k-SAT is NP-hard, so we would not expect this algorithm to work for all instances.  Moser and Tardos \cite{moser2010constructive} prove that it runs in expected polynomial time when the when the events are not ``too dependent.''  Let $p$ be the maximum probability that any single event happens under a random assignment to all the variables.  Construct an undirected dependency graph for the events, in which there is an edge between $E_j$ and $E_k$ if they are statistically dependent under random assignment to the $A_i$.  In k-SAT, there is an edge between clauses $E_j$ and $E_k$ if they share any variable.  Let $d$ be the maximum degree of events in this graph.  Then the above algorithm finishes in expected polynomial time if $e p d < 1$, where $e$ is the base of the natural logarithm.

The dependency graph is useful for seeing an easy way to parallelize the simple LLL algorithm.  Notice that if independent events $E_j$ and $E_k$ happen under the current assignment to $\mathcal{A}$, and event $E_j$'s variables are chosen for resampling, then (by independence) $E_k$ will still happen after the resampling, so the next iteration could resample it.  This means we can resample variables depended-on by any independent set of events that currently happen, and this resampling can be done in parallel.  Chung et al \cite{chung2014distributed} and Srinivasan et al \cite{haeupler2011new} give alternative parallel algorithms that are less obviously correct, but which can be run on distributed computers.

\subsection{Project goals}

% Is Moser's algorithm useful in practice?  It seems that common local search methods for problems like $k$-SAT and graph coloring are similar to Moser's algorithm, but not exactly the same.  Since Moser's algorithm is so simple (and so similar to algorithms like Papadimitriou's algorithm for $2$-SAT \cite{papadimitriou1991selecting}), it seems likely that practitioners would have tried it heuristically.  But I have not (so far) found any studies of this; perhaps this idea has not yet migrated out of the theory community.  So I would like to write a reasonably performant implementation of the parallel version of Moser's algorithm, and then compare it on CSP benchmarks with off-the-shelf solvers and perhaps with an SDP relaxation \cite{steurer2010fast} using an off-the-shelf SDP solver like SeDuMi \cite{polik2007sedumi}.

\bibliographystyle{plain}
\bibliography{\jobname}

\end{document}

### 2015-05-28 paper/nonatomic/paper.tex
Some work on an expanded version of the 270 survey.

--- 2015-05-19 paper/cs270-report/paper.tex
@@
 }
+@article{formentin2009symmetric,
+  title={A symmetric entropy bound on the non-reconstruction regime of Markov chains on Galton-Watson trees},
+  author={Formentin, Marco and K{\"u}lske, Christof},
+  journal={arXiv preprint arXiv:0903.2962},
+  year={2009}
+}
 @book{alon2004probabilistic,
@@
 }
+@article{prufer1918neuer,
+  title={Neuer beweis eines satzes {\"u}ber permutationen},
+  author={Pr{\"u}fer, Heinz},
+  journal={Arch. Math. Phys},
+  volume={27},
+  pages={742--744},
+  year={1918}
+}
+@inproceedings{neville1953codifying,
+  title={The codifying of tree-structure},
+  author={Neville, EH},
+  booktitle={Mathematical Proceedings of the Cambridge Philosophical Society},
+  volume={49},
+  number={03},
+  pages={381--385},
+  year={1953},
+  organization={Cambridge Univ Press}
+}
+@article{otter1948number,
+  title={The number of trees},
+  author={Otter, Richard},
+  journal={Annals of Mathematics},
+  pages={583--599},
+  year={1948},
+  publisher={JSTOR}
+}
+@inproceedings{jacobson1989space,
+  title={Space-efficient static trees and graphs},
+  author={Jacobson, Guy},
+  booktitle={Foundations of Computer Science, 1989., 30th Annual Symposium on},
+  pages={549--554},
+  year={1989},
+  organization={IEEE}
+}
+@book{drmota2009random,
+  title={Random trees: an interplay between combinatorics and probability},
+  author={Drmota, Michael},
+  year={2009},
+  publisher={Springer Science \& Business Media}
+}
 \end{filecontents}
@@
 In the less abstract \emph{variable version} of the LLL problem, the space is a product space of, say, $n$ independent variables and each event corresponds to a binary function of some of the variables.
-In 2010, Moser and Tardos \cite{moser2009constructive,moser2010constructive} made the LLL method constructive for variable LLL problems, giving a surprisingly simple expected-polynomial-time algorithm to find variable assignments that make no events happen.
+In 2010, Moser and Tardos \cite{moser2009constructive,moser2010constructive} made the LLL method efficiently constructive for variable LLL problems, giving a surprisingly simple expected-polynomial-time algorithm to find variable assignments that make no events happen.
 They used a novel entropy compression argument, proving that their algorithm could be used to compress random bits if it took more than a certain amount of expected time to find a solution.
@@
 \section{Introduction}
-The \lovasz Local Lemma (LLL) is a probabilistic method for proving the existence of elements of a probability space $\mathcal{P} = (\Omega, \Sigma, \mu)$ that miss all of a set of events $\mathcal{A} = \{A_1, \cdots, A_m\}$, the search for which we call an ``LLL problem''.
+The \lovasz Local Lemma (LLL) is a probabilistic method for proving the existence of elements of a probability space $\mathcal{P} = (\Omega, \Sigma, \mu)$ that are flawless -- elements that miss all of a set of bad events (``flaws'') $\mathcal{A} = \{A_1, \cdots, A_m\}$.
+We call any search for such elements an ``LLL problem''.
 It was introduced by \lovasz and Erd\"{o}s in 1975 \cite{erdos1975problems}; the proof is a fairly straightforward application of the probabilistic method \cite{alon2004probabilistic}.
 The LLL has many applications in combinatorial search problems.
-A useful restricted version of the problem is the ``variable version,'' in which $\mathcal{P}$ is a product space of $n$ independent random variables $X = (X_1, \cdots, X_n)$.
-(We will sometimes write $A_i(X)$ for the indicator that event $i$ happens when the variables take value $X$.)  For example, we can use the LLL to prove the existence of solutions to certain \ksat problems by imagining a probability space on $n$ random variables in which each event $A_j$ occurs if a corresponding disjunctive clause $C_j \in \mathcal{C} = \{C_1, \cdots, C_m\}$ of $k$ variables is false.
-In that setup, ``missing all events'' corresponds to satisfying all clauses.
+
+A less abstract (though somewhat restricted) version of the problem is the ``variable version,'' in which each element of $\Omega$ can be expressed as a vector of $n$ variables, and those variables are independent.
+More formally, in a variable-LLL problem, $\mathcal{P}$ is a product of $n$ probability spaces on which we define $n$ corresponding independent random variables $(X_1, \cdots, X_n) = X$ product measure $\prod_{j = 1}^{n} \mu_j  = \mu$.
+Each $A_i$ is a set of values of $X$ that exhibit some abstract ``flaw.''  In a slight abuse of notation, we will write an event $A_i$ as a function of $X$ that returns $1$ when the corresponding event happens and $0$ otherwise.
+As a function, $A_i$ may be functionally dependent on only a subvector of $X$, and we will write $\operatorname{vbl}(A_i) \subseteq [n]$ for the indices on which function $A_i$ depends.
+Finally, following \cite{achlioptas2014random}, we will denote by $U(X)$ the events that happen under $X$, or more colloquially the ``flaws of $X$.''
+
+For example, we can use the LLL to prove the existence of solutions to certain \ksat problems.
+In applying the LLL to \ksat, we imagine a probability space on $n$ independent uniform random binary variables in which each event $A_j$ occurs if a corresponding disjunctive clause $C_j \in \mathcal{C} = \{C_1, \cdots, C_m\}$ of $k$ of the variables is false under the variable setting $X$.
+(We denote by $\operatorname{vbl}(C_j)$ the set of $k$ variables of clause $C_j$.)  A flawless element $X$ is just one that satisfies all clauses.
 Not every \ksat problem has a satisfying solution, so we should expect the conditions of the LLL to be restrictive.
@@
 
-A seminal paper by Beck \cite{beck1991algorithmic} showed that the LLL could be made \emph{algorithmic} -- that is, when the LLL conditions hold, certain algorithms will \emph{find} a solution quickly.
+The original LLL does not tell us how to \emph{construct} a solution when its conditions are met.
+There is an obvious Las Vegas algorithm for finding a solution -- since a solution exists, just sample from $\mathcal{P}$ until one is found.
+In fact, the LLL guarantees more than the existence of a solution; it provides a lower bound on $\mu(\cap_{i \in [m]} \bar{A_i})$.
+Unfortunately, this lower bound can be exponentially small, so there is no guarantee that the Las Vegas algorithm is efficient.
+But a seminal paper by Beck \cite{beck1991algorithmic} showed that the LLL \emph{could} be made algorithmic.
+That is, when the LLL conditions hold, certain algorithms will find a solution quickly.
 This started a literature around the ``constructive'' or ``algorithmic'' LLL.
@@
 \begin{algorithmic}
-\State Initialize the $X_i$ IID uniform.
-\While{Any of the $A_j$ happen under the current value of $X$}
-  \State Let $A_j$ be an arbitrary event that currently happens.
-  \State Resample each $X_i \in \operatorname{vbl}(A_j)$ IID uniform.
+\Function{Solve}{$\mathcal{P}$, $\mathcal{A}$}
+  \State Initialize $X \hasDist \mu$.
+  \While{$X$ has a flaw}
+    \State Let $A_j$ be an arbitrary flaw of $X$.
+    \State Resample each $X_i \in \operatorname{vbl}(A_j) \hasDist \mu_i$.
   \EndWhile
+\EndFunction
 \end{algorithmic}
@@
 
-For example, specialized to \ksat, the MT algorithm becomes:
+For example, specialized to our formulation of \ksat, the MT algorithm becomes:
 
@@
 \begin{algorithmic}[1]
-\State Initialize the $X_i$ IID uniform Bernoulli.
-\While{Any clause is violated by $X$}
+\Function{Solve}{$n$, $\mathcal{C}$}
+  \State Initialize $X_1, \cdots, X_n$ i.i.d. uniform Bernoulli.
+  \While{Any clause in $\mathcal{C}$ is violated by $X$}
     \State Let $C_j$ be an arbitrary clause that is violated.
-  \State Resample each $X_i \in \operatorname{vbl}(C_j)$ IID uniform Bernoulli.
+    \State Resample each $X_i \in \operatorname{vbl}(C_j)$ i.i.d. uniform Bernoulli.
   \EndWhile
+\EndFunction
+
 \end{algorithmic}
@@
   \item[$e$] is the base of the natural logarithm;
-  \item[$p$] is the maximum marginal probability of any event when $X \hasDist \operatorname{Unif}(\Omega)$; and
-  \item[$d$] is the maximum degree of any event in the \emph{dependency graph} on events, in which two events are connected by an edge if they are statistically dependent when $X \hasDist \operatorname{Unif}(\Omega)$ .
+  \item[$p$] is the maximum marginal probability of any flaw under distribution $\mu$, $\sup_{i \in [m]} \mu(A_i)$; and
+  \item[$d$] is the maximum degree of any flaw in the \emph{dependency graph} on flaws, in which two flaws are connected by an edge if they are statistically dependent when $X \hasDist \mu$.
+(We will denote by $\Gamma(i)$ the neighborhood of $i$, the set of indices of flaws that neighbor flaw $i$ in this dependency graph.
+$\Gamma^{+}(i) = \Gamma(i) \cup \{i\}$ is the inclusive neighborhood.)  Then $d = \max_{i \in [m]} |\Gamma(i)|$ and $d+1 = \max_{i \in [m]} |\Gamma^{+}(i)|$.
 \end{description}
 
+Their proof uses a novel \emph{entropy compression} argument.
+They show that the MT algorithm must terminate in a certain number of steps (on average) or it would be possible to represent the algorithm's state trajectory using fewer random bits than were used to generate it, which is impossible.
+Tao \cite{tao2009entropy} and Fortnow \cite{fortnow2009entropy} make this argument more explicit.
+Notice that taking $\mu$ to be the uniform distribution maximizes the entropy created at each step of the algorithm (at least on finite spaces), so it is easiest to make an entropy compression argument in that case.
+
 After the Moser and Tardos paper, interest in the subject has only increased; follow-up work has given new algorithms, new problem domains, new applications, and a range of sufficient conditions weaker than the LLL condition.
+Our interest, and the subject of this paper, is in strengthening the Moser-Tardos result from the \emph{variable} version of the problem to more general versions, under analogous conditions.
+The key proof idea is taking the entropy compression argument seriously.
+%FIXME: Language.
+Sounds arrogant.
 
 \subsection{This paper}
-In this paper, we first review the result by Moser and Tardos, using a more illuminating proof by Tao \cite{tao2009entropy} and Fortnow \cite{fortnow2009entropy} that directly highlights the role of entropy compression.
-In section \ref{sec:perfect}, we review the extension of the algorithmic LLL to non-variable problems by Achlioptas and Iliopoulos \cite{achlioptas2014random}.
-Using these two ideas, in section \ref{sec:nonatomic} we sketch an argument that suggests the entropy compression argument could be extended to novel LLL-like applications in which the state vector of an algorithm cannot be uniquely recovered.
-Achlioptas and Iliopoulos call such algorithms (or rather, technically, their associated random walk graphs) ``non-atomic''.
-The original work in this paper is preliminary and incomplete, and we look forward to improving it in the near future.
+In this paper, we first review the result by Moser and Tardos, using the more illuminating proof by Tao \cite{tao2009entropy} and Fortnow \cite{fortnow2009entropy} that directly highlights the role of entropy compression.
+In section \ref{sec:perfect}, we review an extension of the algorithmic LLL to non-variable problems by Achlioptas and Iliopoulos \cite{achlioptas2014random}.
+Achlioptas and Iliopoulos give a general condition on \emph{algorithms} rather than problems, and bring us partway to a general constructive LLL, but they require that the space $\Omega$ is finite, and this is the restriction we will remove.
+In section \ref{sec:examples} we give a series of examples suggesting the entropy compression argument could be fruitfully extended to novel LLL-like applications in which the state vector of an algorithm cannot be uniquely recovered.
+Achlioptas and Iliopoulos call such algorithms (or, technically, their associated random walk graphs) ``non-atomic''.
+By removing the atomicity assumption, we can develop a more general constructive LLL.
+In section \ref{sec:nonatomic}, we give a formal proof.
+%FIXME: Language.
 
@@
 \begin{algorithm}[H]
-\caption{A more complicated recursive version of the Moser-Tardos (MT) algorithm.  Here $\Gamma(i)$ denotes the neighbors of event $i$ in the event dependency graph.}
+\caption{A more complicated recursive version of the Moser-Tardos (MT) algorithm.
+This version replaces the \emph{arbitrary} order of resampling in algorithm \ref{alg:mt} with a fixed order that resamples neighborhoods of events in the dependency graph.}
 \label{alg:mt-recursive}
 \begin{algorithmic}
-\State Initialize the $X_i$ IID uniform.
+\Function{Solve}{$\mathcal{P}$, $\mathcal{A}$}
+  \State Initialize $X \hasDist \mu$.
   \ForAll{$i \in [m]$}
@@
   \EndFor
+  \Return $X$
+\EndFunction
 
@@
   \EndIf
-  \State Resample each $X_i \in \operatorname{vbl}(A_i)$ IID uniform.
-  \ForAll{$j \in [m]: j \in \Gamma(i), A_j(X)$}
+  \State \Call{Resample}{i}
+  \ForAll{$k \in [m] \texttt{s.t.} k \in \Gamma^{+}(i) \texttt{ and } A_k(X) = 1$}
     \Call{Fix}{j}
   \EndFor
+\EndFunction
+
+\Function{Resample}{i}
+  \State Resample each $X_j \in \operatorname{vbl}(A_i) \hasDist \mu_j$.
 \EndFunction
@@
 
-We would like to prove that this algorithm terminates in polynomial time when the LLL condition is met.
-We will briefly review an argument by Tao \cite{tao2009entropy} and Fortnow \cite{fortnow2009entropy} that emphasizes the entropy compression aspect of the proof.
-The argument is specialized to \ksat to make the exposition simpler.
+An an introduction to later parts of this paper, we will give a proof that this algorithm terminates in polynomial time when the LLL condition is met.
+Our argument mostly follows one by Tao \cite{tao2009entropy} and Fortnow \cite{fortnow2009entropy} that emphasizes the entropy compression aspect of the proof.
+The argument is specialized to \ksat to make the exposition simpler, but we will later discuss the generalization to arbitary LLL problems.
+We will prove that the specialization of algorithm \ref{alg:mt-recursive} to \ksat terminates before $m/2$ calls to \texttt{Resample}.
+%FIXME: Might be extra O(1) terms.
+Recall that in the specialization to \ksat, $\mu$ is the uniform distribution on $\{0,1\}^n$-vectors and $A_i(X) = 1$ if clause $C_i$ is false.
 
-The general idea is to imagine running algorithm \ref{alg:mt-recursive} until \texttt{Fix} is called $t$ times.
-If $i(s)$ denotes the argument to \texttt{Fix} on iteration $s$, then after $t$ calls, $\sum_{j = 1}^{t} \operatorname{vbl}(A_{t(j)}) = k t$ binary variables have been resampled independently using some random source, and the initial assignment used $n$ bits.
-Together, the algorithm state vector, which we call $Z$, encodes $n + k t$ purely random bits.
-We will use the assumption $2^C_0 p d < 1$ to give a method to encode the algorithm state in an alternative way $Y$, which we call the ``alternative information''.
-(We will use $C_0 = 3$, though the proof is possible under the weaker condition $e p (d + 1) < 1$.)  When $t$ is large, $Y$ will contain fewer bits than $Z$, which contradicts Shannon's source coding theorem \cite{shannon2001mathematical} (it is impossible to encode purely random bits with a scheme that uses fewer bits in expectation).
+%FIXME: This could do with division into subsections, since it is so long.
 
-We note first that, given an innocuous assumption about the problem (which is true for \ksat and can be made true for other problems), there is a 1-to-1 correspondence between the pair $((A_{t(i)})_{i=1}^{t}, \sigma_{t+1})$ and $Z$.
-Here $\sigma_t$ denotes the state of $X$ before call $t$.
-The assumption we need is that each event $A_i$ happens for only one setting of $\operatorname{vbl}(A_i)$.
+\subsection{A nice property of algorithm \ref{alg:mt-recursive}}
+First, algorithm \ref{alg:mt-recursive} has a nice property that the simpler algorithm \ref{alg:mt} does not have, which motivates our switch to the more complicated version.
+If \texttt{Fix}$(i)$ returns, $X$ cannot have flaw $A_i$, nor can it have any flaw it did not have before \texttt{Fix}$(i)$ was called.
+The proof is by induction on the tree of calls to \texttt{Resample} caused by calling \texttt{Fix}$(i)$, which we (following Moser and Tardos) call a ``witness tree.''  In the witness tree (which we will be important later), each node is labeled (possibly nonuniquely) with a flaw index.
+A call to \texttt{Fix}$(i)$ may cause a call to \texttt{Resample}$(i)$ and zero or more additional calls to \texttt{Fix} in the following loop.
+There is a node in the tree for the call to \texttt{Resample}$(i)$, and that node has a child for each call to \texttt{Resample} caused by the calls to \texttt{Fix} in the loop.
+%FIXME: Picture
+
+The inductive argument is fairly simple.
+If the tree is empty, then $X$ must not have flaw $A_i$, and no resampling occurred, so no additional flaws are added.
+Otherwise, the variables $\operatorname{vbl}(A_i)$ were resampled and then calls to \texttt{Fix} were made for elements of clause $A_i$'s inclusive neighborhood (including $A_i$ itself) that were unsatisfied.
+No other clauses could have become unsatisfied by this resampling, since clauses outside $\Gamma^{+}(i)$ do not have variables $\operatorname{vbl}(A_i)$.
+By induction, these calls to \texttt{Fix} add no new flaws and satisfy each clause in $\Gamma^{+}(i)$.
+So our call to \texttt{Fix}$(i)$ added no new flaws and ensured that $A_i$ was satisfied.
+%FIXME: A little loose with the A_i and C_i notation here.
+%FIXME: Should merge the witness tree description here, or break it into its own section.
+
+This property implies that \ref{alg:mt-recursive} finds a satisfying solution if \texttt{Solve} returns at all, since it loops over all flaws and fixes each without introducing any new ones.
+So it remains to show when (and how quickly) that happens.
+
+We should emphasize that this fact is only \emph{convenient}, not necessary, for the following proof.  Algorithm \ref{alg:mt} enjoys many of the same guarantees as algorithm \ref{alg:mt-recursive}.
+
+\subsection{The entropy compression argument}
+The idea is to imagine running algorithm \ref{alg:mt-recursive} until \texttt{Resample} is called $t$ times.
+We will give a scheme for compressing random bits -- which is impossible -- if $t$ is too large in expectation.
+Let $\sigma_s$ denote the value of $X$ at the beginning of the $s$th call to \texttt{Resample}, and let $i(s)$ denote the argument to \texttt{Resample} on that iteration.
+We will say that $A_{i(s)}$ was ``addressed'' on iteration $s$.
+After $t$ calls, $\sum_{j = 1}^{t} \operatorname{vbl}(A_{t(j)}) = k t$ binary variables have been resampled independently using some random source.
+The initial assignment used $n$ bits.
+Let $Z \in \Omega^{t+1} = (\sigma_0, \sigma_1, \cdots, \sigma_t)$ denote the algorithm state vector.
+Then $Z$ encodes $n + k t$ purely random bits.
+We will use the assumption $2^C_0 p (d+1) < 1$ to give a method to encode the algorithm state in an alternative way $Y$, which we call the ``alternative information''.
+(We will use $C_0 = 2$ and later $C_0 = \log_2 \frac{1}{0.3383}$, though the tightest constant is $C_0 = \log_2 e$.)  When $t$ is large on average, $Y$ will contain fewer bits than $Z$ on average.
+That contradicts Shannon's source coding theorem \cite{shannon2001mathematical}, which formalizes the intuitive fact that it is impossible to encode purely random bits with a scheme that uses fewer bits in expectation.
+
+We note first that, given an innocuous assumption about the problem (which is true for \ksat and can be made true for other problems), there is a 1-to-1 correspondence between the pair $((A_{t(i)})_{i=1}^{t}, \sigma_{t})$ and $Z$.
+In other words, we assume that if we know the final return value and the sequence of flaws that were addressed, we know the whole sequence of states the algorithm passed through.
+The assumption we need is that each flaw $A_i$ happens for only one setting of $\operatorname{vbl}(A_i)$.
 For example, in \ksat, a disjunctive clause can only be violated by one setting of its variables.
 In section \ref{subsec:trivial} we will explain why this assumption is innocuous in general.
-Under the assumption, there is an easy inductive argument: If $\sigma_{i+1}$ and $A_{t(i)}$ are known, then $\sigma_i$ must equal $\sigma_{i+1}$ outside $\operatorname{vbl}(A_{t(i)})$, and on $\operatorname{vbl}(A_{t(i)})$ it must equal the one violating setting, since it was chosen to be \texttt{Fix}ed.
-But $\sigma_{t+1}$ and all the events are known, so the whole state vector $Z$ (including the initial state) is determined.
+Under the assumption, there is an easy inductive argument: If $\sigma_{i+1}$ and $A_{t(i)}$ are known, then $\sigma_i$ must equal $\sigma_{i+1}$ outside $\operatorname{vbl}(A_{t(i)})$, and on $\operatorname{vbl}(A_{t(i)})$ it must equal the one violating setting, since it was chosen to be \texttt{Resample}d.
+But $\sigma_{t+1}$ and all the flaws are known, so the whole state vector $Z$ (including the initial state) is determined.
+%FIXME: Include a picture.
 
-It remains to encode $(A_{t(i)})_{i=1}^{t}$ efficiently given $\sigma_{t+1}$.
-Note that we cannot simply encode the events using $\log m$ bits, since this would not give us the bound we want.
-Instead we will use a structure called (by Moser and Tardos) a ``witness forest.''  In this forest, each event called directly by the outer loop of the algorithm (and which actually causes a resample) is a root, and an event called inside $\texttt{Fix}(i)$ is a child of event $A_i$.
-For \ksat, $p = 2^{-k}$, so by the assumption $2^C_0 p d < 1$, each event is independent of all but at most $2^{k-C_0}$ other events.
-So if an event is a child, it can be encoded using at most $2^{k-C_0}$ bits.
-Each event is either a root or not, so the roots can be encoded using $m$ bits.
-Finally, to encode each tree, we can use an extra bit per element to denote the end of the tree.
+It remains to encode $(A_{t(i)})_{i=1}^{t}$ efficiently.
+Note that we cannot simply encode the addressed flaws using the natural encoding that takes $\log m$ bits, since this would lead us a different, much stronger condition than the LLL condition.
+Instead we will use a structure called (by Moser and Tardos) a ``witness forest'' to hold instances in which flaws are addressed.
+In this forest, there are two kinds of nodes:
+\begin{description}
+  \item[Roots:] Whenever the outer loop of the algorithm calls \texttt{Resample}$(i)$ and flaw $A_i$ currently happens, that flaw's variables are resampled, and then its neighbors in the dependency graph are potentially addressed.
+We designate such outer-loop flaw-addressing instances as roots in the witness forest.
+(When \texttt{Fix}$(i)$ is called but flaw $A_i$ does not currently happen, we add nothing to the forest.)
+  \item[Children:] Whenever the inner loop in \texttt{Fix}$(i)$ calls \texttt{Fix}$(j)$, there must already be a node in the tree (either a root or, inductively, a child) corresponding to that flaw-addressing instance for flaw $i$.
+When such a call results in a call to \texttt{Resample}$(j)$, we add a child to that node corresponding to flaw $A_j$.
+\end{description}
 
-So we can encode $Y$ using at most $n + 2m + (k - C_0 + 1)t$ bits, and we can determine $X$ from $Y$, so by Shannon's source coding theorem
+For \ksat, $p = 2^{-k}$, so by the assumption $2^C_0 p (d+1) < 1$, each flaw is independent of all but at most $2^{k-C_0}$ other flaws, including itself.
+So if an flaw-addressing instance is a child in the witness forest, it can be encoded using only $k-C_0$ bits.
+Unfortunately, the argument is more complicated than that, because we must also encode the structure of the forest itself.
+Encoding the identities and order of the roots is easy enough, since each flaw can appear at most once, and the order is by flaw index.
+But encoding the individual trees is harder.
+Given a complete graph on $t$ nodes, there are $t^{t-2}$ distinct spanning trees, so it takes $(t-2) \log_2 t$ bits to encode such trees \cite{prufer1918neuer,neville1953codifying}.
+Since we also need to know the order of calls in the inner loop, and we can have potentially $m$ disjoint trees, the encoding length of a witness forest appears to be even greater than this.
+If this were true, the average encoding length of a node would eventually become larger than $k$, ruining the proof.
+
+However, we actually only need to encode the \emph{unlabelled structure} of each tree in the witness forest, and this is easier.
+There is a simple encoding scheme, but let us add some intuition.
+First, note that recording the order of calls in the inner loop is unnecessary, since the calls are made in a fixed order by flaw identity.
+So if we are told only the identity of the root of each tree in the forest and the set of children of each node, we can reconstruct the order of each child set.
+%FIXME: This is a bad explanation.
+Needs pictures.
+Second, we do not actually need to label the nodes of the tree.
+We can just store the $k-C_0$-bit child identifiers in a some fixed order, for example the order in which a DFS visits nodes of the tree when children are ordered by their indices.
+%FIXME: Also needs a picture.
+So the number of distinct witness trees to encode (ignoring the issue of encoding the roots, which we have already) is at most the number of \emph{unlabelled rooted trees} on $t$ nodes, which is much smaller than the number of labelled trees.
+While there is no known exact formula for counting such trees, Otter \cite{otter1948number} gives an asymptotic estimate of
+\seqn
+  |\{T: T\texttt{ is an unlabelled tree}, |T| = n\}| \tilde O(1) (\frac{1}{0.3382\cdots})^t + O(1) .
+\eeqn
+Therefore (by Shannon's noisy channel coding theorem) there exists an encoding scheme for witness trees that uses only, say, $(\log_2 \frac{1}{0.3383}) t + O(1)$ bits.
+Note that $\log_2 \frac{1}{0.3383}$ is only an upper bound on the number of required bits per iteration of the algorithm, since it assumes the maximum-entropy (uniform) distribution on unlabelled trees of size $n$, while the algorithm may generate witness trees with less entropy than that.
+
+If we are willing to weaken the upper bound a bit, there is a very simple concrete encoding scheme that follows the algorithm execution.
+First, we record the $k-C_0$-bit flaw data in the order in which \texttt{Resample} was called.
+Then we record the history of the depth-first search executed in each tree by logging a $0$ whenever \texttt{Fix} is called inside \texttt{Fix}, and a $1$ whenever a call to \texttt{Fix} returns.
+This uses at most $2$ bits per edge in the witness trees, thus at most $2t$ total bits.
+
+So we can encode $Y$ using at most $n + m + (k - C_0 + 2)t$ bits, and we can determine $Z$ from $Y$, so by Shannon's source coding theorem
 \seqn
@@
 
-This is violated if $t \geq \frac{m}{C_0 - 1}$, so the algorithm must finish (in expectation) before then.
-(In particular, it must finish \emph{eventually}, so there must exist a satisfying solution, and this argument is another proof of the LLL.)
+This is violated if $t \geq \frac{m}{C_0 - 2}$, so the algorithm must finish (in expectation) before that many iterations.
+We have also shown that the algorithm must finish in expectation before $\frac{m}{C_0 - \log_2 \frac{1}{0.3383}} + O(1)$ iterations.
+A more careful argument using a Galton-Watson process gets the constant down to $\log_2 e$.
+%FIXME: Explain this better.
+
+\subsection{General LLL problems}
+%FIXME
+
+\subsection{The simple Moser-Tardos algorithm}
+%FIXME
+
+\subsection{A weaker condition}
+Moser and Tardos actually prove that their algorithm terminates successfully in polynomial time under a weaker condition than $e p (d+1) < 1$.
+Under some assignment of a weight to each flaw $x: \mathcal{A} \to (0,1)$, the following condition must hold for every flaw $A_i$:
+\seqn
+  P(A_i) \leq x(A_i) \prod_{j \in \Gamma(i)} (1-x(A_j)) (\forall A_i \in \mathcal{A}) \label{eqn:general-lll-req}
+\eeqn
+
+This condition is more complicated and unnecessary for exposition, so we have  focused mostly on the simpler one.
+However, it is often the condition cited in the literature, so it is important to see the connection between the two conditions.
+Let us briefly show that $e p (d+1) < 1$ implies the weaker condition with $x(A_i) = \frac{1}{d+1}$ for all $i$.
+$P(A_i) \leq p$ for all $i$ by definition of $p$, so the (claimed) stronger condition gives us:
+\seqn
+  P(A_i) \leq p &< \frac{1}{d+1} \frac{1}{e}  \\
+  &< \frac{1}{d+1} (1 - \frac{1}{d+1})^d \\
+  &\leq \frac{1}{d+1} \prod_{j \in \Gamma(i)} (1-\frac{1}{d+1}) \\
+  &= x(A_i) \prod_{j \in \Gamma(i)} (1 - x(A_j)) .
+\eeqn
 
@@
 Thus far, we have seen only a variable-version LLL problem in which our probability space was a product of $n$ variables.
-The MT algorithm requires explicit access to those variables, and it is restricted to handling every event in the same way -- by resampling all of its variables.
-Both assumptions are fairly restrictive.
-The sample space for some combinatorial problems is better represented in other forms, like permutations \cite{achlioptas2014random}.
+The MT algorithm requires explicit access to those variables, and it is restricted to handling every flaw in the same way -- by resampling all of its variables from the projection of $\mu$ on the lower-dimensional space of those variables.
+Both assumptions are satisfied for many interesting problems, but it is still interesting to see how they can be relaxed.
+The sample space for some combinatorial problems is better represented in other forms, such as permutations \cite{achlioptas2014random}.
 And contra the second assumption, many local search algorithms for \ksat (e.g.
-\texttt{WalkSAT} \cite{papadimitriou1991selecting}) are similar to MT, but will flip only a smaller number of variables on each iteration.
-Achlioptas and Iliopoulos \cite{achlioptas2014random} provide an elegant framework that allows the probability space to be arbitrary and the algorithm to choose actions more flexibly, for which variable-version problems and algorithms are a special case.
-Examining this generalization is instructive.
+\texttt{WalkSAT} \cite{papadimitriou1991selecting}) are similar to MT, but will flip only a smaller number of variables on each iteration, and with a more complicated random component.
+The proof of Moser and Tardos assumes a \emph{particular} search algorithm and applies to a \emph{somewhat broad} class of problems, but it would be nice to have a way to prove efficiency of a broader class of algorithms (and perhaps for an even broader class of problems).
+Achlioptas and Iliopoulos \cite{achlioptas2014random} take a step in this direction, providing an elegant framework for proving efficiency of random local search algorithms in arbitrary finite spaces.
+Examining this framework is instructive.
 
-Again take our $\sigma$-algebra to be $(\Omega, \Sigma)$, but this time without a product structure and without a measure $\mu$.
-Our set of events, which we will now call ``flaws'', are the same as before, but now there is no notion of $\operatorname{vbl}(A_i)$.
+Again take our $\sigma$-algebra to be $(\Omega, \Sigma)$, but this time without a product structure and without even a measure $\mu$.
+Our set of flaws is the same as before, but now there can be no notion of $\operatorname{vbl}(A_i)$.
 Let $U(\sigma)$ denote the set of flaws that happen under state $\sigma$.
-Again, our goal is to find any element $\sigma^* \in \Omega$ such that none of the events happen; we call such elements with $U(\sigma^*) = \emptyset$ ``flawless objects.''  Consider any randomized iterative algorithm whose random state is Markovian and stationary, which attempts to find a flawless object.
+Again, our goal is to find a flawless element $\sigma^* \in \Omega$.
+Consider any randomized iterative algorithm whose random state is Markovian and stationary, which attempts to find a flawless object.
 (The MT algorithm is an example of such an algorithm.)  We can always encode such an algorithm in a random walk graph $\mathcal{T}$ on $\Omega$.
@@
 
-The first requirement is ``accountability of action.''  In Tao's proof, we needed the fact that the flaw (event) $A_t$ addressed on iteration $t$ uniquely determined the values of $\operatorname{vbl}(A_t)$ in the previous state $\sigma_t$.
+The first requirement is ``accountability of action.''  In Tao's proof, we needed the fact that the flaw $A_t$ addressed on iteration $t$ uniquely determined the values of $\operatorname{vbl}(A_t)$ in the previous state $\sigma_t$.
 We proved this by assuming, by induction, that we knew the values of the variables in $\sigma_{t+1}$.
@@
 Assuming atomicity (recall its two requirements: (1) simple constraints and (2) algorithmic focusing) then gave us a way to exactly reconstruct the state of the algorithm before the step was taken.
-Now, let us abstractly write $Z$ for a vector of algorithm states (in the MT algorithm for \ksat, the concatenation of the original uniform assignment bits and the vectors of $k$ bits used at each iteration); $Y$ for the hypothetical observed information that encodes information sufficient to reconstruct the algorithm states (in our explication of Tao's MT proof, this includes the witness forest, the set of initially-violated events, and the final assignment); and $\operatorname{Infer}$ for the deterministic function that maps $Y$ to $Z$.
+Now, let us abstractly write $Z$ for a vector of algorithm states (in the MT algorithm for \ksat, the concatenation of the original uniform assignment bits and the vectors of $k$ bits used at each iteration); $Y$ for the hypothetical observed information that encodes information sufficient to reconstruct the algorithm states (in our explication of Tao's MT proof, this includes the witness forest, the set of flaws of the initial assignment, and the final assignment); and $\operatorname{Infer}$ for the deterministic function that maps $Y$ to $Z$.
 Then the purpose of the atomicity assumption is that $P(Z | Y)$ is a point mass measure at $\operatorname{Infer}(Y)$.
@@
 
-\subsection{Example 2 (Uncountable feasible sets)}
+\subsection{Example 2 (Uncountable feasible set)}
 \label{subsec:uncountable}
-Consider the following feasibility problem:
+Now let us see an example of a problem and algorithm that are covered by the Moser-Tardos framework under some burdensome assumptions.  Consider the following feasibility problem:
 \seqn
@@
   \operatorname{Find} & x \in \Reals^n \texttt{s.t.} \\
-  x \in \mathcal{B}(1, n, ||\cdot||_\infty) \\
   A_i(x) = 0 \forall i \in [m] ,
 \eeqn
-where the $A_i$ are \emph{arbitrary} indicator functions $\Reals^{m} \to \{0,1\}$ and $\mathcal{B}(1, n, ||\cdot||_\infty)$ is the unit $\infty$-norm ball in $\Reals^n$ (a.k.a.
-the solid hypercube).
+where the $A_i$ are indicator functions $\Reals^{m} \to \{0,1\}$.
+
+We can write down an MT algorithm to solve this problem.
+This will add some assumptions, two of which we must discuss before even writing down the algorithm.
+First, we assume that each function depends only on a subset of the elements of the vector $x$; let $\operatorname{vbl}(f) \subseteq [n]$ denote the indices on which a function $f$ depends, and let $y[\operatorname{vbl}(f)]$ denote the corresponding elements of a vector $y$.
+Second, we must write down a \emph{product measure} $\mu = \prod_{j=1}^{n} \mu_j$ on $\Reals^n$, where each $\mu_j$ is a measure on $\Reals$.
+During the algorithm, we must be able to sample from $\mu$ and from product measures of the form $\prod_{j \in \operatorname{vbl}(A_i)} \mu_j$.
+Later we will have to impose additional constraints on $\mu$, but for now imagine that $\mu_j = N(0, 1)$ for each $j$.
+
+\begin{algorithm}[H]
+\caption{An MT-like algorithm for problem \ref{prob:feas}.}
+\label{alg:mt-feas}
+\begin{algorithmic}[1]
+
+\State Initialize $x \hasDist \mu$.
+\ForAll{$i \in [m]$}
+  \Call{Fix}{i}
+\EndFor
+
+\Function{Fix}{i}
+  \If{$A_i(x) = 0$}
+    \Return
+  \EndIf
+  \State Resample $x[\operatorname{vbl}(A_i)]$ from $\prod_{j \in \operatorname{vbl}(A_i)}$.
+  \ForAll{$k \in [m]: k \in \Gamma(i), A_k(x) = 1$}
+    \Call{Fix}{$k$}
+  \EndFor
+\EndFunction
+
+\end{algorithmic}
+\end{algorithm}
+
+Moser and Tardos prove that this algorithm finishes in polynomial time under the following conditions:
+
+\begin{enumerate}
+  \item Each $A_i$ is $\mu$-measurable and satisfies $\mu(\{x: A_i(x) = 1\}) \leq p$ for some small $p$.
+That is, random sampling from our measure $\mu$ satisfies each single constraint with high probability.
+(The difficulty of the problem comes from the requirement of satisfying the constraints \emph{jointly}.)  Otherwise the constraints can be arbitrarily complicated.
+\label{eqn:feas-margprob-condition}
+  \item The dependency graph induced by the constraints and the measure $\mu$ meets the usual LLL condition: $e p (d+1) < 1$. \label{eqn:feas-lowdep-condition}
+\end{enumerate}
+
+%FIXME
+
+, but their proof relies on the finiteness of $n$.
+The proof of Achlioptas and Iliopoulos would need no assumptions on $n$, but it requires that $m$ be finite and fails when the sample space $\Omega$ is infinite (even countably so), as in this case.
+The Moser-Tardos assumptions are:
+
+\begin{enumerate}
+%FIXME: Some of these comments are useful to have here.
+  \item We chose our sample space to be a norm ball in finite dimensions, rather than say all of $\Reals^n$, precisely so that our sample space would be compact, allowing us to choose a simple probability measure as our dominating measure when computing entropies.
+In general, to bound the number of steps taking by the algorithm, we need a probability measure that $Y$-almost surely dominates $P(Z | Y)$, for which the LLL-required marginal flaw probabilities hold, and which induces low dependency degree among the flaws.
+  \item We chose the infinity norm ball in particular for a different reason: the sample space needs to be an axis-aligned rectangle, or it is not easy to ensure that the resampling step keeps us inside it while modifying only $k$ variables at a time and covering a large volume.
+Again, this is not important for proving something about the number of steps taken by the algorithm, but it is important for the efficiency of the individual steps.
+  \item Rather than augment $Y$ to pin down $Z$, we could imagine proving directly that $H(Z | Y)$ is small.
+It is natural to view $Z$ and $Y$ as the states and observations of a hidden Markov model, respectively, and use any of the well-known algorithms for inference in HMMs to compute $P(Z | Y)$.
+(Such inference could be quite difficult if the constraint functions are complicated, but for purposes of analysis we could imagine performing it.)  At least in this case, augmenting $Y$ was much simpler, but this makes the example weaker, since we are not taking advantage of our freedom in making $H(Z | Y) > 0$.
+  \item With respect to the constraint functions themselves, the algorithm can be made efficient as long as we have an oracle that can efficiently find a violated constraint when there is one.
+It may be possible to have an efficient version of the algorithm testing only a core of $\operatorname{poly}(n)$ ``core'' constraints using the ideas of Haeupler et al \cite{haeupler2011new}.
+\end{enumerate}
+
+In our next example, we will remove the finite-dimensionality restriction.
+
+\subsection{Example 3 (Uncountable, infinite-dimensional feasible set)}
+
+\label{subsec:uncountable}
+Now consider the following modified feasibility problem:
+\seqn
+  \label{prob:feas}
+  \operatorname{Find} & x \in L_2(\mu) \texttt{s.t.} \\ %FIXME: L_2?
+  A_i(x) = 0 \forall i \in [m] ,
+\eeqn
+where the $A_i$ are \emph{arbitrary} indicator functions $\Reals^{m} \to \{0,1\}$ and $\mu$ is a probability measure on a probability space $(\Omega, \Sigma)$.
 We will assume later that each function depends only on a subset of the elements of the vector $x$; let $\operatorname{vbl}(z) \subseteq [n]$ denote the indices on which a function depends, and let $y[\operatorname{vbl}(z)]$ denote the corresponding elements of a vector $y$.
@@
 
-Moser and Tardos prove that this algorithm finishes in polynomial time under the usual LLL conditions, but their proof relies on the finiteness of $n$ and $m$.
-The proof of Achlioptas and Iliopoulos would need no assumptions on $n$ and weaker assumptions on $m$, but it fails when the sample space $\Omega$ is infinite (even countably so).
-We will provide a proof that does not depend on the dimension that this algorithm finishes in expected polynomial time under conditions analogous to the LLL conditions.
-Specifically, we will make the following assumptions:
+We will provide a proof that this algorithm finishes in expected polynomial time under conditions analogous to the LLL conditions.  Specifically, we will make the following assumptions:
 \begin{enumerate}
@@
   \item We chose our sample space to be a norm ball in finite dimensions, rather than say all of $\Reals^n$, precisely so that our sample space would be compact, allowing us to choose a simple probability measure as our dominating measure when computing entropies.
-In general, to bound the number of steps taking by the algorithm, we need a probability measure that $Y$-almost surely dominates $P(Z | Y)$, for which the LLL-required marginal event probabilities hold, and which induces low dependency degree among the events.
+In general, to bound the number of steps taking by the algorithm, we need a probability measure that $Y$-almost surely dominates $P(Z | Y)$, for which the LLL-required marginal flaw probabilities hold, and which induces low dependency degree among the flaws.
   \item We chose the infinity norm ball in particular for a different reason: the sample space needs to be an axis-aligned rectangle, or it is not easy to ensure that the resampling step keeps us inside it while modifying only $k$ variables at a time and covering a large volume.
@@
 Moser and Tardos \cite{moser2010constructive} used a similar (but much simplified) idea in their original paper.
-It is easy to see that the simple iterative version of their algorithm (algorithm \ref{alg:mt}) can be parallelized by resampling the variables of independent sets of events, though it is not trivial that this ensures the algorithm terminates in $O(\log m)$ such steps (in expectation).
+It is easy to see that the simple iterative version of their algorithm (algorithm \ref{alg:mt}) can be parallelized by resampling the variables of independent sets of flaws, though it is not trivial that this ensures the algorithm terminates in $O(\log m)$ such steps (in expectation).
 More recently, Chung et al \cite{chung2014distributed} gave more practical parallel algorithms, which trade off the size of the independent set for the time taken to find such a set.

### 2015-05-29 paper/nonatomic/paper.tex
Further work on the paper.

@@
   howpublished = {\url{http://blog.computationalcomplexity.org/2009/06/kolmogorov-complexity-proof-of-lov.html}}
+}
+@article{Lu2007,
+  author = {Lu, Linyuan, Székely, László},
+  journal = {The Electronic Journal of Combinatorics [electronic only]},
+  keywords = {dependency graph; negative dependency graph; Lovasz local lemma; random injections; permutation enumeration; hzpergraph packing; Turan-type extremal problems},
+  language = {eng},
+  number = {1},
+  pages = {Research Paper R63, 13 p.-Research Paper R63, 13 p.},
+  publisher = {Prof. André Kündgen, Deptartment of Mathematics, California State University San Marcos, San Marcos},
+  title = {Using Lovász local lemma in the space of random injections.},
+  url = {http://eudml.org/doc/117186},
+  volume = {14},
+  year = {2007},
 }
@@
 
-For \ksat, $p = 2^{-k}$, so by the assumption $2^C_0 p (d+1) < 1$, each flaw is independent of all but at most $2^{k-C_0}$ other flaws, including itself.
+For \ksat, $p = 2^{-k}$, since only one of the $2^k$ assignments to a disjunctive clause can violate it.
+The assumption $2^C_0 p (d+1) < 1$ means that each flaw is independent of all but at most $2^{k-C_0}$ flaws, including itself.
 So if an flaw-addressing instance is a child in the witness forest, it can be encoded using only $k-C_0$ bits.
-Unfortunately, the argument is more complicated than that, because we must also encode the structure of the forest itself.
+This is the crux of the argument.
+Unfortunately, things are slightly more complicated than that, because we must also encode the structure of the forest itself.
 Encoding the identities and order of the roots is easy enough, since each flaw can appear at most once, and the order is by flaw index.
@@
 \seqn
-  n + m + (k - C_0 + 1)t \leq n + k t .
+  n + m + (k - C_0 + 2)t \leq n + k t \label{eqn:ksat-entropy-inequality}.
 \eeqn
@@
 We have also shown that the algorithm must finish in expectation before $\frac{m}{C_0 - \log_2 \frac{1}{0.3383}} + O(1)$ iterations.
-A more careful argument using a Galton-Watson process gets the constant down to $\log_2 e$.
+Moser and Tardos give a more careful argument considering the witness tree explicitly as a draw from a particular Galton-Watson process, bringing $C_0$ down to $\log_2 e$.
+(This implies that the witness trees generated by algorithm \ref{alg:mt-recursive} are not uniformly distributed on the possible trees of fixed size.)
+
+\subsection{Arbitrary starting state}
+Note also that a similar proof would still work, with a worse bound, if we eliminated the requirement that the algorithm started in a uniform random state.
+If the starting state is arbitrary, then it may have entropy $0$, and inequality \ref{eqn:ksat-entropy-inequality} would become
+\seqn
+  n + m + (k - C_0 + 2)t &\leq k t \\
+  \implies t \leq \frac{n+m}{C_0-2} .
+\eeqn
+So we pay an additive penalty (at least according to this upper bound) equal to the dimension of the problem.  We will see this later when we review \cite{achlioptas2014random}.
 %FIXME: Explain this better.
@@
 \subsection{General LLL problems}
-%FIXME
+The proof above was restricted to \ksat.  In this subsection we outline the proof of the constructive LLL in full generality.
+
+The important difference between \ksat and general LLL problems is that \ksat events are \emph{elementary}: For every flaw $A_i$, only a single assignment to the variables in $\operatorname{vbl}(A_i)$ causes $A_i$ to happen.
+This allowed us to reconstruct the algorithm state vector $Y$ given only the final state $\sigma_{t}$ and the vector of flaws addressed $(A_1, \cdots, A_t)$.
+Each flaw could be addressed using $\log_2 (d+1) + C_0$ bits, but from its identity we reconstructed one out of $2^k$ possible assignments, $k = \log_2 2^k = \log_2 (1/p)$ bits of information.
+The contradiction arose because $\log_2 (1/p) > \log_2 (d+1) + C_0$, which is exactly the LLL condition.
+
+Now that we have stated the argument in this way, the general case is straightforward.
+Assume that we know $\sigma_{i+1}$ and $A_i$.
+Then we know that $\operatorname{vbl}(A_i)$ take values in $\sigma_i$ with joint probability at most $p$.
+If $\Omega$ is discrete, then we have gained $\log_2 (1/p)$ bits of information about $\sigma_i$.
+If we used only $\log_2 (d+1) + C_0$ bits to encode it, we eventually get the same contradiction.
+
+If $\Omega$ is continuous, then the same is true, except that we must make our arguments in terms of entropy rather than bits.
+We have also ignored the fact that our argument was inductive, and when there is some uncertainty remaining in $\sigma_{i}$, we no longer know it exactly when we try to infer $\sigma_{i-1}$.
+Both issues are surmountable, but we will defer that to section \ref{subsec:trivial}.
 
 \subsection{The simple Moser-Tardos algorithm}
+Algorithm \ref{alg:mt-recursive} is convenient for analysis, but algorithm \ref{alg:mt} is easier to describe and implement, and it leads to parallel algorithms.
+In this subsection we outline the proof of the constructive LLL with algorithm \ref{alg:mt}.
 %FIXME
@@
 
+The proof of the constructive LLL under condition \ref{eqn:general-lll-req} uses the condition to construct a Galton-Watson process whose entropy lower-bounds that of the witness trees actually generated by algorithm \ref{}.
+Moser and Tardos do not use those terms explicitly.
+Instead, they bound the probability of any tree appearing with a particular root and sum those probabilities over trees and roots to bound the expected number of elements of a witness tree.
+It would be interesting to give a version of the proof that more explicitly relies on the entropy compression argument, but we will not do that here.
+%FIXME
+
 \section{The algorithmic LLL on general sample spaces}
@@
 The MT algorithm requires explicit access to those variables, and it is restricted to handling every flaw in the same way -- by resampling all of its variables from the projection of $\mu$ on the lower-dimensional space of those variables.
-Both assumptions are satisfied for many interesting problems, but it is still interesting to see how they can be relaxed.
-The sample space for some combinatorial problems is better represented in other forms, such as permutations \cite{achlioptas2014random}.
+Both assumptions are satisfied for many interesting problems, but it is still useful to see how they can be relaxed.
+The sample space for some combinatorial problems is better represented in other forms, such as permutations \cite{achlioptas2014random} or more generally random injections \cite{lu2007injections}.
 And contra the second assumption, many local search algorithms for \ksat (e.g.
 \texttt{WalkSAT} \cite{papadimitriou1991selecting}) are similar to MT, but will flip only a smaller number of variables on each iteration, and with a more complicated random component.
+%FIXME: These are weak examples.
 The proof of Moser and Tardos assumes a \emph{particular} search algorithm and applies to a \emph{somewhat broad} class of problems, but it would be nice to have a way to prove efficiency of a broader class of algorithms (and perhaps for an even broader class of problems).
@@
 Consider any randomized iterative algorithm whose random state is Markovian and stationary, which attempts to find a flawless object.
-(The MT algorithm is an example of such an algorithm.)  We can always encode such an algorithm in a random walk graph $\mathcal{T}$ on $\Omega$.
+(The MT algorithm is an example of such an algorithm.)  By definition, we can view such algorithms as random walks $\mathcal{T}$ on $\Omega$.
+
+The argument in section \ref{sec:lll} was really an argument about %FIXME
 Now let us try to write down the properties of this graph that are really needed for something like the proof in section \ref{sec:lll} to apply.
 
+%FIXME: s/_t/_s/
 The first requirement is ``accountability of action.''  In Tao's proof, we needed the fact that the flaw $A_t$ addressed on iteration $t$ uniquely determined the values of $\operatorname{vbl}(A_t)$ in the previous state $\sigma_t$.

### 2015-06-03 paper/nonatomic/paper.tex
Further revisions to initial draft of paper.

@@
-\documentclass[twocolumn]{article}
+\documentclass{article}
 \usepackage{amsmath, amssymb}
@@
 \usepackage{hyperref}
+\usepackage{tikz}
+\usetikzlibrary{fit,positioning}
+\usepackage{mdframed}
 \author{Henry Milner}
@@
 
+\DeclareMathOperator*{\esssup}{ess\,sup}
+
+\DeclareMathOperator*{\essinf}{ess\,inf}
+
 \newcommand{\inprod}[2]%
@@
   {\mathbb{R}}
+\newcommand{\Nats}%
+  {\mathbb{N}}
 
@@
 }
-@article{Lu2007,
+@article{lu2007injections,
   author = {Lu, Linyuan, Székely, László},
@@
 
+\begin{figure}[ht]
+  \begin{mdframed}
+    \centering
+    \includegraphics[scale=0.5]{figures/mt-directed-example.png}
+    \caption{One iteration of an example run of the Moser-Tardos algorithm for \ksat.
+The problem is to find an assignment to the binary variables $a$, $b$, $c$, and $d$ so that $(!a|b)\&(b|!d)\&(a|c)$.
+The variables are initialized uniformly at random before the algorithm starts.}
+    \label{fig:mt-directed-example}
+  \end{mdframed}
+\end{figure}
+
+\begin{figure}[ht]
+  \begin{mdframed}
+    \centering
+    \includegraphics[scale=0.5]{figures/mt-undirected-example.png}
+    \caption{The dependency graph for the problem pictured in figure \ref{fig:mt-directed-example}.  Here $d = 2$ and $p = 2^{-2}$, so the LLL condition $e p (d + 1)$ does not hold.}
+    \label{fig:mt-undirected-example}
+  \end{mdframed}
+\end{figure}
+
 Their proof uses a novel \emph{entropy compression} argument.
@@
 Achlioptas and Iliopoulos give a general condition on \emph{algorithms} rather than problems, and bring us partway to a general constructive LLL, but they require that the space $\Omega$ is finite, and this is the restriction we will remove.
-In section \ref{sec:examples} we give a series of examples suggesting the entropy compression argument could be fruitfully extended to novel LLL-like applications in which the state vector of an algorithm cannot be uniquely recovered.
+In section \ref{sec:nonatomic-intro} we suggest that the entropy compression argument could be fruitfully extended to novel LLL-like applications in which the state vector of an algorithm cannot be uniquely recovered.
 Achlioptas and Iliopoulos call such algorithms (or, technically, their associated random walk graphs) ``non-atomic''.
 By removing the atomicity assumption, we can develop a more general constructive LLL.
-In section \ref{sec:nonatomic}, we give a formal proof.
-%FIXME: Language.
+In section \ref{sec:examples} we give a series of motivating examples.
+In section \ref{sec:markov}, we discuss an interpretation of the entropy compression argument as a statement about Markov chains and associated hidden Markov chains.
+Finally, in section \ref{sec:nonatomic}, we give formal proofs.
+%FIXME: Not clear this is a good organization.
 
@@
 \subsection{Arbitrary starting state}
+\label{subsec:arbitrary}
 Note also that a similar proof would still work, with a worse bound, if we eliminated the requirement that the algorithm started in a uniform random state.
@@
 \eeqn
-So we pay an additive penalty (at least according to this upper bound) equal to the dimension of the problem.  We will see this later when we review \cite{achlioptas2014random}.
+So we pay an additive penalty (at least according to this upper bound) equal to the dimension of the problem.
+More generally, we pay a cost of $\log_2 |\Omega|$ when $\Omega$ is finite.
+We will see this later when we review \cite{achlioptas2014random}.
 %FIXME: Explain this better.
+%FIXME: Forward-reference the section where we discuss this.
 
 \subsection{General LLL problems}
+\label{subsec:general-lll}
 The proof above was restricted to \ksat.  In this subsection we outline the proof of the constructive LLL in full generality.
@@
 
-The proof of the constructive LLL under condition \ref{eqn:general-lll-req} uses the condition to construct a Galton-Watson process whose entropy lower-bounds that of the witness trees actually generated by algorithm \ref{}.
-Moser and Tardos do not use those terms explicitly.
+The proof of the constructive LLL under condition \ref{eqn:general-lll-req} uses the condition to construct a Galton-Watson process whose entropy lower-bounds that of the witness trees actually generated by algorithm \ref{alg:mt-recursive}.
+Moser and Tardos do not use the word ``entropy'' explicitly anywhere in their paper.
 Instead, they bound the probability of any tree appearing with a particular root and sum those probabilities over trees and roots to bound the expected number of elements of a witness tree.
@@
 Consider any randomized iterative algorithm whose random state is Markovian and stationary, which attempts to find a flawless object.
-(The MT algorithm is an example of such an algorithm.)  By definition, we can view such algorithms as random walks $\mathcal{T}$ on $\Omega$.
+(The MT algorithm is an example of such an algorithm.)  By definition, we can view any such algorithm as a random walk $\mathcal{T}$ on a graph whose nodes are $\Omega$ and whose edge-weights define a transition matrix for the walk.
 
-The argument in section \ref{sec:lll} was really an argument about %FIXME
-Now let us try to write down the properties of this graph that are really needed for something like the proof in section \ref{sec:lll} to apply.
+%The argument in section \ref{sec:lll} was really an argument about %FIXME
+Now let us try to write down the properties of these graphs that are really needed for something like the proof in section \ref{sec:lll} to apply.
 
-%FIXME: s/_t/_s/
-The first requirement is ``accountability of action.''  In Tao's proof, we needed the fact that the flaw $A_t$ addressed on iteration $t$ uniquely determined the values of $\operatorname{vbl}(A_t)$ in the previous state $\sigma_t$.
-We proved this by assuming, by induction, that we knew the values of the variables in $\sigma_{t+1}$.
-Given the identity of the flaw, we knew that $\sigma_t$ and $\sigma_{t+1}$ differed only on the variables in $\operatorname{vbl}(A_t)$ differed.
-Further, since each flaw could be violated by only one configuration of its variables, we knew what that configuration must have been in $\sigma_t$.
+The first requirement is ``atomicity,'' which generalizes Moser's and Tardos' concept of ``elementary'' flaws in section \ref{subsec:general-lll}.
+We needed the fact that the flaw $A_s$ addressed on iteration $s$ uniquely determined the values of $\operatorname{vbl}(A_s)$ in the previous state $\sigma_s$.
+We proved this by assuming, by induction, that we knew the values of the variables in $\sigma_{s+1}$.
+Given the identity of the flaw, we knew that $\sigma_s$ and $\sigma_{s+1}$ differed only on the variables in $\operatorname{vbl}(A_s)$ differed.
+Further, since each flaw could be violated by only one configuration of its variables, we knew what that configuration must have been in $\sigma_s$.
 Without variables, we need a different tack to get a correspondence between flaw vectors and algorithm state vectors.
 We will still assume that the algorithm ``addresses'' a single flaw on each iteration; given a current state $\sigma_t$ and a flaw to address, the algorithm draws from some distribution to find its next action.
-To achieve accountability, we assume that, for every (flaw, state) pair $(A_{t}, \sigma_{t+1})$, there is a unique previous state $\sigma_t$.
+(The transition probabilities for each node are obtained by marginalizing out the flaw selection distribution.)  To achieve accountability, we assume that, for every (flaw, state) pair $(A_{s}, \sigma_{s+1})$, there is a unique previous state $\sigma_s$.
 In terms of the random walk graph, this means that if we label each weighted edge in the graph with the flaw we addressed at that step, each node has at most one incoming edge for each flaw.
-This requirement is called ``atomicity.''
+This property is called atomicity.
 
@@
 This condition is unimportant, since we can always modify the algorithm (for purposes of proofs) to make it so.
-We will discuss this further in section \ref{sec:nonatomic}.
-  \item[Algorithmic focusing:] After choosing a flaw $A_t$, the algorithm modifies \emph{only} the variables in $\operatorname{vbl}(A_t)$ on iteration $t$.  This condition is nontrivial.
+We will discuss this further in section \ref{subsec:trivial}.
+  \item[Algorithmic focusing:] After choosing a flaw $A_s$, the algorithm modifies \emph{only} the variables in $\operatorname{vbl}(A_s)$ on iteration $s$.  This condition is nontrivial.
 \end{description}
@@
 In section \ref{sec:lll}, we saw that it was desirable (from the perspective of the entropy compression argument) to maximize the entropy of each step.
-So we will assume that, having chosen a flaw $A_t$ to address, our algorithm chooses uniformly at random from all the possible next states it could reach.
-We call this ``action set'' $A(A_t, \sigma_t)$.
-(In the variable version, this would have been of size $2^{\operatorname{vbl}(A_t)}$ for any $\sigma_t$.)  Then, the requirement is that no flaw $A_i$ has too many neighbors with small action sets; specifically,
+So we will assume that, having chosen a flaw $A_s$ to address, our algorithm chooses uniformly at random from all the possible next states it could reach.
+We call this ``action set'' $A(A_s, \sigma_s)$.
+(In the variable version with binary variables, this would have been of size $2^{\operatorname{vbl}(A_s)}$ for any $\sigma_s$.)  Then, the requirement is that no flaw $A_i$ has too many neighbors with small action sets; specifically,
 \seqn
@@
 for some positive $\delta$.
+%FIXME: Rephrase the previous results in terms of \delta; this language is clearer.
+%FIXME: Note that A-I have the uniformity assumption, which we could relax later.
 
 Under these two requirements, Achlioptas and Iliopoulos show that any such algorithm will terminate quickly.
-In that case, the running time of the random walk algorithm is $\delta^{-1} (\log_2 |\Omega| + |U(\sigma_1)| + s)$ steps with probability at least $1 - 2^{-s}$.
-
-Interestingly, instead of using a measure $\mu$ on $(\Omega, \Sigma)$, requirement \ref{eqn:degree-req} is entirely in terms of the random walk graph, and it is independent of the starting location $\sigma_1$.
-The proof does not require any entropy in the starting state.
-Still, the main point of our interest in \cite{achlioptas2014random} is the clear statement of the atomicity requirement.
+The running time of the random walk algorithm is $\delta^{-1} (\log_2 |\Omega| + |U(\sigma_1)| + s)$ steps with probability at least $1 - 2^{-s}$.
+Though this is a high-probability bound rather than a bound in expectation, it is similar to the running time we found in section \ref{sec:lll}.
+Here $|U(\sigma_1)|$, the number of flaws in the initial assignment, stands in for $m$, which we used to upper-bound the same quantity in section \ref{sec:lll}.
+The $\log_2 |\Omega|$ term is introduced because Achlioptas and Iliopoulos assume an arbitrary starting location; we found the same additive constant in section \ref{subsec:arbitrary}.
+And $\delta^{-1}$ is a slack term analogous to $C_0 - 2$.
+%FIXME: When we rephrase the above stuff in terms of \delta there will be a direct correspondence.
+The main point of our interest in \cite{achlioptas2014random} is the clear statement of the atomicity requirement, and we will not revise their proof.
+Note that, by working directly with the random walk graph and paying the extra $\log_2 |\Omega|$ cost in running time, Achlioptas and Iliopoulos avoid assuming a measure $\mu$ on $(\Omega, \Sigma)$.
+We will have to reintroduce the measure when we attempt to generalize their result to infinite spaces.
+This seems to be a very mild restriction.
 
 \section{Non-atomic algorithms}
-\label{sec:nonatomic}
+\label{sec:nonatomic-intro}
 In sections \ref{sec:lll} and \ref{sec:perfect}, we saw that the \emph{atomicity} of an \mt algorithm was a key assumption in the entropy-compression argument.
-In this section, we conjecture that similar arguments could be made without the atomicity assumption.
-It would be very interesting if this conjecture held, since it could open up the algorithmic LLL to applications outside combinatorial search.
-We look forward to following up on this conjecture in future work, but for now we will just sketch some arguments.
-The work in this section is, to our knowledge, original.
+In this section, we show that similar arguments can be made without the atomicity assumption.
+This raises the possibility of using the algorithmic LLL in applications outside combinatorial search.
 
@@
 Now, let us abstractly write $Z$ for a vector of algorithm states (in the MT algorithm for \ksat, the concatenation of the original uniform assignment bits and the vectors of $k$ bits used at each iteration); $Y$ for the hypothetical observed information that encodes information sufficient to reconstruct the algorithm states (in our explication of Tao's MT proof, this includes the witness forest, the set of flaws of the initial assignment, and the final assignment); and $\operatorname{Infer}$ for the deterministic function that maps $Y$ to $Z$.
-Then the purpose of the atomicity assumption is that $P(Z | Y)$ is a point mass measure at $\operatorname{Infer}(Y)$.
+Write $P(Z | Y)$ for the posterior distribution (after Bayesian updating) on the true algorithm state vector given the alternative information.
+The purpose of the atomicity assumption is that $P(Z | Y)$ is a point mass measure at $\operatorname{Infer}(Y)$.
 
@@
 \eeqn
-This is exactly the fact that we needed in Tao's proof to show a contradiction when $t$ was too large.
+This is exactly the fact that we needed in section \ref{sec:lll} to show a contradiction when $t$ was too large.
 
-From this perspective, it does not seem strictly necessary that $H(Z | Y) = 0$.
-Perhaps in some cases we can use inequality \ref{eqn:entropy-encoded} or equation \ref{eqn:entropy-general} with some other upper bounds on $H(Z | Y)$ or $H(Z)$ to make an argument like Tao's.
-We will consider two examples to illustrate this point.
+From this perspective, it does not seem strictly necessary that $H(Z | Y) = 0$ or that $Y$ and $Z$ be representable by finitely many bits.
+Perhaps in some cases we can use inequality \ref{eqn:entropy-encoded} or equation \ref{eqn:entropy-general} with some other upper bounds on $H(Z | Y)$ or $H(Z)$ to make an entropy compression argument.
+We will consider several examples to illustrate this point.
 
+\section{Examples}
+\label{sec:examples}
 \subsection{Example 1 (trivial)}
@@
 It only violates the first condition for atomicity in the framework of Achlioptas and Iliopoulos, which is that if we observe state $\sigma_{t+1}$ after a flaw $f$ is addressed, there is only one possibility for the previous state $\sigma_t$.
-They call this condition ``syntactic,'' since it can always be addressed by expanding the flaw set to make flaws more granular.
+They call this condition ``syntactic,'' since it can always be addressed (when $\Omega$ is finite) by expanding the flaw set to make flaws more granular.
 For example, say that we have a modified version of \ksat in which $2^c$ out of $2^k$ assignments to each clause can make it fail.
@@
 So the argument in our trivial example does not buy us any extra power in handling non-atomicity, only an alternative proof of Achlioptas' and Iliopoulos' comment.
+And in the framework of Moser and Tardos, it buys us nothing, since they do not require atomicity, but rather a bound $p$ on the marginal probability of flaws.
 
@@
 First, we assume that each function depends only on a subset of the elements of the vector $x$; let $\operatorname{vbl}(f) \subseteq [n]$ denote the indices on which a function $f$ depends, and let $y[\operatorname{vbl}(f)]$ denote the corresponding elements of a vector $y$.
-Second, we must write down a \emph{product measure} $\mu = \prod_{j=1}^{n} \mu_j$ on $\Reals^n$, where each $\mu_j$ is a measure on $\Reals$.
+Second, we must write down a \emph{measure} $\mu = \prod_{j=1}^{n} \mu_j$ on $\Reals^n$, where each $\mu_j$ is a measure on $\Reals$.
 During the algorithm, we must be able to sample from $\mu$ and from product measures of the form $\prod_{j \in \operatorname{vbl}(A_i)} \mu_j$.
@@
   \EndIf
-  \State Resample $x[\operatorname{vbl}(A_i)]$ from $\prod_{j \in \operatorname{vbl}(A_i)}$.
+  \State Resample $x[\operatorname{vbl}(A_i)]$ from $\prod_{j \in \operatorname{vbl}(A_i)} \mu_j$.
   \ForAll{$k \in [m]: k \in \Gamma(i), A_k(x) = 1$}
@@
 \begin{enumerate}
+  \item $n$ is finite. \label{eqn:feas-finitedim-condition}
   \item Each $A_i$ is $\mu$-measurable and satisfies $\mu(\{x: A_i(x) = 1\}) \leq p$ for some small $p$.
@@
 \label{eqn:feas-margprob-condition}
+  \item $\mu$ is a product measure $\prod_{j=1}^{n} \mu_j$, where each $\mu_j$ is a measure on $\Reals$.
+During the algorithm, we must be able to sample from $\mu$ and from product measures of the form $\prod_{j \in \operatorname{vbl}(A_i)} \mu_j$.
+\label{eqn:feas-prodmeasure-condition}
   \item The dependency graph induced by the constraints and the measure $\mu$ meets the usual LLL condition: $e p (d+1) < 1$. \label{eqn:feas-lowdep-condition}
@@
 
-%FIXME
-
-, but their proof relies on the finiteness of $n$.
-The proof of Achlioptas and Iliopoulos would need no assumptions on $n$, but it requires that $m$ be finite and fails when the sample space $\Omega$ is infinite (even countably so), as in this case.
-The Moser-Tardos assumptions are:
-
-\begin{enumerate}
-%FIXME: Some of these comments are useful to have here.
-  \item We chose our sample space to be a norm ball in finite dimensions, rather than say all of $\Reals^n$, precisely so that our sample space would be compact, allowing us to choose a simple probability measure as our dominating measure when computing entropies.
+% \begin{enumerate}
+% %FIXME: Some of these comments are useful to have here.
+%   \item We chose our sample space to be a norm ball in finite dimensions, rather than say all of $\Reals^n$, precisely so that our sample space would be compact, allowing us to choose a simple probability measure as our dominating measure when computing entropies.
 In general, to bound the number of steps taking by the algorithm, we need a probability measure that $Y$-almost surely dominates $P(Z | Y)$, for which the LLL-required marginal flaw probabilities hold, and which induces low dependency degree among the flaws.
-  \item We chose the infinity norm ball in particular for a different reason: the sample space needs to be an axis-aligned rectangle, or it is not easy to ensure that the resampling step keeps us inside it while modifying only $k$ variables at a time and covering a large volume.
+%   \item We chose the infinity norm ball in particular for a different reason: the sample space needs to be an axis-aligned rectangle, or it is not easy to ensure that the resampling step keeps us inside it while modifying only $k$ variables at a time and covering a large volume.
 Again, this is not important for proving something about the number of steps taken by the algorithm, but it is important for the efficiency of the individual steps.
-  \item Rather than augment $Y$ to pin down $Z$, we could imagine proving directly that $H(Z | Y)$ is small.
+%   \item Rather than augment $Y$ to pin down $Z$, we could imagine proving directly that $H(Z | Y)$ is small.
 It is natural to view $Z$ and $Y$ as the states and observations of a hidden Markov model, respectively, and use any of the well-known algorithms for inference in HMMs to compute $P(Z | Y)$.
 (Such inference could be quite difficult if the constraint functions are complicated, but for purposes of analysis we could imagine performing it.)  At least in this case, augmenting $Y$ was much simpler, but this makes the example weaker, since we are not taking advantage of our freedom in making $H(Z | Y) > 0$.
-  \item With respect to the constraint functions themselves, the algorithm can be made efficient as long as we have an oracle that can efficiently find a violated constraint when there is one.
+%   \item With respect to the constraint functions themselves, the algorithm can be made efficient as long as we have an oracle that can efficiently find a violated constraint when there is one.
 It may be possible to have an efficient version of the algorithm testing only a core of $\operatorname{poly}(n)$ ``core'' constraints using the ideas of Haeupler et al \cite{haeupler2011new}.
-\end{enumerate}
+% \end{enumerate}
 
@@
 
-\subsection{Example 3 (Uncountable, infinite-dimensional feasible set)}
+\subsection{Example 3 (Infinite-dimensional feasible set)}
 
-\label{subsec:uncountable}
+\label{subsec:infinitedim}
 Now consider the following modified feasibility problem:
 \seqn
-  \label{prob:feas}
-  \operatorname{Find} & x \in L_2(\mu) \texttt{s.t.} \\ %FIXME: L_2?
-  A_i(x) = 0 \forall i \in [m] ,
+  \label{prob:feas-infinitedim}
+  \operatorname{Find} & x \in \Omega \\
+  A_i(x) = 0 \forall A_i \in \mathcal{A} .
 \eeqn
-where the $A_i$ are \emph{arbitrary} indicator functions $\Reals^{m} \to \{0,1\}$ and $\mu$ is a probability measure on a probability space $(\Omega, \Sigma)$.
-We will assume later that each function depends only on a subset of the elements of the vector $x$; let $\operatorname{vbl}(z) \subseteq [n]$ denote the indices on which a function depends, and let $y[\operatorname{vbl}(z)]$ denote the corresponding elements of a vector $y$.
-Finally, let $\mu^d$ denote the uniform measure on $\mathcal{B}(1, d, ||\cdot||_\infty)$.
+Again the $A_i$ are \emph{arbitrary} indicator functionals $\Omega \to \{0,1\}$, though now there may be infinitely many of them.
+We assume $\Omega$ is a Hilbert space so that it admits a countable basis, and we endow it with a probability measure $\mu$.
+(For example, we could be optimizing over functions on some underlying space, and $\mu$ could be a Gaussian process measure.)  Let $(\phi_j)_{j \in \Nats} \in \Omega^{\Nats}$ be a countable basis for $\Omega$, and let $\phi(x) = (\inprod{\phi_j}{x} \phi_j)_{j \in \Nats} \in \Omega^{\Nats}$ be the basis representation of an element $x \in \Omega$.
+Again, we will assume that each constraint functional depends only on a subset of the elements of $\phi(x)$; let $\operatorname{vbl}(z) \subseteq \Nats$ denote the indices of basis elements on which a constraint functional $z$ depends, and let $\phi(y)[\operatorname{vbl}(z)]$ denote the corresponding elements of the basis representation of $y \in \Omega$.
+Finally, for $A \subset \Nats$ a (possibly infinite) subset of indices for basis elements, and $a$ a vector of weights for those elements, let $P_{|B}(E | a)$ denote the conditional $\mu$-measure of event $E$ given the event $(\phi[A])^{-1}(a)$.
+This generalizes the projections of product measures from which we resampled variable subsets in the finite example.
+Having eliminated the assumption that $\mu$ is a product measure, this is a natural distribution for resampling.
 
-It is straightforward to write down an MT algorithm for solving this problem:
+We can write down an MT-\emph{like} algorithm for solving this problem:
 
 \begin{algorithm}[H]
-\caption{An MT-like algorithm for problem \ref{prob:feas}.}
-\label{alg:mt-feas}
+\caption{An MT-like algorithm for problem \ref{prob:feas-infinitedim}.}
+\label{alg:mt-feas-infinitedim}
 \begin{algorithmic}[1]
 
-\State Initialize $x \hasDist \operatorname{Unif(\mathcal{B}(1, n, ||\cdot||_\infty))}$.
+\State Initialize $x \hasDist \mu$.
 \ForAll{$i \in [m]$}
@@
   \EndIf
-  \State Resample $x[\operatorname{vbl}(A_i)]$ from $\mu^{|\operatorname{vbl}(A_i)}$.
+  \State Resample $\phi(x)[\operatorname{vbl}(A_i)]$ from $P_{|\Nats \setminus \operatorname{vbl}(A_i)}(\cdot | \phi(x)[\Nats \setminus \operatorname{vbl}(A_i)])$.
   \ForAll{$j \in [m]: j \in \Gamma(i), A_j(x) = 1$}
@@
 
-We will provide a proof that this algorithm finishes in expected polynomial time under conditions analogous to the LLL conditions.  Specifically, we will make the following assumptions:
+Soon we will provide a proof that this algorithm finishes in expected polynomial time under conditions analogous to the LLL conditions.  Specifically, we will make the following assumptions:
+
+%FIXME: No idea what these should be, really.
 \begin{enumerate}
-  \item Each $A_i$ is a function of only $k$ elements of $x$.  \label{eqn:feas-sparsity-condition}
-  \item Each $A_i$ is $\mu^n$-measurable and satisfies $\mu^n(\{x: A_i(x) = 1\}) \leq 2^{-k}$.
-(That is, most of the hypercube is feasible for any single constraint.)  Otherwise the constraints can be arbitrarily complicated(!)  \label{eqn:feas-margprob-condition}
-  \item The dependency graph induced by our constraints meets a slightly-strengthened version of the usual LLL condition: $2^{C_0} 2^{-k} d < 1$, for some absolute constant $C_0$.
-(For reference, the usual condition is $e 2^{-k} (d + 1) < 1$.
-The requirement for a strengthened condition may be an artifact of the proof.) \label{eqn:feas-lowdep-condition}
+  \item Each $A_i$ is $\mu$-measurable.
+  \item Further, each flaw must be addressed with positive probability when we resample it.
+More precisely, there is a positive lower bound $p^*$ on the probability of each flaw not happening when we resample it conditional on the variables on which it does not depend, regardless of the value of those variables:
+    \seqn
+      \label{eqn:feas-condprob-condition}
+      p^* = \inf_{A_i \in \mathcal{A}} \essinf_{x \in \Omega} P_{|\Nats \setminus \operatorname{vbl}(A_i)}(\bar{A_i} | \phi(x)[\Nats \setminus \operatorname{vbl}(A_i)]) .
+    \eeqn
+    $p^*$ is analogous to $p$ in the standard LLL.
+  \item The dependency graph induced by our constraints meets a slightly-strengthened version of the usual LLL condition: $2^{C_0} p^* (d+1) < 1$, for some absolute constant $C_0$.
+(If $C_0 = \log_2 e$ then this result specializes to the standard LLL.) \label{eqn:feas-lowdep-strong-condition}
 \end{enumerate}
 
-This problem now has many of the ingredients of an LLL problem.
+% Some possible alternative conditions, from A-I paper:
+% \begin{enumerate}
+%   \item Each $A_i$ is $\mu$-measurable.
+%   \item Further, each flaw must be addressed with positive probability when we resample it.
+More precisely, there is a positive lower bound $p^*$ on the probability of each flaw not happening when we resample it conditional on the variables on which it does not depend, regardless of the value of those variables:
+%     \seqn
+%       \label{eqn:feas-margprob-condition}
+%       p^* = \inf_{A_i \in \mathcal{A}} \essinf_{x \in \Omega} P_{|\Nats \setminus \operatorname{vbl}(A_i)}(\bar{A_i} | \phi(x)[\Nats \setminus \operatorname{vbl}(A_i)]) .
+%     \eeqn
+%     $p^*$ is analogous to $p$ in the standard LLL.
+%   \item The dependency graph induced by our constraints meets a slightly-strengthened version of the usual LLL condition: $2^{C_0} p^* (d+1) < 1$, for some absolute constant $C_0$.
+\label{eqn:feas-lowdep-condition}
+% \end{enumerate}
+
+%FIXME: Explain that this algorithm is not obviously implementable because the space is infinite-dimensional.
+For now, could just cite some sources that use infinite-dimensional objects by representing them implicitly, and argue that this could be possible if the constraint functions behave that way.
+Admit that we don't currently know of an application.
+
+%
+% This problem now has many of the ingredients of an LLL problem.
 (In fact, we can encode any instance of \ksat that satisfies the LLL conditions in this problem, mapping $x_i$ to \texttt{true} if it is in $[0,1]$ and \texttt{false} otherwise, and encoding constraints accordingly.)  The process for encoding the alternative information and constructing a witness forest from it is almost identical to that in Tao's argument.
-A technical difficulty is that the sample space is uncountable, so we can no longer speak of encoding a point using bits; this is where it seems useful to use the entropy inequality \ref{eqn:entropy-general} directly.
+A technical difficulty is that the sample space is uncountable, so we can no longer speak of encoding a point using bits; this is where it is useful to use the entropy inequality \ref{eqn:entropy-general} directly.
 A more practical difficulty is that we can no longer exactly reconstruct the algorithm state given the history of corrected flaws.
 To fix this, it will suffice to augment the alternative information slightly.
-
-Let us first calculate $H(Z)$ for a run of algorithm \ref{alg:mt-feas} that takes $t$ steps.
+%
+% Let us first calculate $H(Z)$ for a run of algorithm \ref{alg:mt-feas} that takes $t$ steps.
 This part is easy.
@@
 Each of our events is independent, so $H(Z) = H(U([-1,1])) (n + t k) = n + k t$.
-
-Now we define the alternative information $Y$ and find an upper bound for its entropy $H(Y)$.  We use almost the same construction as in Tao's proof: $Y$ is the concatenation of
-\begin{enumerate}
-  \item the witness forest (a number of bits, whose entropy is therefore bounded by its length),
-  \item the end state (whose entropy is at most $n$, the same as that of the initial algorithm state),
-  \item the initial set of violated constraints (which uses at most $m$ bits), and
-  \item one additional piece of information for each iteration, which was not needed in Tao's argument.
+%
+% Now we define the alternative information $Y$ and find an upper bound for its entropy $H(Y)$.  We use almost the same construction as in Tao's proof: $Y$ is the concatenation of
+% \begin{enumerate}
+%   \item the witness forest (a number of bits, whose entropy is therefore bounded by its length),
+%   \item the end state (whose entropy is at most $n$, the same as that of the initial algorithm state),
+%   \item the initial set of violated constraints (which uses at most $m$ bits), and
+%   \item one additional piece of information for each iteration, which was not needed in Tao's argument.
 To apply Tao's argument directly, we need to exactly reconstruct each state $\sigma_t$ from $Y$.
@@
 We are considering each item independently, which gives an upper bound, which is all we want.
-\end{enumerate}
-
-From these pieces and their individual entropy bounds (the sum of which bounds the joint entropy), $H(Y) \leq m + n + (k - C_0 + 2) t$, where $C_0$ is the constant in assumption \ref{eqn:feas-lowdep-condition}.
-
-Finally, since we have pinned down $Z$ using $Y$, $H(Z | Y) = 0$.
+% \end{enumerate}
+%
+% From these pieces and their individual entropy bounds (the sum of which bounds the joint entropy), $H(Y) \leq m + n + (k - C_0 + 2) t$, where $C_0$ is the constant in assumption \ref{eqn:feas-lowdep-condition}.
+%
+% Finally, since we have pinned down $Z$ using $Y$, $H(Z | Y) = 0$.
 Putting everything together, we get $m + n + (k - C_0 + 2) t \leq n + k t$, so that the algorithm must terminate (in expectation) before $t > \frac{m}{C_0 - 2}$.
@@
 (Note that the ordinary non-constructive LLL applies to this problem, and our assumptions are slightly stronger than the LLL conditions, so a solution must exist.)
-
-Some final notes on this algorithm:
-\begin{enumerate}
-  \item We chose our sample space to be a norm ball in finite dimensions, rather than say all of $\Reals^n$, precisely so that our sample space would be compact, allowing us to choose a simple probability measure as our dominating measure when computing entropies.
+%
+% Some final notes on this algorithm:
+% \begin{enumerate}
+%   \item We chose our sample space to be a norm ball in finite dimensions, rather than say all of $\Reals^n$, precisely so that our sample space would be compact, allowing us to choose a simple probability measure as our dominating measure when computing entropies.
 In general, to bound the number of steps taking by the algorithm, we need a probability measure that $Y$-almost surely dominates $P(Z | Y)$, for which the LLL-required marginal flaw probabilities hold, and which induces low dependency degree among the flaws.
-  \item We chose the infinity norm ball in particular for a different reason: the sample space needs to be an axis-aligned rectangle, or it is not easy to ensure that the resampling step keeps us inside it while modifying only $k$ variables at a time and covering a large volume.
+%   \item We chose the infinity norm ball in particular for a different reason: the sample space needs to be an axis-aligned rectangle, or it is not easy to ensure that the resampling step keeps us inside it while modifying only $k$ variables at a time and covering a large volume.
 Again, this is not important for proving something about the number of steps taken by the algorithm, but it is important for the efficiency of the individual steps.
-  \item Rather than augment $Y$ to pin down $Z$, we could imagine proving directly that $H(Z | Y)$ is small.
+%   \item Rather than augment $Y$ to pin down $Z$, we could imagine proving directly that $H(Z | Y)$ is small.
 It is natural to view $Z$ and $Y$ as the states and observations of a hidden Markov model, respectively, and use any of the well-known algorithms for inference in HMMs to compute $P(Z | Y)$.
 (Such inference could be quite difficult if the constraint functions are complicated, but for purposes of analysis we could imagine performing it.)  At least in this case, augmenting $Y$ was much simpler, but this makes the example weaker, since we are not taking advantage of our freedom in making $H(Z | Y) > 0$.
-  \item With respect to the constraint functions themselves, the algorithm can be made efficient as long as we have an oracle that can efficiently find a violated constraint when there is one.
+%   \item With respect to the constraint functions themselves, the algorithm can be made efficient as long as we have an oracle that can efficiently find a violated constraint when there is one.
 It may be possible to have an efficient version of the algorithm testing only a core of $\operatorname{poly}(n)$ ``core'' constraints using the ideas of Haeupler et al \cite{haeupler2011new}.
-\end{enumerate}
+% \end{enumerate}
 
@@
 
-\section{Future work}
-Our primary interest is in seeing whether we can find a general framework in which to apply the relative entropy compression argument.
-It would be helpful to find an example of an algorithm for which the ``focusing'' requirement of atomicity -- the requirement that only variables related to the addressed flaw be modified at each step -- is violated in a way that can be rescued by the relative entropy method.
-%FIXME: When example 3 above is filled in, delete this.
-We would also like to understand whether the results of \cite{achlioptas2014random} can be restated using an entropy compression argument, with the hope that our results can be extended to the non-variable setting.
+\section{A non-ergodic Markov chain interpretation}
+\label{sec:markov}
 
-In our review of the literature, several other topics interested us, though we did not get to them in this paper.  Here is a brief literature review:
-\begin{description}
-  \item[Connections to graphical model inference:] Graphical models are a very natural way to think about constraint satisfaction problems, though they are perhaps too general; sampling from the conditional distribution in a graphical model allows us to solve any CSP.
+\section{Nonatomic algorithms}
+\label{sec:nonatomic}
+
+% \section{Future work}
+% Our primary interest is in seeing whether we can find a general framework in which to apply the relative entropy compression argument.
+% It would be helpful to find an example of a real algorithm for which the ``focusing'' requirement of atomicity -- the requirement that only variables related to the addressed flaw be modified at each step -- is violated in a way that can be rescued by the relative entropy method.
+% We would also like to understand whether the results of \cite{achlioptas2014random} can be restated using an entropy compression argument, with the hope that our results can be extended to the non-variable setting.
+%
+% In our review of the literature, several other topics interested us, though we did not get to them in this paper.  Here is a brief literature review:
+% \begin{description}
+%   \item[Connections to graphical model inference:] Graphical models are a very natural way to think about constraint satisfaction problems, though they are perhaps too general; sampling from the conditional distribution in a graphical model allows us to solve any CSP.
 These connections are touched on in \cite{freer2010probabilistic} and more thoroughly in \cite{haeupler2011new}.
@@
 It would be interesting to see whether an MT-like algorithm could be a useful version of MCMC for some statistical problems (perhaps bearing provable guarantees, via a relative entropy compression argument, that are not available for other algorithms), or whether the LLL could apply to existing algorithms from machine learning.
-  \item[Parallel LLL algorithms:] Beck's algorithm was not parallel, but Alon \cite{alon1991parallel} published a parallel version soon after.
+%   \item[Parallel LLL algorithms:] Beck's algorithm was not parallel, but Alon \cite{alon1991parallel} published a parallel version soon after.
 Moser and Tardos \cite{moser2010constructive} used a similar (but much simplified) idea in their original paper.
@@
 It would also be interesting to see whether the general algorithm of Achlioptas and Iliopoulos \cite{achlioptas2014random} can be parallelized in a similar fashion.
-\end{description}
+% \end{description}
 

### 2015-06-18 julia/Applications/KsatTests.jl
Made solvers work for generic CSPs.

@@ -1,5 +1,5 @@
 using ArgParse
-using Utils, GeneratedData, Problems, Solvers
+using Utils, GeneratedData, Data, Problems, Solvers
 
 function compareGroundTruth(problem:: KsatProblem)
   println("Using exhaustive search to find a solution.")
@@ -7,7 +7,7 @@ function compareGroundTruth(problem:: KsatProblem)
   const trueSolution:: AnnotatedKsatSolution = solve(exhaustiveSolver, problem)
   if (isSuccessful(trueSolution))
     println("The problem has a solution: $(string(trueSolution))")
-    println("$(trueSolution.numSatisfyingSolutions)/$(trueSolution.numPotentialSolutions) ($(trueSolution.numSatisfyingSolutions/trueSolution.numPotentialSolutions)) satisfying solutions.")
+    println("($(trueSolution.numSatisfyingSolutions/trueSolution.numPotentialSolutions)) satisfying solutions.")
   else
     println("The problem has no solution.")
   end
@@ -37,18 +37,17 @@ end
 
 function run()
   const args = parseArgs()
-  srand(args["problem-seed"])
+  if args["problem-seed"] != nothing srand(args["problem-seed"]) end
   const problemGenerator = eval(parse(args["data-generator"]))
   const problem = generate(problemGenerator)
-  srand(args["solver-seed"])
+  if args["solver-seed"] != nothing srand(args["solver-seed"]) end
   const solver = eval(parse(args["solver"]))
   if args["compare-ground-truth"]
     compareGroundTruth(problem)
   end
   const solution = solve(solver, problem)
-  #println("$(isSuccessful(solution) ? "Successful" : "Unsuccessful") solution found for $(problem.k)-SAT problem with n=$(problem.numVariables), numClauses=$(length(problem.clauses))")
-  #println("Check: $(checkSuccess(solution.assignment, problem))")
-  #println("Solution: $(solution.assignment)")
+  println("$(isSuccessful(solution) ? "Successful" : "Unsuccessful") solution found for problem with n=$(problem.numVariables), numClauses=$(length(problem.clauses))")
+  println("Solution: $(solution.assignment)")
 end
 
 run()

### 2015-06-18 julia/Applications/KsatTests_solution.jl

@@ -55,7 +55,7 @@ function run()
     const problem = generate(problemGenerator)
     #println("[Seed:$(i)] $(string(problem))")
 
-    const graph = makeKsatDependencyGraph(problem)
+    const graph = makeGraphWithCriterion(problem, SharedVariable)
     max = maxDegree(graph)
     println("[Seed:$(i)] $(max)")
     push!(sum_maxDegree,max)
@@ -85,7 +85,7 @@ function run()
 
   #const solution = solve(solver, problem)
   #println("$(isSuccessful(solution) ? "Successful" : "Unsuccessful") solution #found for $(problem.k)-SAT problem with n=$(problem.numVariables), numClauses=$(length(problem.clauses))")
-  #println("Check: $(checkSuccess(solution.assignment, problem))")
+  #println("Check: $(isSatisfied(problem, solution.assignment))")
   #println("Solution: $(solution.assignment)")
 end
 

### 2015-06-18 julia/Applications/RandomKsatProblemStatistics.jl

@@ -1,0 +1,99 @@
+using ArgParse
+using GeneratedData, Problems, Solvers, Utils
+
+using Gadfly
+
+
+function parseArgs()
+  s = ArgParseSettings()
+  
+  @add_arg_table s begin
+    "--data-generator", "-d"
+      help = "the DataGenerator to use, a Julia string"
+    "--seed", "-e"
+      help = "the random seed" #FIXME
+      arg_type = Int
+  end
+
+  return parse_args(s)
+end
+
+
+
+function computeDegreeStatistics(problemGenerator:: DataGenerator)
+  maxDegrees = []
+  averageDegrees = []
+  for i = 1:1000
+    const problem = generate(problemGenerator)
+    const problemGraph = makeGraphWithCriterion(problem, SharedVariable)
+    push!(maxDegrees, maxDegree(problemGraph))
+    push!(averageDegrees, totalDegree(problemGraph) / length(nodes(problemGraph)))
+  end
+  println(maxDegrees)
+  println(averageDegrees)
+end
+
+
+function independenceLrTestStatistic(samples:: AbstractVector{AbstractVector{Float64}})
+  const numSamples = length(samples)
+  const n = numSamples > 0 ? length(samples[1]) : 0
+  const jointEmpiricalDist = [k => v / numSamples for (k, v) in countUniques(samples)]
+  const dependentLikelihood = mapreduce(p -> p*log(p), +, values(jointEmpiricalDist))
+  const marginalEmpiricalDists = map(1:n) do edgeIdx
+    mapreduce(sample -> sample[edgeIdx] / numSamples, +, samples)
+  end
+  const nullLikelihood = mapreduce(p*log(p)+(1-p)*log(1-p), +, marginalEmpiricalDists)
+  dependentLikelihood / nullLikelihood
+end
+
+
+function testEdgeMutualIndependence(problemGenerator:: DataGenerator, edgesToTest:: AbstractVector{(Int64,Int64)})
+  # We test the null hypothesis that the presence of @edgesToTest are mutually 
+  # independent under the distribution @problemGenerator.
+  # 
+  # We use a likelihood ratio test.  Let n = |edgesToTest|.  The test statistic 
+  # is:
+  #   [\sup_{probability measures m on {0,1}^n} l_m(data)] /
+  #   [\sup_{product measures m on {0,1}^n}] l_m(data)]
+  #   = f(data) / g(data),
+  # where:
[…]

### 2015-06-18 julia/Data/CnfSatReader.jl

@@ -1,0 +1,50 @@
+export CnfSatReader, generate
+
+using GeneratedData
+using Distributions
+using Problems
+
+immutable CnfSatReader <: DataGenerator
+  path:: String
+end
+
+function decompressed(f:: Function, path:: String)
+  if endswith(path, ".lzma")
+    open(f, `lzcat $path`, "r", STDIN)
+  else
+    open(f, path, "r", STDIN)
+  end
+end
+
+function readProblemDescription!(cnfFile)
+  for (lineIdx, line) in enumerate(eachline(cnfFile))
+    if beginswith(line, "c")
+      continue
+    elseif beginswith(line, "p cnf")
+      const problemDescriptionMatch = match(r"([0-9]+) ([0-9]+)", line, length("p cnf")+1)
+      const n = int(problemDescriptionMatch.captures[1])
+      const m = int(problemDescriptionMatch.captures[2])
+      return (n, m)
+    else
+      error("Cannot process file $(f): Syntax error at line $(lineIdx) ($(line)).")
+    end
+  end
+end
+
+function GeneratedData.generate(this:: CnfSatReader)
+  decompressed(this.path) do file
+    const n, m = readProblemDescription!(file)
+    const clauses = map(eachline(file)) do line
+      # The last number on each line is a 0, which is to be ignored.
+      const varStrings = split(line, " ")[1:(end-1)]
+      # We use 1-based-indexing because Julia is weird that way.
+      const vars = map(v -> abs(int(v)), varStrings)
+      const signs = map(v -> !beginswith(v, "-"), varStrings)
+      SatClause(vars, signs)
+    end
+    assert(length(clauses) == m)
+    assert(maximum(map(clause -> maximum(vbl(clause)), clauses)) <= n)
+    SatProblem(n, clauses)
+  end
+end
+

### 2015-06-18 julia/Data/SatToKsatConvertingGenerator.jl

@@ -1,0 +1,13 @@
+export SatToKsatConvertingGenerator, generate
+
+using GeneratedData
+using Problems
+
+immutable SatToKsatConvertingGenerator <: DataGenerator
+  satGenerator:: DataGenerator
+  strategy:: SatToKsatStrategy
+end
+
+function GeneratedData.generate(this:: SatToKsatConvertingGenerator)
+  satToKsat(generate(this.satGenerator), this.strategy)
+end

### 2015-06-18 julia/Data/module.jl

@@ -1,3 +1,6 @@
 module Data
 
+include("./SatToKsatConvertingGenerator.jl")
+include("./CnfSatReader.jl")
+
 end

### 2015-06-18 julia/GeneratedData/RandomKsat.jl

@@ -17,3 +17,16 @@ function generate(this:: RandomKsatGenerator)
   end
   KsatProblem(this.k, this.numVariables, clauses)
 end
+
+# See http://arxiv.org/pdf/cs/0309020.pdf
+SMALL_K_THRESHOLDS = [4.267, 9.931, 21.117, 43.37, 87.79]
+MAX_SMALL_K = 7
+function makeHardGenerator(k:: Int64, numVariables:: Int64)
+  const numClauses = if k <= MAX_SMALL_K
+    SMALL_K_THRESHOLDS[k-2]
+  else
+    # See https://users.soe.ucsc.edu/~optas/papers/ksat-ams.pdf
+    #FIXME
+  end
+  RandomKsatGenerator(k, numVariables, numClauses)
+end

### 2015-06-18 julia/Problems/BinaryAssignment.jl

@@ -1,0 +1,18 @@
+export BinaryAssignment, defaultAssignment, uniformRandomBinaryAssignment, emptyBinaryAssignment
+
+type BinaryAssignment <: VariableAssignment{Binary}
+  #FIXME: Maybe should use a concrete type here for performance.
+  vars:: AbstractVector{Bool}
+end
+
+function defaultAssignment(numVariables:: Int64, :: Type{BinaryAssignment})
+  emptyBinaryAssignment(numVariables)
+end
+
+function uniformRandomBinaryAssignment(numVariables:: Int64)
+  BinaryAssignment(randbool(numVariables))
+end
+
+function emptyBinaryAssignment(numVariables:: Int64)
+  BinaryAssignment(falses(numVariables))
+end

### 2015-06-18 julia/Problems/KsatProblem.jl

@@ -1,78 +1,76 @@
-export KsatProblem, numClauses, SatClause, KsatSolution, isSuccessful, unsuccessfulKsatSolution, KsatAssignment, isSatisfied, checkSuccess, uniformRandomAssignment
+export KsatProblem, k, constraints, numVars, marginalSample!, constraintProb
+export SatToKsatStrategy, SatToPaddedKsatStrategy, satToKsat
+export KsatSolution, isSuccessful, unsuccessfulKsatSolution
 
-# A disjunction of variables.
-immutable SatClause
-  variables:: AbstractVector{Int64}
-  signs:: AbstractVector{Bool}
-end
-
-function Base.string(this:: SatClause)
-  join(
-    map(1:length(this.variables)) do varIdx
-      const prefix = this.signs[varIdx] ? "" : "!"
-      "$(prefix)$(this.variables[varIdx])"
-    end,
-    "|")
-end
-
+#FIXME: Mostly copied from SatProblem.  Since Julia does not have inheritance
+# for concrete types, it's not straightforward to fix that.
 # A conjunction of disjunctive clauses (an AND of numVariables clauses, each
-# clause being an OR of k variables).
-immutable KsatProblem <: Problem
+# clause being an OR of k variables).  Each clause is of length exactly k.
+immutable KsatProblem <: ProductMeasureCsp{Binary, BinaryAssignment, SatClause}
   k:: Int64
   numVariables:: Int64
   clauses:: AbstractVector{SatClause}
 end
 
-const MAX_DISPLAYED_LITERALS = 20000000
-function Base.string(this:: KsatProblem)
-  const maxDisplayedClauses = int(MAX_DISPLAYED_LITERALS / this.k)
-  const numDisplayedClauses = min(maxDisplayedClauses, numClauses(this))
-  const clausesString = join(map(c -> "($(string(c)))", this.clauses[1:numDisplayedClauses]), "&")
-  "$(this.k)-SAT problem $(clausesString)$(numDisplayedClauses < numClauses(this) ? "..." : "")"
+function k(this:: KsatProblem)
+  this.k
 end
 
-function numClauses(this:: KsatProblem)
-  length(this.clauses)
+function constraints(this:: KsatProblem)
+  this.clauses
 end
 
+function numVars(this:: KsatProblem)
+  this.numVariables
+end
 
-typealias KsatAssignment AbstractVector{Bool}
-
-function isSatisfied(clause:: SatClause, assignment:: KsatAssignment)
-  # A clause is disjunctive -- an OR of variables.
[…]

### 2015-06-18 julia/Problems/Problems.jl

@@ -1,9 +1,201 @@
-export Problem, Solution, isSuccessful
+export Assignment
+export Problem
+export Constraint, isSatisfied
+export Csp, m, constraints
+export VariableType, Binary, Discrete, Real, NonnegativeReal
+export VariableAssignment
+export VariableConstraint, vbl, k
+export VariableCsp, numVars, variables
+export ProductMeasureCsp, jointSample, marginalSample!, constraintProb
+export ProblemSolution, isSuccessful
+export Solution, successfulSolution, unsuccessfulSolution
 
-abstract Problem
+using Utils
 
-abstract Solution
+# A member of the domain of a problem.  For example, a vector of 5 real
+# numbers.  Not necessarily a feasible or optimal solution.  For performance
+# reasons, instances should be mutable.
+abstract Assignment
+
+#FIXME: Probably unnecessary.
+# Some kind of zero or default element of the domain of a problem.  Examples:
+# an assignment of all zeros; the trivial permutation.
+function defaultAssignment{A <: Assignment}(:: Type{A})
+  raiseAbstract("defaultAssignment", A)
+end
+
+
+abstract Problem{A <: Assignment}
+
+
+abstract Constraint{A <: Assignment}
+
+function isSatisfied{A <: Assignment}(this:: Constraint{A}, assignment:: A)
+  raiseAbstract("isSatisfied", this)
+end
+
+
+# A problem in which we want to find an element of some domain that satisfies
+# m separate constraints.  m is finite, and it is assumed that the constraints
+# (whatever their internal semantics) are indexed by integers from 1 to m.
+# C <: Constraint{A}
+abstract Csp{C <: Constraint, A <: Assignment} <: Problem{A}
+
+function m(this:: Csp)
+  length(constraints(this))
+end
+
+#FIXME: numClauses() is deprecated.
+function numClauses(this:: Csp)
+  m(this)
+end
+
+function constraints(this:: Csp)
+  raiseAbstract("constraints", this)
+end
+
+#FIXME: Why doesn't the compiler like this?
+# C <: Constraint{A}
+function isSatisfied{C <: Constraint, A <: Assignment}(this:: Csp{C, A}, assignment:: A)
+  for c in constraints(this)
+    if !isSatisfied(c, assignment)
+      return false
+    end
+  end
+  return true
+end
+
+const MAX_DISPLAYED_CONSTRAINTS = 100
+function Base.string(this:: Csp)
[…]

### 2015-06-18 julia/Problems/SatClause.jl

@@ -1,0 +1,87 @@
+export SatClause, vbl, isSatisfied, shareVariable, haveNegativeDependency
+
+# A disjunction of variables (for example, "A or B or (not D) or G").
+immutable SatClause <: VariableConstraint{BinaryAssignment, Binary}
+  variables:: Vector{Int64}
+  signs:: BitArray
+  
+  function SatClause(variables:: Vector{Int64}, signs:: BitVector)
+    assert(length(variables) == length(signs))
+    new(variables, signs)
+  end
+end
+
+function SatClause(variables:: AbstractVector{Int64}, signs:: AbstractVector{Bool})
+  SatClause([variables], bitpack(signs))
+end
+
+function SatClause(variables:: AbstractVector{Int64}, signs:: AbstractVector{Int64})
+  SatClause([variables], bitpack(signs))
+end
+
+function vbl(this:: SatClause)
+  this.variables
+end
+
+function isSatisfied(this:: SatClause, assignment:: BinaryAssignment)
+  # A clause is disjunctive -- an OR of variables.
+  for (variableIdx, variable) in enumerate(this.variables)
+    const value = assignment.vars[variable]
+    if this.signs[variableIdx] == value
+      return true
+    end
+  end
+  return false
+end
+
+function Base.string(this:: SatClause)
+  join(
+    map(1:length(this.variables)) do varIdx
+      const prefix = this.signs[varIdx] ? "" : "!"
+      "$(prefix)$(this.variables[varIdx])"
+    end,
+    "|")
+end
+
+function shareVariable(clauseA:: SatClause, clauseB:: SatClause)
+  if length(vbl(clauseA)) < 50
+    for varA in vbl(clauseA)
+      for varB in vbl(clauseB)
+        if varA == varB
+          return true
+        end
+      end
+    end
+    false
+  else
+    # Use a linear-time algorithm with higher overhead when k is large.
+    #TODO: Should probably store Set(vbl(clause)) in the clauses anyway.
+    const varsA = Set(vbl(clauseA))
+    const varsB = Set(vbl(clauseB))
+    !isempty(intersect(varsA, varsB))
+  end
+end
+
+# True if A and B overlap on a variable where A's copy and B's copy have 
+# opposite signs.
+function haveNegativeDependency(clauseA:: SatClause, clauseB:: SatClause)
+  if length(vbl(clauseA)) < 50
+    for (signA, varA) in zip(clauseA.signs, clauseA.variables)
[…]

### 2015-06-18 julia/Problems/SatProblem.jl

@@ -1,0 +1,37 @@
+export SatLikeProblem
+export SatProblem, constraints, numVars, marginalSample, constraintProb
+export SatSolution, isSuccessful, unsuccessfulSatSolution
+
+typealias SatLikeProblem ProductMeasureCsp{Binary, BinaryAssignment, SatClause}
+
+# A conjunction of disjunctive clauses (an AND of numVariables clauses, each
+# clause being an OR of variables).
+immutable SatProblem <: SatLikeProblem
+  numVariables:: Int64
+  clauses:: AbstractVector{SatClause}
+end
+
+function constraints(this:: SatProblem)
+  this.clauses
+end
+
+function numVars(this:: SatProblem)
+  this.numVariables
+end
+
+function marginalSample!(this:: SatProblem, vbls:: AbstractVector{Int64}, assignment:: BinaryAssignment)
+  const newValues = randbool(length(vbls))
+  for (i, variableIdx) in enumerate(vbls)
+    assignment.vars[variableIdx] = newValues[i]
+  end
+end
+
+# The marginal probability that constraint @constraintIdx is satisfied under
+# the measure on the variables of @this.
+function constraintProb(this:: SatProblem, constraintIdx:: Int64)
+  2^(-k(this.clauses[constraintIdx]))
+end
+
+function couldBeNegativelyCorrelated(this:: ProductMeasureCsp, constraintA:: Int64, constraintB:: Int64)
+  haveNegativeCorrelation(constraintA, constraintB)
+end

### 2015-06-18 julia/Problems/module.jl

@@ -1,6 +1,9 @@
 module Problems
 
 include("./Problems.jl")
+include("./BinaryAssignment.jl")
+include("./SatClause.jl")
+include("./SatProblem.jl")
 include("./KsatProblem.jl")
 
 end

### 2015-06-18 julia/Solvers/CspDependencyGraph.jl

@@ -1,4 +1,8 @@
-export DependencyGraph, nodes, edges, hasEdge, hasNode, degree, neighbors, neighborhood, maxDegree, MarkedGraph, KsatDependencyGraph, makeKsatDependencyGraph, inducedSubgraph
+export DependencyGraph, nodes, edges, hasEdge, hasNode, degree, neighbors, neighborhood, maxDegree, inducedSubgraph
+export MarkedGraph
+export CspDependencyGraph, makeGraphWithEdgeCriterion, inducedSubgraph
+export EdgeCriterion, SharedVariable, NegativeCorrelation
+
 
 abstract DependencyGraph
 
@@ -6,7 +10,7 @@ function nodes(this:: DependencyGraph)
   raiseAbstract("nodes", this)
 end
 
-# A Set{Int64} of directed edges in the graph (each undirected edge (i,j)
+# A Set{(Int64,Int64)} of directed edges in the graph (each undirected edge (i,j)
 # counts as both (i,j) and (j,i)).
 function edges(this:: DependencyGraph)
   raiseAbstract("edges", this)
@@ -23,6 +27,11 @@ end
 # True if there is an edge between @a and @b.
 function hasEdge(this:: DependencyGraph, a:: Int64, b:: Int64)
   (a, b) in edges(this)
+end
+
+# True if there is an edge e = (a,b).
+function hasEdge(this:: DependencyGraph, e:: (Int64, Int64))
+  e in edges(this)
 end
 
 # True if this graph contains a node with ID @node.
@@ -48,7 +57,7 @@ function totalDegree(this:: DependencyGraph)
 end
 
 function inducedSubgraph(this:: DependencyGraph, nodeSubset:: Set{Int64})
-  KsatDependencySubgraph(this, intersect(nodes(this), nodeSubset))
+  CspDependencySubgraph(this, intersect(nodes(this), nodeSubset))
 end
 
 
@@ -58,8 +67,8 @@ immutable MarkedGraph{MarkType}
 end
 
 
-immutable KsatDependencyGraph <: DependencyGraph
-  problem:: KsatProblem
+immutable CspDependencyGraph <: DependencyGraph
+  problem:: VariableCsp
   #TODO: Could use a full IntSet for this.
   nodes:: Set{Int64}
   # Two clauses have an edge if they share a variable (its sign doesn't
@@ -69,57 +78,72 @@ immutable KsatDependencyGraph <: DependencyGraph
[…]

### 2015-06-18 julia/Solvers/ExhaustiveBinarySolver.jl

@@ -1,0 +1,28 @@
+export ExhaustiveBinarySolver, solve
+
+immutable ExhaustiveBinarySolver <: Solver{Problem{BinaryAssignment}} end
+
+function solve(this:: ExhaustiveBinarySolver, problem:: Problem{BinaryAssignment})
+  const n = numVars(problem)
+  assignment = defaultAssignment(n, BinaryAssignment)
+  assignmentBits = assignment.vars
+  satisfyingAssignment = defaultAssignment(n, BinaryAssignment)
+  numSatisfyingSolutions = 0
+  const numPotentialSolutions = 2^n
+
+  for binaryAssignment = 0:(numPotentialSolutions-1)
+    #TODO: Could be done much faster with a Gray code or a direct conversion
+    # from binary to BitVector.
+    for variableIdx = 1:n
+      assignmentBits[variableIdx] = (binaryAssignment >> (variableIdx-1)) & 0x1
+    end
+    if isSatisfied(problem, assignment)
+      if numSatisfyingSolutions == 0
+        satisfyingAssignment = deepcopy(assignment)
+      end
+      numSatisfyingSolutions += 1
+    end
+  end
+
+  return AnnotatedKsatSolution(satisfyingAssignment, numSatisfyingSolutions > 0, numSatisfyingSolutions, numPotentialSolutions)
+end

### 2015-06-18 julia/Solvers/ExhaustiveKsatSolver.jl

@@ -1,35 +1,0 @@
-export ExhaustiveKsatSolver, solve
-
-immutable ExhaustiveKsatSolver <: KsatSolver end
-
-function solve(this:: ExhaustiveKsatSolver, problem:: KsatProblem)
-  const n = problem.numVariables
-  assignment = BitVector(n)
-  satisfyingAssignment = uniformRandomAssignment(n)
-  numSatisfyingSolutions = 0
-
-
-#  const numPotentialSolutions = 2^n
-  if n < 11
-    const numPotentialSolutions = 2^n
-  else
-    const numPotentialSolutions = 2^11
-  end
-
-  for binaryAssignment:: Uint64 = 1:numPotentialSolutions
-    #TODO: Could be done much faster with a Gray code or a direct conversion
-    # from binary to BitVector.
-    for variableIdx = 1:n
-      assignment[variableIdx] = (binaryAssignment >> (variableIdx-1)) & 0x1
-    end
-    if checkSuccess(assignment, problem)
-      if numSatisfyingSolutions == 0
-        satisfyingAssignment = deepcopy(assignment)
-      end
-      numSatisfyingSolutions += 1
-    end
-  end
-
-
-  return AnnotatedKsatSolution(satisfyingAssignment, numSatisfyingSolutions > 0, numSatisfyingSolutions, numPotentialSolutions)
-end

### 2015-06-18 julia/Solvers/IndependentSets.jl

@@ -3,7 +3,7 @@ export maximalIndependentSet, lubyIndependentSet
 using Distributions
 
 # Some algorithms for finding independent sets in undirected graphs.
-# Depends on KsatDependencyGraph.jl.
+# Depends on CspDependencyGraph.jl.
 # 
 # Some of the algorithms here could, in principle, be parallelized or 
 # distributed.  Currently we have not done that, but we have marked places that

### 2015-06-18 julia/Solvers/KsatUtils.jl

@@ -1,40 +1,31 @@
-export AnnotatedKsatProblem, unsatisfiedClauses, variablesToClauses, toSuccessfulSolution, toUnsuccessfulSolution, annotateKsatProblem, updateAssignment!, registerUpdate!, AnnotatedKsatSolution, isSuccessful
+export AnnotatedSatProblem, unsatisfiedClauses, variablesToClauses, annotateSatProblem, updateAssignment!, registerUpdate!, AnnotatedSatSolution, isSuccessful
 
 using Problems
 
 # A k-SAT problem with some extra metadata useful for an iterative solver,
 # including a current assignment to the variables.
-type AnnotatedKsatProblem
-  problem:: KsatProblem
+type AnnotatedSatProblem
+  problem:: SatLikeProblem
   # Mutable.
-  currentAssignment:: KsatAssignment
+  currentAssignment:: BinaryAssignment
   variablesToClauses:: Dict{Int64, Set{Int64}}
   # Mutable.
   unsatisfiedClauses:: Set{Int64} #TODO: Use IntSet instead?
 end
 
-function unsatisfiedClauses(this:: AnnotatedKsatProblem)
+function unsatisfiedClauses(this:: AnnotatedSatProblem)
   this.unsatisfiedClauses
 end
 
-function variablesToClauses(this:: AnnotatedKsatProblem)
+function variablesToClauses(this:: AnnotatedSatProblem)
   this.variablesToClauses
 end
 
-function toSuccessfulSolution(this:: AnnotatedKsatProblem)
-  #TODO: Assert unsatisfiedClauses isempty.
-  KsatSolution(this.currentAssignment, true)
-end
-
-function toUnsuccessfulSolution(this:: AnnotatedKsatProblem)
-  KsatSolution(this.currentAssignment, false)
-end
-
-function annotateKsatProblem(p:: KsatProblem, initialAssignment:: KsatAssignment)
+function annotateSatProblem(p:: SatLikeProblem, initialAssignment:: BinaryAssignment)
   variablesToClauses = Dict{Int64, Set{Int64}}()
   unsatisfiedClauses = Set{Int64}()
-  for (clauseIdx, clause) in enumerate(p.clauses)
-    for variable in clause.variables
+  for (clauseIdx, clause) in enumerate(constraints(p))
+    for variable in vbl(clause)
       if !haskey(variablesToClauses, variable)
         variablesToClauses[variable] = Set(clauseIdx)
       else
[…]

### 2015-06-18 julia/Solvers/LllConditionChecker.jl

@@ -1,0 +1,23 @@
+export checkSimpleLllCondition, checkStandardLllCondition
+
+function checkSimpleLllCondition(problem:: ProductMeasureCsp)
+  checkSimpleLllCondition(problem, makeGraphWithCriterion(problem, SharedVariable))
+end
+
+function checkSimpleLllCondition(problem:: ProductMeasureCsp, dependencyGraph:: DependencyGraph)
+  const p = maximum(map(i -> constraintProb(problem, i), 1:m(problem)))
+  const d = maxDegree(dependencyGraph)
+  e*p*(d+1) < 1
+end
+
+function checkStandardLllCondition(problem:: ProductMeasureCsp)
+  checkStandardLllCondition(problem, makeGraphWithCriterion(problem, SharedVariable))
+end
+
+function checkLopsidedLllCondition(problem:: ProductMeasureCsp)
+  checkStandardLllCondition(problem, makeGraphWithCriterion(problem, NegativeCorrelation))
+end
+
+function checkStandardLllCondition(problem:: ProductMeasureCsp, dependencyGraph:: DependencyGraph)
+  #FIXME
+end

### 2015-06-18 julia/Solvers/ParallelLllKsatSolver.jl

@@ -33,14 +33,15 @@ immutable ParallelLllKsatSolver <: Solver{KsatProblem}
   independentSetFinder:: IndependentSetFinder
 end
 
+#FIXME
 function solve(this:: ParallelLllKsatSolver, problem:: KsatProblem)
   # See Moser and Tardos 2010, algorithms 1 and 2.
-  const n = problem.numVariables
+  const n = numVars(problem)
   #println("Problem: $(string(problem))")
-  assignment = uniformRandomAssignment(n)
-  annotatedProblem = annotateKsatProblem(problem, assignment)
+  assignment = uniformRandomBinaryAssignment(n)
+  annotatedProblem = annotateSatProblem(problem, assignment)
   #NOTE: Could be parallelized.
-  const graph = makeKsatDependencyGraph(problem)
+  const graph = makeGraphWithCriterion(problem, SharedVariable)
   const maxNumIterations = calculateNumIterations(this.independentSetFinder, problem, graph)
   const independentSetFunc = buildFinderFunc(this.independentSetFinder, graph)
   #println("Using at most $(maxNumIterations) iterations for $(problem.k)-SAT problem with $(n) variables and $(numClauses(problem)) clauses.")
@@ -48,9 +49,8 @@ function solve(this:: ParallelLllKsatSolver, problem:: KsatProblem)
     #NOTE: Could be parallelized.
     const unsat = unsatisfiedClauses(annotatedProblem)
     if isempty(unsat)
-      println("$(i)")
-      #println("Found successful solution after $(i) iterations.")
-      return toSuccessfulSolution(annotatedProblem)
+      println("Found successful solution after $(i) iterations.")
+      return successfulSolution(annotatedProblem)
     end
     const independentSet = independentSetFunc(inducedSubgraph(graph, unsat))
     # println("Using an independent set of size $(length(independentSet)) on iteration $i")
@@ -61,10 +61,10 @@ function solve(this:: ParallelLllKsatSolver, problem:: KsatProblem)
     # graph.
     for clauseIdx in independentSet
       const clause = problem.clauses[clauseIdx]
-      for variable in clause.variables
+      for variable in vbl(clause)
         annotatedProblem.currentAssignment[variable] = randbool()
       end
-      registerUpdate!(annotatedProblem, clause.variables)
+      registerUpdate!(annotatedProblem, vbl(clause))
     end
   end
   unsuccessfulKsatSolution(n)

### 2015-06-18 julia/Solvers/ParallelLllKsatSolver_graphOnly.jl

@@ -1,9 +1,9 @@
-export ParallelLllKsatSolver, solve
+#DEPRECATED
 
+export ParallelLllKsatSolver, solve
 export IndependentSetFinder, MoserTardosIndependentSetFinder, SimpleChungIndependentSetFinder, WeakMisFinder, buildFinderFunc, calculateNumIterations
 
 using Problems
-
 using Distributions
 
 
@@ -37,40 +37,36 @@ function solve(this:: ParallelLllKsatSolver, problem:: KsatProblem)
   # See Moser and Tardos 2010, algorithms 1 and 2.
   const n = problem.numVariables
   println("Problem: $(string(problem))")
-  assignment = uniformRandomAssignment(n)
-  annotatedProblem = annotateKsatProblem(problem, assignment)
+  assignment = uniformRandomBinaryAssignment(n)
+  annotatedProblem = annotateSatProblem(problem, assignment)
   #NOTE: Could be parallelized.
-  println("here")
-#  const graph = makeKsatDependencyGraph(problem)
-
-
-
-#  const maxNumIterations = calculateNumIterations(this.independentSetFinder, problem, graph)
-#  const independentSetFunc = buildFinderFunc(this.independentSetFinder, graph)
-#  println("Using at most $(maxNumIterations) iterations for $(problem.k)-SAT problem with $(n) variables and $(numClauses(problem)) clauses.")
-#  for i = 1:maxNumIterations
-#    #NOTE: Could be parallelized.
-#    const unsat = unsatisfiedClauses(annotatedProblem)
-#    if isempty(unsat)
-#      println("Found successful solution after $(i) iterations.")
-#      return toSuccessfulSolution(annotatedProblem)
-#    end
-#    const independentSet = independentSetFunc(inducedSubgraph(graph, unsat))
-    # println("Using an independent set of size $(length(independentSet)) on iteration $i")
-    #NOTE: Could be parallelized.  For example, if clauses were distributed
+ const graph = makeGraphWithCriterion(problem, SharedVariable)
+ const maxNumIterations = calculateNumIterations(this.independentSetFinder, problem, graph)
+ const independentSetFunc = buildFinderFunc(this.independentSetFinder, graph)
+ println("Using at most $(maxNumIterations) iterations for $(problem.k)-SAT problem with $(n) variables and $(numClauses(problem)) clauses.")
+ for i = 1:maxNumIterations
+   #NOTE: Could be parallelized.
+   const unsat = unsatisfiedClauses(annotatedProblem)
+   if isempty(unsat)
[…]

### 2015-06-18 julia/Solvers/RandomWalkKsatSolver.jl

@@ -14,7 +14,7 @@ end
 
 function solve(this:: RandomWalkKsatSolver, problem:: KsatProblem)
   # See Mitzenmacher and Upfal 2005, section 7.1.2, algorithm 7.3.
-  const n = problem.numVariables
+  const n = numVars(problem)
   const numAssignmentsPerAttempt = 3*n
   const expectedNumStepsRequired = n^1.5 * (4.0/3.0)^n
   const numAttempts = int(ceil(2*(-1)*log2(this.failureProbability)*expectedNumStepsRequired))
@@ -25,6 +25,7 @@ end
 
 function findSolution(numAttempts:: Int64, problem:: KsatProblem)
   for attemptIdx in 1:numAttempts
+    # We call SimpleRandomKsatSolver.solve() on each attempt.
     solution = solve(SimpleRandomKsatSolver(), problem)
     if isSuccessful(solution)
       println("Found successful solution after $(attemptIdx) attempts.")
@@ -42,24 +43,24 @@ function solve(this:: SimpleRandomKsatSolver, problem:: KsatProblem)
   # See Mitzenmacher and Upfal 2005, section 7.1.2, algorithm 7.3.
   const n = problem.numVariables
   const numAssignments = 3*n
-  assignment = uniformRandomAssignment(n)
-  annotatedProblem = annotateKsatProblem(problem, assignment)
+  assignment = uniformRandomBinaryAssignment(n)
+  annotatedProblem = annotateSatProblem(problem, assignment)
   # println("Initially unsatisfied clauses: $(unsatisfiedClauses(annotatedProblem))")
   for assignmentIdx in 1:numAssignments
     # Pick an arbitrary unsatisfied clause.  The first one will do.
     unsat = unsatisfiedClauses(annotatedProblem)
     if isempty(unsat)
       println("Found successful solution after $(assignmentIdx) flips.")
-      return toSuccessfulSolution(annotatedProblem)
+      return successfulSolution(annotatedProblem.currentAssignment)
     end
     # println("Unsatisfied clauses at round $(assignmentIdx): $(unsatisfiedClauses(annotatedProblem))")
     clauseIdx = first(unsat)
     clause = problem.clauses[clauseIdx]
     variableInClauseIdx = rand(1:problem.k)
-    variable = clause.variables[variableInClauseIdx]
+    variable = vbl(clause)[variableInClauseIdx]
[…]

### 2015-06-18 julia/Solvers/SimpleMoserTardosKsatSolver.jl

@@ -1,30 +1,0 @@
-export SimpleMoserTardosKsatSolver, solve
-
-using Problems
-
-immutable SimpleMoserTardosKsatSolver <: Solver{KsatProblem}
-end
-
-function solve(this:: SimpleMoserTardosKsatSolver, problem:: KsatProblem)
-  # See Moser and Tardos 2010, algorithm 1.1.
-  const n = problem.numVariables
-  const maxNumIterations = 1000 #FIXME
-  println("Using at most $(maxNumIterations) iterations for $(problem.k)-SAT problem with $(n) variables and $(numClauses(problem)) clauses.")
-  println("Problem: $(string(problem))")
-  assignment = uniformRandomAssignment(n)
-  annotatedProblem = annotateKsatProblem(problem, assignment)
-  for i = 1:maxNumIterations
-    const unsat = unsatisfiedClauses(annotatedProblem)
-    if isempty(unsat)
-      println("Found successful solution after $(i) iterations.")
-      return toSuccessfulSolution(annotatedProblem)
-    end
-    const clauseIdx = first(unsat)
-    const clause = problem.clauses[clauseIdx]
-    for variable in clause.variables
-      annotatedProblem.currentAssignment[variable] = randbool()
-    end
-    registerUpdate!(annotatedProblem, clause.variables)
-  end
-  unsuccessfulKsatSolution(n)
-end

### 2015-06-18 julia/Solvers/SimpleMoserTardosSolver.jl

@@ -1,0 +1,59 @@
+export SimpleMoserTardosSolver, SimpleMtSatLikeSolver, solve
+
+using Utils
+using Problems
+
+# A <: VariableAssignment{V}
+# C <: VariableConstraint{A, V}
+immutable SimpleMoserTardosSolver{V <: VariableType, A <: VariableAssignment, C <: VariableConstraint} <: Solver{ProductMeasureCsp{V, A, C}}
+end
+
+typealias SimpleMtSatLikeSolver SimpleMoserTardosSolver{Binary, BinaryAssignment, SatClause}
+
+
+# A <: VariableAssignment{V}
+# C <: VariableConstraint{A, V}
+function solve{V <: VariableType, A <: VariableAssignment, C <: VariableConstraint}(this:: SimpleMoserTardosSolver{V, A, C}, problem:: ProductMeasureCsp{V, A, C})
+  # See Moser and Tardos 2010, algorithm 1.1.
+  const n = numVars(problem)
+  const maxNumIterations = 100000 #FIXME
+  println("Using at most $(maxNumIterations) iterations for $(string(problem)).")
+  assignment = jointSample(problem)
+  for i = 1:maxNumIterations
+    const unsat = filter(constraint -> isSatisfied(constraint, assignment), constraints(problem))
+    if isempty(unsat)
+      println("Found successful solution after $(i) iterations.")
+      return successfulSolution(assignment)
+    end
+    const constraintIdx = first(unsat)
+    const constraint = constraints(problem)[constraintIdx]
+    marginalSample!(problem, vbl(constraint), assignment)
+  end
+  unsuccessfulSolution(A)
+end
+
+# solve(), specialized to SAT and k-SAT for performance.
+#TODO: A lot of copied code.  Could probably merge this with the above.
+# I think the method of tracking unsatisfied constraints is completely general,
+# and we can get rid of this special case entirely.
+function solve(this:: SimpleMtSatLikeSolver, problem:: SatLikeProblem)
+  # See Moser and Tardos 2010, algorithm 1.1.
+  const n = numVars(problem)
+  const maxNumIterations = 100000 #FIXME
+  println("Using at most $(maxNumIterations) iterations for $(string(problem)).")
+  assignment = jointSample(problem)
+  annotatedProblem = annotateSatProblem(problem, assignment)
+  for i = 1:maxNumIterations
+    const unsat = unsatisfiedClauses(annotatedProblem)
+    if isempty(unsat)
[…]

### 2015-06-18 julia/Solvers/WalksatSolver.jl

@@ -1,0 +1,35 @@
+export WalksatSolver, solve
+
+using Problems
+
+# An implementation of the WALKSAT algorithm.
+immutable WalksatSolver <: Solver{KsatProblem}
+end
+
+function solve(this:: WalksatSolver, problem:: KsatProblem)
+  const n = problem.numVariables
+  const maxNumAssignments = 3*n #FIXME
+  assignment = uniformRandomBinaryAssignment(n)
+  annotatedProblem = annotateSatProblem(problem, assignment)
+  # println("Initially unsatisfied clauses: $(unsatisfiedClauses(annotatedProblem))")
+  for assignmentIdx in 1:numAssignments
+    # Pick an arbitrary unsatisfied clause.  The first one will do.
+    unsat = unsatisfiedClauses(annotatedProblem)
+    if isempty(unsat)
+      println("Found successful solution after $(assignmentIdx) flips.")
+      return successfulSolution(annotatedProblem.currentAssignment)
+    end
+    println("Unsatisfied clauses at round $(assignmentIdx): $(unsatisfiedClauses(annotatedProblem))")
+    clauseIdx = first(unsat)
+    clause = problem.clauses[clauseIdx]
+    variableIdx = argmax(1:k) do variableInClauseIdx
+      
+      #FIXME: Return # of satisfied clauses.
+    end
+    variable = vbl(clause)[variableInClauseIdx]
+    # println("Flipping variable $(variable) ($(variableInClauseIdx) in clause $(clauseIdx)/$(numClauses(problem)) $(clause))")
+    currentValue = annotatedProblem.currentAssignment[variable]
+    updateAssignment!(annotatedProblem, variable, currentValue, !currentValue)
+  end
+  unsuccessfulSolution(BinaryAssignment)
+end

### 2015-06-18 julia/Solvers/module.jl

@@ -2,11 +2,12 @@ module Solvers
 
 include("./Solvers.jl")
 include("./KsatUtils.jl")
-include("./KsatDependencyGraph.jl")
+include("./CspDependencyGraph.jl")
+include("./LllConditionChecker.jl")
 include("./IndependentSets.jl")
-include("./ExhaustiveKsatSolver.jl")
+include("./ExhaustiveBinarySolver.jl")
 include("./RandomWalkKsatSolver.jl")
-include("./SimpleMoserTardosKsatSolver.jl")
+include("./SimpleMoserTardosSolver.jl")
 include("./ParallelLllKsatSolver.jl")
 
 end

### 2015-06-18 julia/modules.jl

@@ -1,5 +1,5 @@
 require("./Utils/module.jl")
 require("./Problems/module.jl")
-require("./Data/module.jl")
 require("./GeneratedData/module.jl")
+require("./Data/module.jl")
 require("./Solvers/module.jl")

### 2015-06-18 julia/packages.jl

@@ -2,3 +2,4 @@
 
 Pkg.add("Distribution")
 Pkg.add("ArgParse")
+Pkg.add("Gadfly")

### 2015-06-18 paper/nonatomic/paper.tex
Some incomplete thoughts about the Markov chain interpretation of the entropy argument.

@@
 The original LLL does not tell us how to \emph{construct} a solution when its conditions are met.
-There is an obvious Las Vegas algorithm for finding a solution -- since a solution exists, just sample from $\mathcal{P}$ until one is found.
-In fact, the LLL guarantees more than the existence of a solution; it provides a lower bound on $\mu(\cap_{i \in [m]} \bar{A_i})$.
+There is an obvious Las Vegas algorithm for finding a solution: just sample i.i.d.
+from $\mu$ until one is found.
+The running time of this algorithm has a geometric distribution with parameter $\mu(\cap_{i \in [m]} \bar{A_i})$ (the probability that no flaw happens), so its expectation is $\frac{1}{\mu(\cap_{i \in [m]} \bar{A_i})}$.
+In fact, the LLL guarantees more than the existence of a solution; it provides a strictly-positive lower bound on $\mu(\cap_{i \in [m]} \bar{A_i})$ and thus an upper bound on the expected running time of the Las Vegas algorithm.
 Unfortunately, this lower bound can be exponentially small, so there is no guarantee that the Las Vegas algorithm is efficient.
@@
 
-If $\Omega$ is continuous, then the same is true, except that we must make our arguments in terms of entropy rather than bits.
+If $\Omega$ is continuous, then the same is true, except that we must make our arguments in terms of an abstract differential entropy rather than bits.
 We have also ignored the fact that our argument was inductive, and when there is some uncertainty remaining in $\sigma_{i}$, we no longer know it exactly when we try to infer $\sigma_{i-1}$.
@@
 Then the ``prior'' on $Z$, the distribution before we observe anything, is $P(Z) = \operatorname{Unif}(2^{m+kt})$.
-By Bayes' rule, $H(Y | Z) = H(Z | Y) - H(Z) + H(Y)$.
-If we make the reasonable assumption that $Y$ can be recovered from the actual state vector $Z$, then $H(Y | Z) = 0$, and we have
+By Bayes' rule,
 \seqn
   \label{eqn:entropy-general}
-  H(Z) = H(Z | Y) + H(Y) .
-\eeqn
-Typically we will proceed by finding a simple expression for the left hand side (which is typically determined explicitly by the algorithm) and upper-bounding the right hand side.
-For example, if we can encode the alternative information $Y$ using $N$ bits, an upper bound on the prior entropy $H(Y)$ is $N$, and then equation \ref{eqn:entropy-general} becomes
+  H(Z) = H(Z | Y) + H(Y) - H(Y | Z) .
+\eeqn
+
+Typically we will proceed by finding a simple expression or lower bound for the left hand side (which is typically determined explicitly by the algorithm) and upper-bounding the right hand side.
+Since entropies are nonnegative, we can simply drop the $H(Y | Z)$ term when searching for an upper bound, if it is convenient.
+And if, for example, we can encode the alternative information $Y$ using $\hat{H}(Y)$ bits, an upper bound on the prior entropy $H(Y)$ is $\hat{H}(Y)$, and then equation \ref{eqn:entropy-general} becomes
 \seqn
@@
 \eeqn
-When $P(Z | Y)$ is a point mass, $H(Z | Y) = 0$, and inequality \ref{eqn:entropy-encoded} reduces to
+(Note that we need $\operatorname{Unif}(\{0,1\}^d)$ to be a suitable base measure for computing or bounding $H(Z)$ and $H(Z | Y)$.)  When $P(Z | Y)$ is a point mass, $H(Z | Y) = 0$, and inequality \ref{eqn:entropy-encoded} reduces to
 \seqn
@@
 \eeqn
-This is exactly the fact that we needed in section \ref{sec:lll} to show a contradiction when $t$ was too large.
+This is exactly the fact that we needed in section \ref{sec:lll} to show a contradiction when $t$ was too large, having shown that both $H(Z)$ and $N$ were affine functions of $t$.
 
@@
 This ensures there are $2^{ct}$ possible flaw-encodings, and thus (by atomicity of the underlying MT algorithm for \ksat) our posterior on $Z$, having observed $Y'$, is a uniform distribution over $2^{ct}$ state vectors.
-Then we could use Tao's argument to show that inequality \ref{eqn:entropy-encoded} is violated under exactly the conditions when it was violated for $Y$, since $H(Z | Y') + N' = H(Z | Y) + 2^{ct} + N - 2^{ct} = H(Z | Y) + N$.
+Then we could use Tao's argument to show that inequality \ref{eqn:entropy-encoded} is violated under exactly the conditions when it was violated for $Y$, since $H(Z | Y') + N' = H(Z | Y) + ct + N - ct = H(Z | Y) + N$.
 
 We can use exactly the same argument to apply Tao's proof in a situation where the ambiguity in $P(Z | Y)$ is induced by non-atomicity of the transformation graph, rather than by the artificial removal of information from $Y$.
+If $c$ states can lead to a single state by addressing the same flaw, then we now require that $e c p (d + 1) < 1$.
+%FIXME
+
 So why do we call this example trivial?
@@
 For example, say that we have a modified version of \ksat in which $2^c$ out of $2^k$ assignments to each clause can make it fail.
-We can rephrase any such problem as an ordinary \ksat problem by making $2^c$ copies of each clause, with each copy only disallowing one state.
-This would increase the size of $F$ by $2^c$ and increase by $c$ the number of bits required to describe any flaw in the alternative information.
+We can rephrase any such problem as an abstract \ksat-like problem by making $2^c$ copies of each clause, with each copy only disallowing one state.
+This would increase the size of $\mathcal{A}$ by $2^c$ and increase by $c$ the number of bits required to describe any flaw in the alternative information.
+(In terms of the conditions in Achlioptas' and Iliopoulos' work, the number of %FIXME
 So the argument in our trivial example does not buy us any extra power in handling non-atomicity, only an alternative proof of Achlioptas' and Iliopoulos' comment.
@@
   \item Further, each flaw must be addressed with positive probability when we resample it.
-More precisely, there is a positive lower bound $p^*$ on the probability of each flaw not happening when we resample it conditional on the variables on which it does not depend, regardless of the value of those variables:
+More precisely, there is an upper bound $p^* < 1$ on the probability of any single flaw happening when we resample conditional on the variables on which it does not depend, regardless of the value of those variables:
     \seqn
       \label{eqn:feas-condprob-condition}
-      p^* = \inf_{A_i \in \mathcal{A}} \essinf_{x \in \Omega} P_{|\Nats \setminus \operatorname{vbl}(A_i)}(\bar{A_i} | \phi(x)[\Nats \setminus \operatorname{vbl}(A_i)]) .
+      p^* = \sup_{A_i \in \mathcal{A}} \esssup_{x \in \Omega} P_{|\Nats \setminus \operatorname{vbl}(A_i)}(A_i | \phi(x)[\Nats \setminus \operatorname{vbl}(A_i)]) .
     \eeqn
@@
 \label{sec:markov}
+Our first step is to prove a general theorem about random walks.  Achlioptas and Iliopoulos discuss random walks that find perfect objects under two conditions:
+\begin{enumerate}
+  \item The random walk must be perfectly reconstructable given a perfect object and a tree-history of addressed flaws.
+  \item The random walk must end up in a $0$-entropy state, which we can think of as either a terminating state or a self-loop with probability $1$.
+These states, and only these states, must be our desired perfect objects.
+\end{enumerate}
+In this section, we generalize this notion to \emph{highly-reconstructable} random walks that hit \emph{low-transition-entropy subgraphs}.
+This is reminiscent of the fact that a random walk will quickly find a subgraph of small conductance.
+
+First, some setup and notation.
+Fix a $\sigma$-algebra $(U, \Sigma)$.
+A \emph{random walk graph} is a weighted digraph $G = (U, E, w)$ for which the edges leading out of each node define a probability distribution.
+That is, $w(i, \cdot)$ is a probability measure on $U$ (perhaps supported only on a ``neighborhood'' of $i$) for all nodes $i$.
+We define the edges and weights of a \emph{node-induced} random walk subgraph $G'$ of $G$ on nodes $U'$ as follows: $G'$ has nodes $U'$, edges $\{(i,j): (i,j) \in E; i, j \in U'\}$, and renormalized weights $w'(i, \cdot) = 1[i \in U'] 1_{U'} \frac{1}{w(i, U')} w(i, \cdot)$.
+(We assume $w(i, U') > 0$ for all $i \in U'$.)
+%FIXME: Kinda sloppy.  Do we need to allow U' to have measure 0 sometimes?
+%FIXME: This repeats some of \ref{sec:perfect}.
+
+Fix a (possibly infinite) random walk graph $G = (U, E, w)$ and a node-induced subgraph $G' = (V, F, w')$.
+We will define $V$ later.
+Think of $V$ as the set of flawed states and $\bar{V} = U \setminus V$ as the set of flawless states.
+
+Let the stochastic process $\sigma(\mu, G) = (\sigma_0, \sigma_1, \cdots) \in U^\infty$ be the infinite Markov chain recording states of a random walk that starts by drawing $\sigma_0 \hasDist \mu$ and drawing $\sigma_{i+1} \hasDist w(\sigma_i, \cdot)$.
+(When the graph and starting measure are clear we will just write $\sigma$.)  And let $\sigma^(t) = (\sigma_i)_{i=0}^{t}$ be the truncation of $\sigma$ to $t+1$ steps.
+Let $h(i, J)$ be the expected hitting time from a starting node $i$ to a destination node set $J$, that is, $\E \inf \{t: \sigma_t \in J\}$ when $\mu = \delta_{\{i\}}(\cdot)$.
+Let $h(\mu, I, J) = \esssup_{\mu, i \in I} h(i, J)$ be the expected hitting time from the ``furthest-away'' node in $I$, possibly up to sets of $\mu$-measure $0$ in $I$.
+We will show several entropy-based conditions for bounding $h(\mu, V, \bar{V})$.
+
+We will choose $V$ to be the set of all ``high-transition-entropy'' nodes in $U$.  More precisely, $V = \{i \in U: \}$ %FIXME
+
+% WTS: H(Z) < 
+% H(Z): = H(\sigma_0) + \sum_{i=1}^{t} H(\sigma_i | \sigma_{i-1}).
+That is, the entropy rate of Markov chain is just equal to the transition entropies.
+So we should make an assumption that lower-bounds H(\sigma_i | \sigma_{i-1}) for all \sigma_{i-1}.
+% H(Y): = H(\sigma_t) + H(Y without \sigma_t | \sigma_t)
+% H(\sigma_t): We would like for the initial distribution \mu to have more entropy than the stationary distribution of the Markov chain, since I think that upper-bounds H(\sigma_t), and we would like H(\sigma_t) \leq H(\sigma_0) to avoid any term that depends on the entropy of \mu.
+That is optional, though, since the entropy of \mu is going to be polynomial in any problem parameter.
+% H(Y without \sigma_t | \sigma_t): We assume Y is just the sequence of addressed flaws.
+Then H(A | \sigma_t) = \sum_{i=t}^{1} H(A_i | A_{i+1}, \cdots, A_{t}, \sigma_t).
+There is a polynomial-time algorithm to compute these quantities, but is there a simple closed-form upper bound?
+% H(Z | Y): 
+% H(Y | Z):
 
@@
 \label{sec:nonatomic}
+
+% \section{A new convex programming approach to the ```lopsided''' LLL?}
+% In this section, we note
+% Inductive hypothesis: In k-SAT, if the (lopsided) LLL is satisfied for a problem p with n variables, then there exists an assignment to one of the variables so that the reduced problem p' satisfies the LLL.
+Stronger: For each variable, there exists an assignment to that variable so that the reduced problem p' satisfies the LLL.
+% n = 1: If the LLL is satisfied for a problem in 1 variable, there exists a satisfying solution to the problem, so we can assign the 1 variable that way.
+A problem with no variables or clauses trivially satisfies the LLL.
+% n = k+1: Consider an assignment to an arbitrary variable i.
+For each clause j not containing i, p_j' = p_j.
+For each clause j containing i, either the clause is now satisfied and has disappeared, or else p_j' = 2*p_j (the clause wanted the other assignment to i and now has half as many remaining good assignments).
+In the simple LLL, we have to show that the degree has dropped by a factor of 2 for at least one potential variable choice i.
 
